### 2018-01-13 manuscript.tex
Initial ShareLaTeX Import

\documentclass[Afour,sageh,times]{sagej}
\usepackage[utf8]{inputenc}
\usepackage[USenglish]{babel}
\usepackage{booktabs}
\usepackage{setspace}
\usepackage{csquotes}
\usepackage{url}
\usepackage{moreverb}
\usepackage[colorlinks,bookmarksopen,bookmarksnumbered,citecolor=red,urlcolor=red]{hyperref}
\newcommand\BibTeX{{\rmfamily B\kern-.05em \textsc{i\kern-.025em b}\kern-.08em T\kern-.1667em\lower.7ex\hbox{E}\kern-.125emX}}

\def\volumeyear{\the\year}

\begin{document}

\runninghead{Author Redacted}
\title{The Importance of Being Nonplanar: Street Network Representation in Urban Form Studies}
\author{Author Redacted \affilnum{1}}
\affiliation{\affilnum{1}Affiliation redacted}
\corrauth{Author Redacted, Address Redacted}
\email{Email Redacted}

\begin{abstract}

\end{abstract}

\keywords{street network, GIS, urban form, transportation, urban design}

\maketitle

\section{Introduction}

In urban planning and urban design research, street networks are routinely used to calculate accessibility between origins and destinations or to compute indicators of the urban form, such as block sizes or intersection density and connectivity.

This article presents two new measures of the degree of nonplanarity for urban street networks that can be generalized to any spatial network. It compares the results and interpretation of these nonplanarity indicators to the formal, mathematical planarity of street networks around the world. Despite common claims in the literature that street networks are planar graphs, this study finds that they generally are not and that planar graphs poorly model the street networks of many cities. Further, the magnitude of this bias varies substantially across cities and urbanization types.

This article is organized as follows. The following section introduces the relevant basics of graph theory urban studies, focusing on discussions in the research literature about street network planarity. The next section introduces the methods used to acquire and analyze the street networks in this study. Then this article presents the results of this analysis before concluding with a discussion of their ramifications for street network research and urban form studies.

\section{Background}

First, a few definitions are in order. Graph theory is the mathematical study of networks \citep{newman_networks:_2010}. Graphs can mathematically model real-world networks such as friendships or the Internet as well as spatial networks, such urban street networks \citep{barthelemy_spatial_2011}. A graph $G$ consists of a set of nodes $N$ connected to one another by a set of edges $E$. An edge $uv$ in a directed graph points in one direction from some node $u$ to some node $v$, but an undirected graph's edges all mutually point in both directions. How a graph's nodes and edges connect to one another defines its topology. For example, a node's degree is a topological trait that represents how many edges connect to that node. A planar graph can be drawn on a two-dimensional plane without any of its edges crossing each other, except where they intersect at nodes. If the graph \emph{cannot} be drawn or redrawn to meet this criterion, then the graph is nonplanar \citep{trudeau_introduction_1994}. Street networks are embedded in space, which provides them with geometry --- such as geographical coordinates, lengths, and area --- alongside their topology.

\begin{table*}[htbp]
\centering
\caption{Recent statements in the urban studies and urban physics literatures regarding the representation of street networks as planar graphs.}
\label{tab:planar_quotes}
\input{table_planar_quotes.tex}
\end{table*}

This creates a minor wrinkle when we consider planarity: we must distinguish between a graph's topological planarity, which we shall refer to as \emph{formal planarity}, and the planarity of its particular spatial embedding, which we shall refer to as \emph{spatial planarity}. A street network might be spatially nonplanar due to its embedding in space (i.e., it contains overpasses or underpasses in the real-world), but it could still be formally planar. For example, if we \enquote{redraw} the graph by moving its nodes and edges around in space without changing how they are connected to one another (i.e., altering its geometry without altering its topology), there may exist some spatial embedding that prevents edges crossing anywhere but nodes \citep[for a more comprehensive discussion see][pp.~6--10]{barthelemy_morphogenesis_2017}. In such a case, the street network is formally planar from a topological perspective, but its real-world embedding is spatially nonplanar.

Consider two theoretical examples: a medieval European city center and a modern American city center. The former's circulation network comprises a set of pedestrian paths and drivable streets without any grade separation, overpasses, or underpasses. In two-dimensions, its edges (i.e., streets and paths) cross each other only at nodes (i.e., intersections), so it is by definition planar. The latter's circulation network comprises pedestrian paths, drivable surface streets, and a grade-separated freeway with overpasses over some of the surface streets. In two-dimensions, its edges occasionally cross each other at non-nodes (i.e., overpasses), so it is by definition nonplanar. However, we might call it \emph{approximately} planar because its nonplanar edge crossings are relatively uncommon. Approximate planarity constrains this nonplanar spatial network such that it does not exhibit certain characteristics found among nonplanar aspatial graphs (e.g., small-world effects or power-law distributed node degrees).

In the urban studies and urban physics literature, street networks are commonly referred to as planar graphs. Table \ref{tab:planar_quotes} presents a survey of statements and reasoning around this claim. Some authors prefer to hedge slightly, arguing that street networks are \emph{approximately} or \emph{essentially} planar graphs that are close enough to be well-modeled as such.

If street networks can be sufficiently well-modeled by planar graphs, there are certain methodological benefits to doing so. Planar graphs offer computational simplicity and tractability. They enable easy polygonal spatial analysis of city blocks and form \citep{fohl_non-planar_1996} as well as the Euler characteristic. In mathematics, there is a bijection between planar graphs and trees, and classifying planar graphs presents a trivial problem \citep{louf_typology_2014}. Planar graphs are easier to visualize and can be faster to run algorithms on \citep{liebers_planarizing_2001}. Accordingly, \citet[p.~3]{barthelemy_spatial_2011} argues that \enquote{planar spatial networks are the most important and most studies have focused on these examples}. But in contrast, \citet{masucci_random_2009} and \citet{masucci_limited_2013} argue that planar graphs remain a compelling research domain for urban scholars because they were understudied until recently for two reasons: they appear topologically trivial and planarity does not lend itself to certain popular graph-theoretic analyses. Discussing the open research area around street networks as planar graphs, \citet[p.~1]{viana_simplicity_2013} state, \enquote{there is still a lack of global, high-level metrics allowing to characterize their structure and geometrical patterns.}

Despite the computational and mathematical advantages of simple planar models, street networks are often nonplanar in reality: many include at least one overpass or underpass that results in the failure of formal proofs of their planarity, such as the \citet{kuratowski_sur_1930} theorem or the \cite{hopcroft_efficient_1974} algorithm \citep[cf.][]{gastner_spatial_2006}. As \citet[p.~7]{levinson_network_2012} points out, \enquote{Real networks are neither perfect, nor planar, nor grids, though they may approximate them.}

Other authors have commented on this characteristic of street networks. \citet[p.~199]{jiang_object-oriented_2010} explain that \enquote{Quite often the transportation network has overpasses and underpasses that require a non-planar network representation.} \citet[p.~1258]{fischer_spatial_2014} explains that \enquote{For many infrastructure networks, {[planarity]} is approximately true, although bridges and tunnels in ground-transport networks are an obvious (but generally minor) exception.} Twenty years ago, \citep[p.~18]{fohl_non-planar_1996} claimed, \enquote{The most commonly used data model for transportation networks is the fully intersected, planar data model} and called for a nonplanar model to better represent truly nonplanar spatial networks. \enquote{The planar network data model has received widespread acceptance and use. Despite its popularity, the model has limitations for some areas of transportation analysis, especially where complex network structures are involved. One major problem is caused by the planar embedding requirement... intersections at grade cannot be distinguished from intersections with an overpass or underpass that do not cross at grade.} \citep[p.~395]{fischer_gis_2004}

If a planar graph models a street network poorly, it could do so in multiple ways. Forcing planarity on a nonplanar street network creates artificial nodes at bridges and tunnels, which breaks routing. As \citet[p.~6]{kwan_review_1996} puts it, \enquote{the difficulty in accurately representing overpasses or underpasses may lead to problems when running various routing algorithms (e.g. recommending that a traveler make a left-turn at an intersection that proves to be an overpass)}. Second, and accordingly, intersection counts and density will be overestimated. Similarly, edge lengths will be underestimated due to these artificial breakpoints. Finally, this bias would likely behave inconsistently across different kinds of cities and street network types.

Given these issues, what do \emph{approximately planar} and \emph{well-modeled} mean for street networks? How close is close enough for a planar graph to competently model a mathematically nonplanar street network? Do the biases of planar models remain consistent across geographies or do they misrepresent different places to different degrees? If street networks are in general approximately planar, then how can we measure how planar or nonplanar a given street network is?

The graph theory literature offers some measures of how \enquote{far off} a nonplanar graph $G$ is from being planar, such as its crossing number --- the minimum number of edge crossings of any drawing of $G$ --- and the skewness --- the minimum number of edges that must be removed from $G$ to produce a planar graph \citep{liebers_planarizing_2001,szekely_successful_2004,chimani_non-planar_2009}. However, these measures are imperfect and hard to compute \citep{chimani_vertex_2012,szekely_successful_2004}. In his discussion of road networks and approximate spatial planarity, \citet[p.~133]{newman_networks:_2010} argues that \enquote{no widely accepted metric for degree of planarity has emerged,} and calls for the development of better indicators.

Such metrics would be particularly useful for street networks, as the extent to which a network is (or is not) planar can characterize the nature of its circulation infrastructure and urban form. For instance, late 20th-century American cities might exhibit lower planarity than older European cities.

In this study, we present two new measures of the extent to which a spatial network is planar. We then analyze various world cities to better understand how well planar graphs model their street networks as well as the extent to which bias (i.e., model misrepresentation) varies across different places or types of urbanization. Finally, we consider what these indicators suggest about the nature of the urban form in different cities.

\section{Methods}

\subsection{Data}

We analyze the drivable and the walkable street networks in the centers of 50 cities worldwide. Following \citet{jacobs_great_1995} and \citet{cardillo_structural_2006}, we analyze a consistently sized, square-mile network at the centers of each of these cities.

OSMnx and how the networks are processed. walk vs drive.

\subsection{Analysis}

Once we have acquired and prepared our networks, we calculate three indicators of nonplanarity. The first indicator, SPS, represents the ratio of nonplanar intersections (i.e., non-dead-end nodes in the nonplanar, spatially-embedded graph) to planar intersections (i.e., line crossings in the planar, spatially-embedded graph). It indicates the extent to which planarity overstates intersections and connectivity in each street network. A truly planar street network with no bridges or tunnels will thus have a SPS score of 1.0, while lower values suggest the extent to which the network is nonplanar.

The second indicator, ELR, represents the ratio of the mean edge length in the planar graph to the mean edge length in the nonplanar graph. It indicates the extent to which planarity understates edge lengths (a linear proxy for block size) in each street network. A truly planar street network with no bridges or tunnels will thus have a ELR score of 1.0, while lower values suggest the extent to which the network is nonplanar.

The third indicator is a formal test of aspatial planarity using the algorithm described by \citet{boyer_subgraph_2012}. A graph is mathematically planar if it can be drawn on a two-dimensional surface with edges crossing only at nodes \citep{trudeau_introduction_1994}. Although a street network may be spatially nonplanar due to overpasses or underpasses, this may be a feature of its configuration in space. That is, is it possible to rearrange the nodes and edges in space --- while preserving the graph's topology --- so that edges cross only at nodes? This indicator tells us if the graph is formally planar, ignoring its real-world spatial embedding.

Finally, we examine a single city more deeply to better understand how these indicators vary across a single municipality. For this, we analyze the drivable street network of Oakland, California, a reasonably representative midsized American city with a variety of urban form types, from gridded street patterns in its flatlands, to winding culs-de-sac in its hills, to freeways and dense blocks around its downtown. We first analyze the entire city of Oakland and then recreate the aforementioned methodology by sampling 100 random points within the city limits and analyzing a square-mile street network centered on each point. This demonstrates the statistical dispersion of the planarity indicators across a single city and the extent to which agglomerating an entire city's neighborhoods into a single graph may obscure neighborhood-scale effects.

\section{Results}

\subsection{Worldwide Cities}

\begin{figure*}[htbp]
    \center
    \includegraphics[width=\textwidth]{world_map_phi_bw.png}
    \caption{Map of world cities in Table \ref{tab:world_cities} grouped by SPS score tertiles.}
    \label{fig:world_map_bw}
\end{figure*}

\begin{table*}[htbp]
\centering
\caption{Indicators of nonplanarity for one square mile street networks at the centers of 50 cities worldwide. SPS = spatial planarity score = the ratio of nonplanar intersections to planar intersections. ELR = edge length ratio = the ratio of mean edge length in planar graph to mean edge length in nonplanar graph. Planar = whether street network passed a formal test of planarity.}
\label{tab:world_cities}
\input{table_world_cities.tex}
\end{table*}

\begin{figure*}[htbp]
    \center
    \includegraphics[width=0.5\textwidth]{regression_phi_split.png}
    \caption{Log-log plot of ELR vs SPS with simple regression lines: drive $r^2=0.98$ and walk $r^2=0.99$.}
    \label{fig:regression_split}
\end{figure*}

\subsection{A Deeper Look: Oakland, California}

We find that the street network of the entire city of Oakland is formally nonplanar when subjected to a formal mathematical test. In terms of the nonplanarity indicators, the city as a whole has a SPS score of 0.918, indicating that it is 91.8\% planar, and a ELR score of 0.936. This suggests that the planar representation of Oakland's drivable street network overstates the number of intersections (and thus, the network's connectivity) by 8.9\% city-wide (i.e., $1 / 0.918$ and understates the average edge length by 6.4\% city-wide. However, these indicators' values vary across the city.

To explore this statistical variation, we examined 100 square-mile samples of Oakland's drivable street network. Table \ref{tab:samples_city} presents descriptive statistics of these planarity indicators.

\begin{table}[htbp]
\centering
\caption{Descriptive statistics of planarity indicators across 100 random square-mile sample runs in Oakland, California's drivable network.}
\label{tab:samples_city}
\input{table_samples_city.tex}
\end{table}

\section{Discussion}

Our findings suggest that most street networks in the centers of major cities are mathematically nonplanar but approximately planar. However, this depends on the scale of measurement: across an entire city there is likely to be some underpass or overpass somewhere, while individual neighborhoods or small suburbs might be formally planar. The type and era of urbanization is another factor. Old medieval European towns or informal settlements in the global south contain fewer grade-separated roads (and thus greater planarity) than 20th-century American or 21st-century Chinese metropolises do. This is a result of prevailing transportation technologies when the urban form was built up, as well as terrain, costs, wealth, culture, and politics.

Street networks are frequently nonplanar in the formal sense because they are spatially embedded in three dimensions --- not two --- and have a z-coordinate along with their x and y. Because they are mostly planar, typically with only a few overpasses or underpasses, they could often be described accurately as \emph{approximately planar}. However, claiming that urban street networks broadly are planar misrepresents them in several ways.

1. Forces false nodes where grade-separated edges cross.

2. Accordingly, underestimates average edge length (a proxy for street segment lengths and block sizes)

3. Misrepresents connectivity for routing, accessibility analysis, and other connectivity studies

Why are they approximately planar? Cost and politics. Expensive to build truly 3-D networks (with z-coordinates as extensive as their x- and y-coordinates) and politically infeasible.

Are street networks planar graphs? Contrary to some of the statements in the urban studies and urban physics literature, our results suggest that this cannot be universally claimed. But perhaps some of it comes down to vocabulary. If road is not synonymous with street, then a road network and a street network may not be synonymous. A road network, including freeways and boulevards, may frequently be nonplanar, but a street network, focusing on municipal streets lined by land parcels, may be at-grade and planar. But this ignores the fact that even residential streets sometimes include bridges and tunnels in hilly neighborhoods, and the fact that our analysis earlier showed that walkable circulation networks in city centers often include pedestrian tunnels and footbridges. A graph is not planar because its edges \emph{usually} or \emph{approximately} intersect only at nodes: it is planar because its edges \emph{exclusively} intersect at nodes.

Thus, street networks are not universally planar graphs. But as George Box famously stated, \enquote{All models are wrong but some are useful.} Can street networks generally be simplified to planar graphs and still be usefully well-modeled? Then answer depends on the place and on the type of analysis. In limited circumstances, where the circulation network and built form exhibit few underpasses, overpasses, or grade-separation, then perhaps yes. But universally, we must answer no.

Most egregiously, imposing planarity on a nonplanar street network forces pseudonodes at underpasses and overpasses, breaking routing and network-based accessibility modeling. For this reason, nonplanar graphs have been the standard for decades in transportation engineering, real-world traffic assignment models, and routing engines. But planar graphs are often used in the literature to study urban form and morphology. So, aside from routing, do planar graphs offer \emph{useful} models for this type of research? Again, only in limited circumstances, including examples such as the drivable networks in the three Italian cities we analyzed, which are formally planar. The results in Table \ref{tab:world_cities} show how common urban form measures such as intersection counts are overstated by planar models, while average edge lengths (a linear proxy for block size) is consequently understated. Moreover, this misrepresentation behaves inconsistently from place to place. Figure \ref{fig:world_map_bw} demonstrates how the magnitude of bias varies across cities and eras of urbanization.

So then why use planar graphs? Tractable. In some cities it doesn't matter (Italy). Polygonal analysis of urban blocks. Otherwise use nonplanar, new tools make it easy.

\section{Conclusion}

The urban studies and urban physics literatures frequently use planar graphs to urban street networks. This study demonstrated that although in limited circumstances these models may be accurate, they can behave inconsistently across different kinds of cities by misrepresenting connectivity, accessibility, routing, intersection counts and densities, and street segment lengths. It found x, y, and z. It also demonstrated how these indicators can be used to characterize the type of urbanization, in particular its three-dimensionality, of different cities. Future research can explore this latter finding, as it likely correlates with other measures of urbanization, development, and era. Finally, future research might examine how nonplanar intersection counts represent true intersections if multiple adjacent edges form multiple graph intersections at a point where only one true intersection exists from an urban design perspective.

\clearpage
\bibliographystyle{apa}
\bibliography{references}

\end{document}

### 2018-01-14 manuscript.tex
add to methods, results, discussion

@@
 In urban planning and urban design research, street networks are routinely used to calculate accessibility between origins and destinations or to compute indicators of the urban form, such as block sizes or intersection density and connectivity.
+Mathematical models of street networks, called graphs, have grown ubiquitous in the urban studies literature in recent years as they can be used to usefully model pedestrian volume, household travel patterns, urban design patterns and spatial morphology, access and equity, and location centrality and polycentricity \citep{marshall_street_2010,porta_alterations_2014,marshall_community_2014,hajrasouliha_impact_2015,parthasarathi_street_2015,knight_metrics_2015,gil_street_2016,zhong_revealing_2017}.
 
@@
 
-First, a few definitions are in order.
 Graph theory is the mathematical study of networks \citep{newman_networks:_2010}.
@@
 An edge $uv$ in a directed graph points in one direction from some node $u$ to some node $v$, but an undirected graph's edges all mutually point in both directions.
-How a graph's nodes and edges connect to one another defines its topology.
-For example, a node's degree is a topological trait that represents how many edges connect to that node.
-A planar graph can be drawn on a two-dimensional plane without any of its edges crossing each other, except where they intersect at nodes.
+In a street network, the nodes represent intersections and dead-ends, and the (directed) edges represent the street segments that connect them.
+How a graph's nodes and edges connect to one another defines its \emph{topology}.
+For example, a node's \emph{degree} is a topological trait that represents how many edges connect to that node.
+A \emph{planar} graph can be drawn on a two-dimensional plane without any of its edges crossing each other, except where they intersect at nodes.
 If the graph \emph{cannot} be drawn or redrawn to meet this criterion, then the graph is nonplanar \citep{trudeau_introduction_1994}.
@@
 Such metrics would be particularly useful for street networks, as the extent to which a network is (or is not) planar can characterize the nature of its circulation infrastructure and urban form.
-For instance, late 20th-century American cities might exhibit lower planarity than older European cities.
-
-In this study, we present two new measures of the extent to which a spatial network is planar.
-We then analyze various world cities to better understand how well planar graphs model their street networks as well as the extent to which bias (i.e., model misrepresentation) varies across different places or types of urbanization.
-Finally, we consider what these indicators suggest about the nature of the urban form in different cities.
+For instance, late 20th-century freeway-oriented American cities might exhibit lower planarity than older walkable European cities.
+What about informal settlements in developing countries or rapidly urbanizing Chinese cities?
+Beyond the question of graph model goodness-of-fit, such indicators could provide useful information about urban development, civil infrastructure, and transportation system character.
 
@@
 
+In this study, we develop two new measures of the extent to which a spatial network is planar.
+We then analyze various world cities to better understand how well planar graphs model their street networks as well as the extent to which bias (i.e., model misrepresentation) varies across different places and types of urbanization.
+Finally, we consider what these indicators suggest about the nature of the urban form and transportation infrastructure in different cities.
+
 \subsection{Data}
 
-We analyze the drivable and the walkable street networks in the centers of 50 cities worldwide.
-Following \citet{jacobs_great_1995} and \citet{cardillo_structural_2006}, we analyze a consistently sized, square-mile network at the centers of each of these cities.
+Following \citet{jacobs_great_1995} and \citet{cardillo_structural_2006}, we analyze a consistently sized, square-mile network at the centers of 50 cities worldwide.
+This allows us to consistently examine central urban street networks without being swamped by metropolitan-scale variation or the idiosyncrasies of individual municipalities' spatial extents.
+For cities, such as Moscow, with a newer commercial central business district (CBD) that lies apart from an \enquote{old town} center, we use the modern CBD as the city center.
+We look separately at their drivable and their walkable street networks.
 
-OSMnx and how the networks are processed. walk vs drive.
+To acquire these street networks, we use OSMnx to download the data for each city and network type from OpenStreetMap.
+OpenStreetMap is a collaborative online mapping platform commonly used by researchers because of its good worldwide coverage \citep{haklay_how_2010,jokar_arsanjani_openstreetmap_2015}.
+OSMnx is a Python-based software tool that allows us to automatically download a street network from OpenStreetMap for any study site in the world, and automatically process it into a length-weighted nonplanar directed graph (citation redacted).
+It differentiates between walkable and drivable routes in the circulation network based on individual elements' metadata that describe how the route may be used.
+Thus the walkable network may contain surface streets, paths through parks, pedestrian passageways between buildings or under roads, and other walkable paths.
+The drivable network may contain surface streets, grade-separated freeways, and other drivable paths.
+
+OpenStreetMap graphs contain many interstitial nodes in the middle of street segments (forming an expansion graph) to allow streets to curve through space via a series of straight-line approximations.
+OSMnx automatically simplifies each graph's topology to retain nodes only at intersections and dead-ends, while faithfully retaining each edge's true spatial geometry.
+This provides us with an accurate count of intersections and an accurate measure of edge lengths for comparison between the planar and nonplanar representations of our networks.
 
@@
 
-Once we have acquired and prepared our networks, we calculate three indicators of nonplanarity.
-The first indicator, SPS, represents the ratio of nonplanar intersections (i.e., non-dead-end nodes in the nonplanar, spatially-embedded graph) to planar intersections (i.e., line crossings in the planar, spatially-embedded graph).
+Once we have acquired and prepared our networks, we calculate three measures of planarity.
+
+The first is a formal test of aspatial planarity using the algorithm described by \citet{boyer_subgraph_2012}.
+That is, is it possible to rearrange the graph's nodes and edges in space --- while preserving its topology --- so that edges cross only at nodes?
+This binary true/false indicator tells us if the graph is formally planar, ignoring its real-world spatial embedding.
+However, street networks are spatially embedded, so we develop the following two continuous measures to assess the \enquote{extent} to which they are planar.
+
+The second measure is the spatial planarity ratio (SPR).
+It represents the ratio of nonplanar intersections (i.e., non-dead-end nodes in the nonplanar, three-dimensional, spatially-embedded graph) to planar intersections (i.e., line crossings in the planar, two-dimensional, spatially-embedded graph).
 It indicates the extent to which planarity overstates intersections and connectivity in each street network.
-A truly planar street network with no bridges or tunnels will thus have a SPS score of 1.0, while lower values suggest the extent to which the network is nonplanar.
+A truly planar network with no bridges or tunnels will thus have an SPR score of 1.0, while lower values suggest the extent to which the network is nonplanar.
 
-The second indicator, ELR, represents the ratio of the mean edge length in the planar graph to the mean edge length in the nonplanar graph.
-It indicates the extent to which planarity understates edge lengths (a linear proxy for block size) in each street network.
-A truly planar street network with no bridges or tunnels will thus have a ELR score of 1.0, while lower values suggest the extent to which the network is nonplanar.
+The third measure is the edge length ratio (ELR).
+It represents the ratio of the mean edge length in the planar graph to the mean edge length in the nonplanar graph.
+It indicates the extent to which planarity understates edge lengths in each street network.
+A truly planar street network with no bridges or tunnels will thus have an ELR score of 1.0, while lower values suggest the extent to which the network is nonplanar.
 
-The third indicator is a formal test of aspatial planarity using the algorithm described by \citet{boyer_subgraph_2012}.
-A graph is mathematically planar if it can be drawn on a two-dimensional surface with edges crossing only at nodes \citep{trudeau_introduction_1994}.
-Although a street network may be spatially nonplanar due to overpasses or underpasses, this may be a feature of its configuration in space.
-That is, is it possible to rearrange the nodes and edges in space --- while preserving the graph's topology --- so that edges cross only at nodes?
-This indicator tells us if the graph is formally planar, ignoring its real-world spatial embedding.
-
-Finally, we examine a single city more deeply to better understand how these indicators vary across a single municipality.
-For this, we analyze the drivable street network of Oakland, California, a reasonably representative midsized American city with a variety of urban form types, from gridded street patterns in its flatlands, to winding culs-de-sac in its hills, to freeways and dense blocks around its downtown.
-We first analyze the entire city of Oakland and then recreate the aforementioned methodology by sampling 100 random points within the city limits and analyzing a square-mile street network centered on each point.
-This demonstrates the statistical dispersion of the planarity indicators across a single city and the extent to which agglomerating an entire city's neighborhoods into a single graph may obscure neighborhood-scale effects.
+Finally, we explore how these indicators might vary across a single municipality.
+To do so we analyze the drivable street network of Oakland, California, a reasonably representative midsized American city with a variety of urban form types from gridded street patterns in its flatlands, to winding culs-de-sac in its hills, to freeways and dense blocks around its downtown.
+First we analyze the entire city of Oakland.
+Then we recreate the aforementioned methodology by randomly sampling 100 points within the city limits and analyzing the square-mile street network centered on each.
+The resulting statistical dispersion of planarity indicators demonstrates the extent to which agglomerating an entire city's neighborhoods into a single graph may obscure neighborhood-scale effects and characteristics.
 
@@
 
+Among drivable street networks, only x\% pass the formal test of planarity.
+On average, they are only x\% planar by the SPR measure and x\% planar by the ELR measure.
+SPR values range from a high of 1.0 in x cities to a low of 57\% in Moscow.
+ELR values range from a high of 1.0 in x cities to a low of 64\% in Los Angeles.
+
+Among walkable street networks, only x\% pass the formal test of planarity.
+On average, they are only x\% planar by the SPR measure and x\% planar by the ELR measure.
+SPR values range from a high of 1.0 in x cities to a low of 74\% in Atlanta.
+ELR values range from a high of 1.0 in x cities to a low of 72\% in Atlanta.
+
+Mogadishu is the only city to show perfect planarity across all three indicators and both network types.
+All three Italian cities show perfect planarity in their drivable networks, but not in their walkable networks.
+
 \begin{figure*}[htbp]
@@
     \includegraphics[width=\textwidth]{world_map_phi_bw.png}
-    \caption{Map of world cities in Table \ref{tab:world_cities} grouped by SPS score tertiles.}
+    \caption{Map of world cities in Table \ref{tab:world_cities} grouped by SPR score tertiles (lower scores mean less planar).}
     \label{fig:world_map_bw}
@@
 \centering
-\caption{Indicators of nonplanarity for one square mile street networks at the centers of 50 cities worldwide.
-SPS = spatial planarity score = the ratio of nonplanar intersections to planar intersections.
-ELR = edge length ratio = the ratio of mean edge length in planar graph to mean edge length in nonplanar graph.
-Planar = whether street network passed a formal test of planarity.}
+\caption{Planarity measures for the central street networks of 50 cities worldwide (SPR = spatial planarity ratio; ELR = edge length ratio; Planar = whether street network passed the formal test of planarity).}
 \label{tab:world_cities}
@@
     \includegraphics[width=0.5\textwidth]{regression_phi_split.png}
-    \caption{Log-log plot of ELR vs SPS with simple regression lines: drive $r^2=0.98$ and walk $r^2=0.99$.}
+    \caption{Log-log plot of ELR vs SPR with simple regression lines: drive $r^2=0.98$ and walk $r^2=0.99$.}
     \label{fig:regression_split}
@@
 
-\subsection{A Deeper Look: Oakland, California}
+\subsection{A Closer Look: Oakland, California}
 
 We find that the street network of the entire city of Oakland is formally nonplanar when subjected to a formal mathematical test.
-In terms of the nonplanarity indicators, the city as a whole has a SPS score of 0.918, indicating that it is 91.8\% planar, and a ELR score of 0.936.
-This suggests that the planar representation of Oakland's drivable street network overstates the number of intersections (and thus, the network's connectivity) by 8.9\% city-wide (i.e., $1 / 0.918$ and understates the average edge length by 6.4\% city-wide.
+In terms of the nonplanarity indicators, the city as a whole has a SPR score of 0.918, indicating that it is 91.8\% planar, and a ELR score of 0.936.
+This suggests that the planar representation of Oakland's drivable street network overstates the number of intersections (and thus, the network's connectivity) by 8.9\% city-wide (i.e., $1 - 1 / 0.918$ and understates the average edge length by 6.4\% city-wide (i.e., $1 - 0.936$).
 However, these indicators' values vary across the city.
@@
 \centering
-\caption{Descriptive statistics of planarity indicators across 100 random square-mile sample runs in Oakland, California's drivable network.}
+\caption{Descriptive statistics of planarity indicators across 100 random samples in Oakland, California's drivable network.}
 \label{tab:samples_city}
@@
 
-Our findings suggest that most street networks in the centers of major cities are mathematically nonplanar but approximately planar.
+Our findings suggest that the street networks in the centers of most major cities are formally and spatially nonplanar.
 However, this depends on the scale of measurement: across an entire city there is likely to be some underpass or overpass somewhere, while individual neighborhoods or small suburbs might be formally planar.
@@
 Because they are mostly planar, typically with only a few overpasses or underpasses, they could often be described accurately as \emph{approximately planar}.
-However, claiming that urban street networks broadly are planar misrepresents them in several ways.
+However, claiming that urban street networks broadly are planar misrepresents them in several ways:
 
-1. Forces false nodes where grade-separated edges cross.
+\begin{enumerate}
+	\item{Forces false nodes where grade-separated edges cross.}
+	\item{Accordingly, underestimates average edge length (a proxy for street segment lengths and block sizes).}
+	\item{Misrepresents connectivity for routing, accessibility analysis, and other topological studies.}
+\end{enumerate}
 
-2. Accordingly, underestimates average edge length (a proxy for street segment lengths and block sizes)
+This is a bigger problem in Los Angeles than in Florence, but even in Florence, planar overcounts intersections around the \textit{sottopassaggio}, or pedestrian subway, near its central train station, Stazione di Santa Maria Novella.
+This is a pedestrian walkway, not a drivable road, but in a walkable town like Florence, we would surely want to model pedestrian access as part the street network, not drivable roads only.
+In this case, the nonplanar public pedestrian network must be considered.
 
-3. Misrepresents connectivity for routing, accessibility analysis, and other connectivity studies
+The problem is consistently and particularly pronounced around the downtowns of North American cities, due to the prevalence of freeways, bridges, and underpasses there.
+Driving networks are affected by these in particular.
+Walking networks are more affected by pedestrian flyovers and subways, as in Florence.
+Even networks of non-freeway, non-pedestrian-only surface streets will often be nonplanar due to bridges or tunnels in hilly neighborhoods or over rivers.
+
+The effect is inconsistent across cities, and across different neighborhoods within individual cities.
 
@@
 
-Are street networks planar graphs?
+\subsection{Are street networks planar graphs?}
+
 Contrary to some of the statements in the urban studies and urban physics literature, our results suggest that this cannot be universally claimed.
@@
 A graph is not planar because its edges \emph{usually} or \emph{approximately} intersect only at nodes: it is planar because its edges \emph{exclusively} intersect at nodes.
+Thus, street networks are not universally planar graphs.
 
-Thus, street networks are not universally planar graphs.
-But as George Box famously stated, \enquote{All models are wrong but some are useful.} Can street networks generally be simplified to planar graphs and still be usefully well-modeled?
-Then answer depends on the place and on the type of analysis.
+\subsection{Are street networks well-modeled by planar graphs?}
+
+As George Box famously stated, \enquote{All models are wrong but some are useful.} Can street networks be simplified to planar graphs and still be usefully well-modeled?
+The answer depends on the study site and on the type of analysis.
 In limited circumstances, where the circulation network and built form exhibit few underpasses, overpasses, or grade-separation, then perhaps yes.
-But universally, we must answer no.
+But universally, we cannot answer yes.
 
@@
 For this reason, nonplanar graphs have been the standard for decades in transportation engineering, real-world traffic assignment models, and routing engines.
+
 But planar graphs are often used in the literature to study urban form and morphology.
 So, aside from routing, do planar graphs offer \emph{useful} models for this type of research?
-Again, only in limited circumstances, including examples such as the drivable networks in the three Italian cities we analyzed, which are formally planar.
+Again, only in limited circumstances, including examples such as the drivable networks in the three Italian cities we analyzed, which are both formally and spatially planar.
 The results in Table \ref{tab:world_cities} show how common urban form measures such as intersection counts are overstated by planar models, while average edge lengths (a linear proxy for block size) is consequently understated.
@@
 The urban studies and urban physics literatures frequently use planar graphs to urban street networks.
-This study demonstrated that although in limited circumstances these models may be accurate, they can behave inconsistently across different kinds of cities by misrepresenting connectivity, accessibility, routing, intersection counts and densities, and street segment lengths.
+This study demonstrated that although in limited circumstances these models may be accurate, they behave inconsistently across different kinds of cities by misrepresenting connectivity, accessibility, routing, intersection counts and densities, and street segment lengths.
 It found x, y, and z.
-It also demonstrated how these indicators can be used to characterize the type of urbanization, in particular its three-dimensionality, of different cities.
+It also demonstrated how these indicators can be used to characterize the type of urbanization, in particular its transportation infrastructure's three-dimensionality, of different cities.
 Future research can explore this latter finding, as it likely correlates with other measures of urbanization, development, and era.
@@
 
-\clearpage
+%\clearpage
 \bibliographystyle{apa}

### 2018-01-14 manuscript.tex
update intro, results

@@
 
-In urban planning and urban design research, street networks are routinely used to calculate accessibility between origins and destinations or to compute indicators of the urban form, such as block sizes or intersection density and connectivity.
-Mathematical models of street networks, called graphs, have grown ubiquitous in the urban studies literature in recent years as they can be used to usefully model pedestrian volume, household travel patterns, urban design patterns and spatial morphology, access and equity, and location centrality and polycentricity \citep{marshall_street_2010,porta_alterations_2014,marshall_community_2014,hajrasouliha_impact_2015,parthasarathi_street_2015,knight_metrics_2015,gil_street_2016,zhong_revealing_2017}.
+In urban planning and design research, street networks are routinely used to calculate accessibility between origins and destinations or to compute indicators of the urban form, such as block sizes or intersection density and connectivity.
+Mathematical models of street networks, called graphs, have grown ubiquitous in the urban studies literature in recent years as they have been used to model household travel patterns, access and equity, pedestrian volume, urban design patterns and spatial morphology, and location centrality and polycentricity \citep{marshall_street_2010,porta_alterations_2014,marshall_community_2014,hajrasouliha_impact_2015,parthasarathi_street_2015,knight_metrics_2015,gil_street_2016,zhong_revealing_2017}.
 
-This article presents two new measures of the degree of nonplanarity for urban street networks that can be generalized to any spatial network.
-It compares the results and interpretation of these nonplanarity indicators to the formal, mathematical planarity of street networks around the world.
-Despite common claims in the literature that street networks are planar graphs, this study finds that they generally are not and that planar graphs poorly model the street networks of many cities.
+In the urban studies literature, street networks are typically referred to as \enquote{planar} or \enquote{approximately planar} (see Table \ref{tab:planar_quotes}), meaning that they can be well-modeled by a flat two-dimensional model that inherently forbids overpasses or underpasses.
+Of course in the real-world, urban street networks are embedded in three-dimensional space and often feature grade separation, bridges, and tunnels.
+This leads to two intertwined questions underexplored in the literature today.
+First, how well do planar graphs model urban street networks?
+Second, what does the extent to which a network is nonplanar tell us about the city's infrastructure and development?
+
+This study tests the planarity of street networks around the world.
+It also presents two new measures of the degree of nonplanarity that can be generalized to other types of spatial networks.
+These indicators help describe the nature of the urban form and transportation infrastructure.
+Despite common claims in the literature that street networks are planar graphs, this study finds that they generally are nonplanar and that planar graphs poorly model the street networks of many cities.
 Further, the magnitude of this bias varies substantially across cities and urbanization types.
@@
 This article is organized as follows.
-The following section introduces the relevant basics of graph theory urban studies, focusing on discussions in the research literature about street network planarity.
-The next section introduces the methods used to acquire and analyze the street networks in this study.
-Then this article presents the results of this analysis before concluding with a discussion of their ramifications for street network research and urban form studies.
+The following section introduces the relevant basics of graph theory in urban studies, focusing on discussions in the research literature about street network planarity.
+The next section discusses the methods used to acquire and analyze the street networks in this study.
+Then we present the results of this analysis before concluding with a discussion of their ramifications for street network research and urban form studies.
 
-\section{Background}
+\section{Background and Motivation}
 
@@
 A \emph{planar} graph can be drawn on a two-dimensional plane without any of its edges crossing each other, except where they intersect at nodes.
-If the graph \emph{cannot} be drawn or redrawn to meet this criterion, then the graph is nonplanar \citep{trudeau_introduction_1994}.
-Street networks are embedded in space, which provides them with geometry --- such as geographical coordinates, lengths, and area --- alongside their topology.
+If the graph \emph{cannot} be drawn --- or redrawn --- to meet this criterion, then the graph is nonplanar \citep{trudeau_introduction_1994}.
+Street networks are embedded in space, which provides them with geometry --- such as geographical coordinates, lengths, and areas --- along with their topology.
 
@@
 
-This creates a minor wrinkle when we consider planarity: we must distinguish between a graph's topological planarity, which we shall refer to as \emph{formal planarity}, and the planarity of its particular spatial embedding, which we shall refer to as \emph{spatial planarity}.
+This creates a minor wrinkle when we consider planarity: we must distinguish between a graph's topological planarity, which we refer to as \emph{formal planarity}, and the planarity of its particular spatial embedding, which we refer to as \emph{spatial planarity}.
 A street network might be spatially nonplanar due to its embedding in space (i.e., it contains overpasses or underpasses in the real-world), but it could still be formally planar.
-For example, if we \enquote{redraw} the graph by moving its nodes and edges around in space without changing how they are connected to one another (i.e., altering its geometry without altering its topology), there may exist some spatial embedding that prevents edges crossing anywhere but nodes \citep[for a more comprehensive discussion see][pp.~6--10]{barthelemy_morphogenesis_2017}.
+For example, if we \enquote{redraw} the graph by moving its nodes and edges around in space without changing how they are connected to one another (i.e., altering its geometry without altering its topology), there may exist some spatial embedding that prevents edges crossing anywhere but at nodes \citep[for a comprehensive discussion see][pp.~6--10]{barthelemy_morphogenesis_2017}.
 In such a case, the street network is formally planar from a topological perspective, but its real-world embedding is spatially nonplanar.
@@
 However, we might call it \emph{approximately} planar because its nonplanar edge crossings are relatively uncommon.
-Approximate planarity constrains this nonplanar spatial network such that it does not exhibit certain characteristics found among nonplanar aspatial graphs (e.g., small-world effects or power-law distributed node degrees).
+Approximate planarity constrains nonplanar spatial networks such that they do not exhibit certain characteristics found among nonplanar aspatial graphs (e.g., small-world effects or power-law distributed node degrees).
 
@@
 Forcing planarity on a nonplanar street network creates artificial nodes at bridges and tunnels, which breaks routing.
-As \citet[p.~6]{kwan_review_1996} puts it, \enquote{the difficulty in accurately representing overpasses or underpasses may lead to problems when running various routing algorithms (e.g.
+As \citet[p.~6]{kwan_review_1996} put it, \enquote{the difficulty in accurately representing overpasses or underpasses may lead to problems when running various routing algorithms (e.g.
 recommending that a traveler make a left-turn at an intersection that proves to be an overpass)}.
-Second, and accordingly, intersection counts and density will be overestimated.
-Similarly, edge lengths will be underestimated due to these artificial breakpoints.
+Second, and accordingly, intersection counts and density will be overestimated by the presence of these pseudonodes.
+Consequently, edge lengths will be underestimated due to these artificial breakpoints in the middle of street segments.
 Finally, this bias would likely behave inconsistently across different kinds of cities and street network types.
 
-Given these issues, what do \emph{approximately planar} and \emph{well-modeled} mean for street networks?
-How close is close enough for a planar graph to competently model a mathematically nonplanar street network?
-Do the biases of planar models remain consistent across geographies or do they misrepresent different places to different degrees?
-If street networks are in general approximately planar, then how can we measure how planar or nonplanar a given street network is?
+Given these issues, what do \emph{approximately planar} and \emph{well-modeled} mean for street network research?
+How close is \enquote{close enough} for a planar graph to competently model a formally nonplanar street network?
+Do the biases of planar models behave consistently across geographies and development eras or do they misrepresent different cities to different degrees?
+And if street networks are at least generally approximately planar, then how can we measure just how planar or nonplanar a given street network is?
 
-The graph theory literature offers some measures of how \enquote{far off} a nonplanar graph $G$ is from being planar, such as its crossing number --- the minimum number of edge crossings of any drawing of $G$ --- and the skewness --- the minimum number of edges that must be removed from $G$ to produce a planar graph \citep{liebers_planarizing_2001,szekely_successful_2004,chimani_non-planar_2009}.
+The graph theory literature offers some measures of how \enquote{far off} a nonplanar graph $G$ is from being planar, including its \emph{crossing number} --- the minimum number of edge crossings of any drawing of $G$ --- and its \emph{skewness} --- the minimum number of edges that must be removed from $G$ to produce a planar graph \citep{liebers_planarizing_2001,szekely_successful_2004,chimani_non-planar_2009}.
 However, these measures are imperfect and hard to compute \citep{chimani_vertex_2012,szekely_successful_2004}.
@@
 For cities, such as Moscow, with a newer commercial central business district (CBD) that lies apart from an \enquote{old town} center, we use the modern CBD as the city center.
-We look separately at their drivable and their walkable street networks.
+We look separately at each city's drivable and walkable street networks.
 
@@
 
-OpenStreetMap graphs contain many interstitial nodes in the middle of street segments (forming an expansion graph) to allow streets to curve through space via a series of straight-line approximations.
+OpenStreetMap's raw data contain many interstitial nodes in the middle of street segments (forming an expansion graph) to allow streets to curve through space via a series of straight-line approximations.
 OSMnx automatically simplifies each graph's topology to retain nodes only at intersections and dead-ends, while faithfully retaining each edge's true spatial geometry.
@@
 
-The first is a formal test of aspatial planarity using the algorithm described by \citet{boyer_subgraph_2012}.
+The first is an aspatial test of formal planarity using the algorithm described by \citet{boyer_subgraph_2012}.
 That is, is it possible to rearrange the graph's nodes and edges in space --- while preserving its topology --- so that edges cross only at nodes?
 This binary true/false indicator tells us if the graph is formally planar, ignoring its real-world spatial embedding.
-However, street networks are spatially embedded, so we develop the following two continuous measures to assess the \enquote{extent} to which they are planar.
+However, street networks are spatially embedded, so we develop the following two continuous indicators to assess the \enquote{extent} to which they are planar.
 
@@
 It indicates the extent to which planarity overstates intersections and connectivity in each street network.
-A truly planar network with no bridges or tunnels will thus have an SPR score of 1.0, while lower values suggest the extent to which the network is nonplanar.
+A truly planar network with no bridges or tunnels will thus have an SPR score of 1.0, while lower values indicate the extent to which the network is nonplanar.
 
@@
 It indicates the extent to which planarity understates edge lengths in each street network.
-A truly planar street network with no bridges or tunnels will thus have an ELR score of 1.0, while lower values suggest the extent to which the network is nonplanar.
+A truly planar street network with no bridges or tunnels will thus have an ELR score of 1.0, while lower values indicate the extent to which the network is nonplanar.
 
@@
 First we analyze the entire city of Oakland.
-Then we recreate the aforementioned methodology by randomly sampling 100 points within the city limits and analyzing the square-mile street network centered on each.
+Then we recreate the aforementioned methodology by randomly sampling 100 points within the city limits and analyzing the square-mile street networks centered on each.
 The resulting statistical dispersion of planarity indicators demonstrates the extent to which agglomerating an entire city's neighborhoods into a single graph may obscure neighborhood-scale effects and characteristics.
@@
 \subsection{Worldwide Cities}
+
+\begin{table*}[htbp]
+	\centering
+	\caption{Planarity measures for the central street networks of 50 cities worldwide (SPR = spatial planarity ratio; ELR = edge length ratio; Planar = whether street network passed the formal test of planarity).}
+	\label{tab:world_cities}
+	\input{table_world_cities.tex}
+\end{table*}
+
+Table \ref{tab:world_cities} presents the results of the analysis of the 50 cities worldwide.
 
@@
 
-Mogadishu is the only city to show perfect planarity across all three indicators and both network types.
-All three Italian cities show perfect planarity in their drivable networks, but not in their walkable networks.
-
 \begin{figure*}[htbp]
@@
 
-\begin{table*}[htbp]
-\centering
-\caption{Planarity measures for the central street networks of 50 cities worldwide (SPR = spatial planarity ratio; ELR = edge length ratio; Planar = whether street network passed the formal test of planarity).}
-\label{tab:world_cities}
-\input{table_world_cities.tex}
-\end{table*}
+Figure \ref{fig:world_map_bw} shows the distribution of SPR values around the world.
+While nearly every European city is in the highest subset, indicating their networks are more planar, most American cities are in the lowest subset, indicating their networks are less planar.
 
-\begin{figure*}[htbp]
-    \center
+Mogadishu is the only city studied that shows perfect planarity across all three indicators and both network types.
+All three Italian cities show perfect planarity in their drivable networks, but not in their walkable networks.
+
+\begin{figure}[htbp]
 	\includegraphics[width=0.5\textwidth]{regression_phi_split.png}
-    \caption{Log-log plot of ELR vs SPR with simple regression lines: drive $r^2=0.98$ and walk $r^2=0.99$.}
+	\caption{Log-log plot of ELR vs SPR with simple regression lines.}
 	\label{fig:regression_split}
-\end{figure*}
+\end{figure}
+
+Figure \ref{fig:regression_split} shows the relationship between the SPR and ELR indicators across all 50 cities for both network types, as a log-log plot.
+The indicators' log-log relationship is linear, positive, and very strong (drivable $r^2=0.98$ and walkable $r^2=0.99$).
 

### 2018-01-14 manuscript.tex
update references

@@
 OpenStreetMap is a collaborative online mapping platform commonly used by researchers because of its good worldwide coverage \citep{haklay_how_2010,jokar_arsanjani_openstreetmap_2015}.
-OSMnx is a Python-based software tool that allows us to automatically download a street network from OpenStreetMap for any study site in the world, and automatically process it into a length-weighted nonplanar directed graph (citation redacted).
+OSMnx is a Python-based software tool that allows us to automatically download a street network from OpenStreetMap for any study site in the world, and automatically process it into a length-weighted nonplanar directed graph \citep{boeing_osmnx:_2017}.
 It differentiates between walkable and drivable routes in the circulation network based on individual elements' metadata that describe how the route may be used.

### 2018-01-16 manuscript.tex
edits

@@
 In the urban studies literature, street networks are typically referred to as \enquote{planar} or \enquote{approximately planar} (see Table \ref{tab:planar_quotes}), meaning that they can be well-modeled by a flat two-dimensional model that inherently forbids overpasses or underpasses.
-Of course in the real-world, urban street networks are embedded in three-dimensional space and often feature grade separation, bridges, and tunnels.
-This leads to two intertwined questions underexplored in the literature today.
-First, how well do planar graphs model urban street networks?
-Second, what does the extent to which a network is nonplanar tell us about the city's infrastructure and development?
+Of course in the real world, urban street networks are embedded in three-dimensional space and often feature grade separation, bridges, and tunnels.
+This leads to two intertwined questions.
+First, how well do these planar graphs model urban street networks?
+Second, what does the extent to which a network is nonplanar tell us about a city's infrastructure and development?
 
@@
 This creates a minor wrinkle when we consider planarity: we must distinguish between a graph's topological planarity, which we refer to as \emph{formal planarity}, and the planarity of its particular spatial embedding, which we refer to as \emph{spatial planarity}.
-A street network might be spatially nonplanar due to its embedding in space (i.e., it contains overpasses or underpasses in the real-world), but it could still be formally planar.
-For example, if we \enquote{redraw} the graph by moving its nodes and edges around in space without changing how they are connected to one another (i.e., altering its geometry without altering its topology), there may exist some spatial embedding that prevents edges crossing anywhere but at nodes \citep[for a comprehensive discussion see][pp.~6--10]{barthelemy_morphogenesis_2017}.
+For example, a street network might be spatially nonplanar due to its embedding in space (i.e., it contains overpasses or underpasses in the real world), but it could still be formally planar.
+If we \enquote{redraw} the graph by moving its nodes and edges around in space without changing how they are connected to one another (i.e., altering its geometry without altering its topology), there may exist some spatial embedding that prevents edges crossing anywhere but at nodes \citep[for a comprehensive discussion see][pp.~6--10]{barthelemy_morphogenesis_2017}.
 In such a case, the street network is formally planar from a topological perspective, but its real-world embedding is spatially nonplanar.
@@
 Do the biases of planar models behave consistently across geographies and development eras or do they misrepresent different cities to different degrees?
-And if street networks are at least generally approximately planar, then how can we measure just how planar or nonplanar a given street network is?
+And if street networks are at least generally approximately planar, how can we measure just how planar or nonplanar a given street network is?
 
@@
 However, these measures are imperfect and hard to compute \citep{chimani_vertex_2012,szekely_successful_2004}.
+They also fail to account for the size or density of a spatial network.
 In his discussion of road networks and approximate spatial planarity, \citet[p.~133]{newman_networks:_2010} argues that \enquote{no widely accepted metric for degree of planarity has emerged,} and calls for the development of better indicators.
@@
 This allows us to consistently examine central urban street networks without being swamped by metropolitan-scale variation or the idiosyncrasies of individual municipalities' spatial extents.
-For cities, such as Moscow, with a newer commercial central business district (CBD) that lies apart from an \enquote{old town} center, we use the modern CBD as the city center.
+For cities such as Moscow, with a newer commercial central business district (CBD) that lies apart from an \enquote{old town} center, we use the modern CBD as the city center.
 We look separately at each city's drivable and walkable street networks.
@@
 Once we have acquired and prepared our networks, we calculate three measures of planarity.
-
 The first is an aspatial test of formal planarity using the algorithm described by \citet{boyer_subgraph_2012}.
@@
 It represents the ratio of nonplanar intersections (i.e., non-dead-end nodes in the nonplanar, three-dimensional, spatially-embedded graph) to planar intersections (i.e., line crossings in the planar, two-dimensional, spatially-embedded graph).
-It indicates the extent to which planarity overstates intersections and connectivity in each street network.
+This indicates the extent to which planarity overstates intersections and connectivity in each street network.
 A truly planar network with no bridges or tunnels will thus have an SPR score of 1.0, while lower values indicate the extent to which the network is nonplanar.
@@
 It represents the ratio of the mean edge length in the planar graph to the mean edge length in the nonplanar graph.
-It indicates the extent to which planarity understates edge lengths in each street network.
+This indicates the extent to which planarity understates edge lengths in each street network.
 A truly planar street network with no bridges or tunnels will thus have an ELR score of 1.0, while lower values indicate the extent to which the network is nonplanar.
@@
 \begin{figure}[htbp]
-	\includegraphics[width=0.5\textwidth]{regression_phi_split.png}
+	\includegraphics[width=0.48\textwidth]{regression_phi_split.png}
 	\caption{Log-log plot of ELR vs SPR with simple regression lines.}
@@
 We find that the street network of the entire city of Oakland is formally nonplanar when subjected to a formal mathematical test.
-In terms of the nonplanarity indicators, the city as a whole has a SPR score of 0.918, indicating that it is 91.8\% planar, and a ELR score of 0.936.
+In terms of the nonplanarity indicators, the city as a whole has an SPR score of 0.918, indicating that it is 91.8\% planar, and an ELR score of 0.936.
 This suggests that the planar representation of Oakland's drivable street network overstates the number of intersections (and thus, the network's connectivity) by 8.9\% city-wide (i.e., $1 - 1 / 0.918$ and understates the average edge length by 6.4\% city-wide (i.e., $1 - 0.936$).
@@
 
-The urban studies and urban physics literatures frequently use planar graphs to urban street networks.
-This study demonstrated that although in limited circumstances these models may be accurate, they behave inconsistently across different kinds of cities by misrepresenting connectivity, accessibility, routing, intersection counts and densities, and street segment lengths.
+Urban researchers frequently use planar graphs to model street networks.
+This study demonstrated that, although in limited circumstances these models may be accurate, they behave inconsistently across different kinds of cities by misrepresenting connectivity, accessibility, routing, intersection counts and densities, and street segment lengths.
 It found x, y, and z.
-It also demonstrated how these indicators can be used to characterize the type of urbanization, in particular its transportation infrastructure's three-dimensionality, of different cities.
+It also demonstrated how these indicators can be used to characterize the type of urbanization, in particular transportation infrastructure's three-dimensionality, in different cities.
 Future research can explore this latter finding, as it likely correlates with other measures of urbanization, development, and era.

### 2018-01-19 manuscript.tex
discussion section, revisions throughout

@@
 
-In the urban studies literature, street networks are typically referred to as \enquote{planar} or \enquote{approximately planar} (see Table \ref{tab:planar_quotes}), meaning that they can be well-modeled by a flat two-dimensional model that inherently forbids overpasses or underpasses.
+In the urban studies literature, street networks are typically referred to as \enquote{planar} or \enquote{approximately planar} (see Table \ref{tab:planar_quotes}), meaning that they can be well-modeled by a flat two-dimensional model that inherently precludes overpasses or underpasses.
 Of course in the real world, urban street networks are embedded in three-dimensional space and often feature grade separation, bridges, and tunnels.
@@
 If the graph \emph{cannot} be drawn --- or redrawn --- to meet this criterion, then the graph is nonplanar \citep{trudeau_introduction_1994}.
-Street networks are embedded in space, which provides them with geometry --- such as geographical coordinates, lengths, and areas --- along with their topology.
+Street networks are embedded in space, which provides them with geometry --- such as geographical coordinates, lengths, areas, and angles --- along with their topology.
 
@@
 \centering
-\caption{Recent statements in the urban studies and urban physics literatures regarding the representation of street networks as planar graphs.}
+\caption{Recent statements in the research literature regarding the representation of street networks as planar graphs.}
 \label{tab:planar_quotes}
@@
 For example, a street network might be spatially nonplanar due to its embedding in space (i.e., it contains overpasses or underpasses in the real world), but it could still be formally planar.
-If we \enquote{redraw} the graph by moving its nodes and edges around in space without changing how they are connected to one another (i.e., altering its geometry without altering its topology), there may exist some spatial embedding that prevents edges crossing anywhere but at nodes \citep[for a comprehensive discussion see][pp.~6--10]{barthelemy_morphogenesis_2017}.
+If we \enquote{redraw} the graph by moving its nodes and edges around in space without changing how they are connected to one another (i.e., altering its geometry without altering its topology), there may exist some spatial embedding that prevents edges crossing anywhere but at nodes \citep[p.~6]{barthelemy_morphogenesis_2017}.
 In such a case, the street network is formally planar from a topological perspective, but its real-world embedding is spatially nonplanar.
@@
 
-In the urban studies and urban physics literature, street networks are commonly referred to as planar graphs.
+In the urban studies literature, street networks are commonly referred to as planar graphs.
 Table \ref{tab:planar_quotes} presents a survey of statements and reasoning around this claim.
@@
 Planar graphs offer computational simplicity and tractability.
-They enable easy polygonal spatial analysis of city blocks and form \citep{fohl_non-planar_1996} as well as the Euler characteristic.
+They enable easy polygonal spatial analysis of city blocks and form \citep{fohl_non-planar_1996,barthelemy_paths_2017} as well as the Euler characteristic.
 In mathematics, there is a bijection between planar graphs and trees, and classifying planar graphs presents a trivial problem \citep{louf_typology_2014}.
@@
 Other authors have commented on this characteristic of street networks.
-\citet[p.~199]{jiang_object-oriented_2010} explain that \enquote{Quite often the transportation network has overpasses and underpasses that require a non-planar network representation.} \citet[p.~1258]{fischer_spatial_2014} explains that \enquote{For many infrastructure networks, {[planarity]} is approximately true, although bridges and tunnels in ground-transport networks are an obvious (but generally minor) exception.} Twenty years ago, \citep[p.~18]{fohl_non-planar_1996} claimed, \enquote{The most commonly used data model for transportation networks is the fully intersected, planar data model} and called for a nonplanar model to better represent truly nonplanar spatial networks.
+\citet[p.~199]{jiang_object-oriented_2010} note that \enquote{quite often the transportation network has overpasses and underpasses that require a non-planar network representation.} \citet[p.~1258]{fischer_spatial_2014} explains that \enquote{for many infrastructure networks, {[planarity]} is approximately true, although bridges and tunnels in ground-transport networks are an obvious (but generally minor) exception.} However, the presence of such nonplanar elements can vex models.
 \enquote{The planar network data model has received widespread acceptance and use.
@@
 One major problem is caused by the planar embedding requirement...
-intersections at grade cannot be distinguished from intersections with an overpass or underpass that do not cross at grade.} \citep[p.~395]{fischer_gis_2004}
+intersections at grade cannot be distinguished from intersections with an overpass or underpass that do not cross at grade} \citep[p.~395]{fischer_gis_2004}.
+Twenty years ago, \citet[p.~18]{fohl_non-planar_1996} called for a nonplanar model to better represent truly nonplanar spatial networks.
 
 If a planar graph models a street network poorly, it could do so in multiple ways.
-Forcing planarity on a nonplanar street network creates artificial nodes at bridges and tunnels, which breaks routing.
-As \citet[p.~6]{kwan_review_1996} put it, \enquote{the difficulty in accurately representing overpasses or underpasses may lead to problems when running various routing algorithms (e.g.
+Forcing planarity on a nonplanar street network creates artificial nodes in the graph at bridges and tunnels, which breaks routing.
+As \citet[p.~6]{kwan_review_1996} describe it, \enquote{the difficulty in accurately representing overpasses or underpasses may lead to problems when running various routing algorithms (e.g.
 recommending that a traveler make a left-turn at an intersection that proves to be an overpass)}.
-Second, and accordingly, intersection counts and density will be overestimated by the presence of these pseudonodes.
-Consequently, edge lengths will be underestimated due to these artificial breakpoints in the middle of street segments.
-Finally, this bias would likely behave inconsistently across different kinds of cities and street network types.
+Intersection counts and densities will be overestimated in the presence of these false nodes.
+Consequently, edge lengths will be underestimated due to these artificial breakpoints splitting up street segments.
+Finally, this bias would likely behave inconsistently across different kinds of cities and street network types based on the extents to which they are planar.
 
@@
 However, these measures are imperfect and hard to compute \citep{chimani_vertex_2012,szekely_successful_2004}.
-They also fail to account for the size or density of a spatial network.
-In his discussion of road networks and approximate spatial planarity, \citet[p.~133]{newman_networks:_2010} argues that \enquote{no widely accepted metric for degree of planarity has emerged,} and calls for the development of better indicators.
+They also fail to correct for the size or density of a spatial network.
+In his discussion of road networks and approximate spatial planarity, \citet[p.~133]{newman_networks:_2010} argues that due to these drawbacks \enquote{no widely accepted metric for degree of planarity has emerged,} and calls for the development of better indicators.
 
-Such metrics would be particularly useful for street networks, as the extent to which a network is (or is not) planar can characterize the nature of its circulation infrastructure and urban form.
+Such measures would be particularly useful for street networks, as the extent to which a network is (or is not) planar can characterize the nature of its circulation infrastructure and urban form.
 For instance, late 20th-century freeway-oriented American cities might exhibit lower planarity than older walkable European cities.
@@
 This allows us to consistently examine central urban street networks without being swamped by metropolitan-scale variation or the idiosyncrasies of individual municipalities' spatial extents.
-For cities such as Moscow, with a newer commercial central business district (CBD) that lies apart from an \enquote{old town} center, we use the modern CBD as the city center.
+The 50 sampled cities span Africa, Asia, Australia, Europe, and North and South America.
 We look separately at each city's drivable and walkable street networks.
+For cities such as Moscow, with a newer commercial central business district (CBD) that lies apart from an \enquote{old town} center, we take the modern CBD as the city center.
 
@@
 The first is an aspatial test of formal planarity using the algorithm described by \citet{boyer_subgraph_2012}.
-That is, is it possible to rearrange the graph's nodes and edges in space --- while preserving its topology --- so that edges cross only at nodes?
+This assesses if it is possible to rearrange the graph's nodes and edges in space, while preserving its topology, so that edges cross only at nodes.
 This binary true/false indicator tells us if the graph is formally planar, ignoring its real-world spatial embedding.
-However, street networks are spatially embedded, so we develop the following two continuous indicators to assess the \enquote{extent} to which they are planar.
+However, street networks \emph{are} spatially embedded, so we have developed the following two indicators to assess the \enquote{extent} to which they are planar.
 
 The second measure is the spatial planarity ratio (SPR).
-It represents the ratio of nonplanar intersections (i.e., non-dead-end nodes in the nonplanar, three-dimensional, spatially-embedded graph) to planar intersections (i.e., line crossings in the planar, two-dimensional, spatially-embedded graph).
+It represents the ratio of nonplanar intersections $i_n$ (i.e., non-dead-end nodes in the nonplanar, three-dimensional, spatially-embedded graph) to planar intersections $i_p$ (i.e., line crossings in the planar, two-dimensional, spatially-embedded graph):
+
+\begin{equation}
+	\label{eq:spr}
+	ELR = \frac{i_n}{i_p}
+\end{equation}
+
 This indicates the extent to which planarity overstates intersections and connectivity in each street network.
-A truly planar network with no bridges or tunnels will thus have an SPR score of 1.0, while lower values indicate the extent to which the network is nonplanar.
+In other words, the SPR tells us what proportion of the line crossings in the planar graph are true intersections in the nonplanar graph.
+A truly planar network with no bridges or tunnels will thus have an SPR score of 1.0, while lower values indicate the percentage extent to which the network is planar.
+From the SPR, we can further calculate by what percentage the planar graph overstates intersection counts as:
 
-The third measure is the edge length ratio (ELR).
-It represents the ratio of the mean edge length in the planar graph to the mean edge length in the nonplanar graph.
-This indicates the extent to which planarity understates edge lengths in each street network.
-A truly planar street network with no bridges or tunnels will thus have an ELR score of 1.0, while lower values indicate the extent to which the network is nonplanar.
+\begin{equation}
+	\label{eq:spr_overstates}
+	1 - \frac{1}{SPR}
+\end{equation}
 
-Finally, we explore how these indicators might vary across a single municipality.
+The third measure is the edge length ratio (ELR). It represents the ratio of the mean edge length in the planar graph $l_p$ to the mean edge length in the nonplanar graph $l_n$:
+
+\begin{equation}
+	\label{eq:elr}
+	ELR = \frac{l_p}{l_n}
+\end{equation}
+
+This indicates the extent to which planarity understates edge lengths in each street network by fragmenting street segments at overpasses or underpasses.
+A truly planar street network with no bridges or tunnels will thus have an ELR score of 1.0, while lower values indicate the percentage extent to which the network is planar.
+From the ELR, we can further calculate by what percentage the planar graph understates the average edge length as:
+
+\begin{equation}
+	\label{eq:elr_understates}
+	1 - ELR
+\end{equation}
+
+Finally, we explore how these three indicators vary across a single city.
 To do so we analyze the drivable street network of Oakland, California, a reasonably representative midsized American city with a variety of urban form types from gridded street patterns in its flatlands, to winding culs-de-sac in its hills, to freeways and dense blocks around its downtown.
@@
 Then we recreate the aforementioned methodology by randomly sampling 100 points within the city limits and analyzing the square-mile street networks centered on each.
-The resulting statistical dispersion of planarity indicators demonstrates the extent to which agglomerating an entire city's neighborhoods into a single graph may obscure neighborhood-scale effects and characteristics.
+The resulting statistical dispersion of planarity demonstrates the extent to which analyzing an entire city's neighborhoods as a single graph may obscure neighborhood-scale planarity characteristics.
 
@@
 
-\subsection{Worldwide Cities}
-
 \begin{table*}[htbp]
 	\centering
-	\caption{Planarity measures for the central street networks of 50 cities worldwide (SPR = spatial planarity ratio; ELR = edge length ratio; Planar = whether street network passed the formal test of planarity).}
+	\caption{Planarity measures for the central street networks of 50 cities worldwide (Planar = whether street network passed the formal test of planarity; SPR = spatial planarity ratio; ELR = edge length ratio).}
 	\label{tab:world_cities}
@@
 
-Table \ref{tab:world_cities} presents the results of the analysis of the 50 cities worldwide.
-
-Among drivable street networks, only x\% pass the formal test of planarity.
-On average, they are only x\% planar by the SPR measure and x\% planar by the ELR measure.
-SPR values range from a high of 1.0 in x cities to a low of 57\% in Moscow.
-ELR values range from a high of 1.0 in x cities to a low of 64\% in Los Angeles.
-
-Among walkable street networks, only x\% pass the formal test of planarity.
-On average, they are only x\% planar by the SPR measure and x\% planar by the ELR measure.
-SPR values range from a high of 1.0 in x cities to a low of 74\% in Atlanta.
-ELR values range from a high of 1.0 in x cities to a low of 72\% in Atlanta.
-
 \begin{figure*}[htbp]
@@
 	\includegraphics[width=\textwidth]{world_map_phi_bw.png}
-	\caption{Map of world cities in Table \ref{tab:world_cities} grouped by SPR score tertiles (lower scores mean less planar).}
+	\caption{Map of world cities from Table \ref{tab:world_cities} grouped by SPR score terciles (lower scores mean less planar).}
 	\label{fig:world_map_bw}
@@
 
-Figure \ref{fig:world_map_bw} shows the distribution of SPR values around the world.
-While nearly every European city is in the highest subset, indicating their networks are more planar, most American cities are in the lowest subset, indicating their networks are less planar.
-
-Mogadishu is the only city studied that shows perfect planarity across all three indicators and both network types.
-All three Italian cities show perfect planarity in their drivable networks, but not in their walkable networks.
-
 \begin{figure}[htbp]
 	\includegraphics[width=0.48\textwidth]{regression_phi_split.png}
-	\caption{Log-log plot of ELR vs SPR with simple regression lines.}
+	\caption{Log-log plot of ELR vs SPR, by network type, with simple regression lines.}
 	\label{fig:regression_split}
@@
 
+Table \ref{tab:world_cities} shows the SPR, ELR, and results of the formal planarity test for the walkable and the drivable networks for each of the 50 cities.
+Among the drivable street networks, only 20\% are formally planar.
+On average, they are 88.5\% planar by the SPR measure and 90\% planar by the ELR measure.
+The individual SPR values indicate that spatial planarity ranges from a high of 100\% in seven of these cities to a low of 57\% in Moscow.
+The ELR values indicate that spatial planarity ranges from a high of 100\% in seven of these cities to a low of 63.5\% in Los Angeles.
+
+Among walkable street networks, only 10\% are formally planar.
+On average, they are 92\% planar by the SPR measure and 91.5\% planar by the ELR measure.
+The individual SPR values indicate that spatial planarity ranges from a high of 100\% in two cities to a low of 67\% in Shanghai.
+The ELR values indicate that spatial planarity ranges from a high of 100\% in two cities to a low of 66\% in Shanghai.
+Fewer walkable than drivable networks are formally planar, but on average these walking networks are slightly more spatially planar than the driving networks.
+
+Not all formally planar street networks are spatially planar.
+For example, Toronto's drivable network is formally planar but only 93\% and 96\% spatially planar according to its SPR and ELR measures.
+In total, three drivable networks (Toronto, Jakarta, and Copenhagen) and three walkable networks (Dallas, Delhi, and Bologna) are formally planar but spatially nonplanar to various extents.
+About a third of the cities studied demonstrate spatial planarity of 97\% or higher according to the SPR, suggesting that they are \enquote{approximately} planar.
+However, another third of the cities studied are less than 87\% planar.
+Dallas, Los Angeles, and Moscow have SPR values below 60\%, suggesting planar graphs poorly model these city centers.
+Moreover, planar graphs overstate the intersection counts in these three networks by 67\%, 72\%, and 74\% respectively (Equation \ref{eq:spr_overstates}).
+
+Mogadishu is the only city studied that demonstrates perfect planarity across all three indicators for both network types.
+All three Italian cities demonstrate perfect planarity in their centers' drivable networks, but not in their walkable networks.
+The extent of planarity is not consistent across network types: Dallas's walkable SPR is 61\% greater than its drivable SPR, while Geneva's drivable SPR is 19\% greater than its walkable SPR.
+Figure \ref{fig:world_map_bw} shows the distribution of SPR values around the world.
+While nearly every European city is in the highest tercile, indicating their networks are more planar, most American cities are in the lowest tercile, indicating their networks are more nonplanar.
+
 Figure \ref{fig:regression_split} shows the relationship between the SPR and ELR indicators across all 50 cities for both network types, as a log-log plot.
 The indicators' log-log relationship is linear, positive, and very strong (drivable $r^2=0.98$ and walkable $r^2=0.99$).
-
-\subsection{A Closer Look: Oakland, California}
-
-We find that the street network of the entire city of Oakland is formally nonplanar when subjected to a formal mathematical test.
-In terms of the nonplanarity indicators, the city as a whole has an SPR score of 0.918, indicating that it is 91.8\% planar, and an ELR score of 0.936.
-This suggests that the planar representation of Oakland's drivable street network overstates the number of intersections (and thus, the network's connectivity) by 8.9\% city-wide (i.e., $1 - 1 / 0.918$ and understates the average edge length by 6.4\% city-wide (i.e., $1 - 0.936$).
-However, these indicators' values vary across the city.
-
-To explore this statistical variation, we examined 100 square-mile samples of Oakland's drivable street network.
-Table \ref{tab:samples_city} presents descriptive statistics of these planarity indicators.
+This strong correlation tells us that these two indicators unsurprisingly provide nearly redundant statistical information about the extent to which a network is planar.
+However, each assesses different implications of this bias for measuring the urban form.
 
@@
 	\centering
-\caption{Descriptive statistics of planarity indicators across 100 random samples in Oakland, California's drivable network.}
+	\caption{Summary statistics of planarity indicators across 100 random samples of Oakland, California's drivable network.}
 	\label{tab:samples_city}
@@
 
+Finally, we examine how these measures behave across an entire city.
+We find that Oakland's city-wide street network is formally nonplanar.
+The city has an SPR score of 91.8\% and an ELR score of 93.6\%.
+This suggests that the planar representation of Oakland's drivable street network overstates the number of intersections --- and thus, the network's connectivity --- by 8.9\% city-wide (Equation \ref{eq:spr_overstates}) and understates the average edge length by 6.4\% city-wide (Equation \ref{eq:elr_understates}).
+
+However, these indicators' values vary across the city.
+To explore this statistical variation, Table \ref{tab:samples_city} presents summary statistics of these planarity indicators across 100 square-mile samples of Oakland's drivable street network.
+Our samples' mean SPR and ELR scores are reasonably close to the city-wide values.
+However, the samples range from spatial planarity lows of 56.9\% (SPR) and 63.7\% (ELR) up to highs of 100\%.
+67\% of the samples pass the formal planarity test.
+However, 63\% of the samples are at least somewhat spatially nonplanar (i.e., with SPR less than 1).
+
 \section{Discussion}
-
-Our findings suggest that the street networks in the centers of most major cities are formally and spatially nonplanar.
-However, this depends on the scale of measurement: across an entire city there is likely to be some underpass or overpass somewhere, while individual neighborhoods or small suburbs might be formally planar.
-The type and era of urbanization is another factor.
-Old medieval European towns or informal settlements in the global south contain fewer grade-separated roads (and thus greater planarity) than 20th-century American or 21st-century Chinese metropolises do.
-This is a result of prevailing transportation technologies when the urban form was built up, as well as terrain, costs, wealth, culture, and politics.
-
-Street networks are frequently nonplanar in the formal sense because they are spatially embedded in three dimensions --- not two --- and have a z-coordinate along with their x and y.
-Because they are mostly planar, typically with only a few overpasses or underpasses, they could often be described accurately as \emph{approximately planar}.
-However, claiming that urban street networks broadly are planar misrepresents them in several ways:
-
-\begin{enumerate}
-	\item{Forces false nodes where grade-separated edges cross.}
-	\item{Accordingly, underestimates average edge length (a proxy for street segment lengths and block sizes).}
-	\item{Misrepresents connectivity for routing, accessibility analysis, and other topological studies.}
-\end{enumerate}
-
-This is a bigger problem in Los Angeles than in Florence, but even in Florence, planar overcounts intersections around the \textit{sottopassaggio}, or pedestrian subway, near its central train station, Stazione di Santa Maria Novella.
-This is a pedestrian walkway, not a drivable road, but in a walkable town like Florence, we would surely want to model pedestrian access as part the street network, not drivable roads only.
-In this case, the nonplanar public pedestrian network must be considered.
-
-The problem is consistently and particularly pronounced around the downtowns of North American cities, due to the prevalence of freeways, bridges, and underpasses there.
-Driving networks are affected by these in particular.
-Walking networks are more affected by pedestrian flyovers and subways, as in Florence.
-Even networks of non-freeway, non-pedestrian-only surface streets will often be nonplanar due to bridges or tunnels in hilly neighborhoods or over rivers.
-
-The effect is inconsistent across cities, and across different neighborhoods within individual cities.
-
-Why are they approximately planar? Cost and politics. Expensive to build truly 3-D networks (with z-coordinates as extensive as their x- and y-coordinates) and politically infeasible.
 
@@
 
-Contrary to some of the statements in the urban studies and urban physics literature, our results suggest that this cannot be universally claimed.
+Our findings suggest that the street networks at the centers of most major cities are formally nonplanar.
+However, this depends on the scale of measurement: across an entire city there is likely to be at least one overpass or underpass somewhere, while individual neighborhoods or small towns might be formally planar in their entirety.
+The type and era of urbanization represent another factor.
+Medieval European towns or informal settlements in the global south may contain fewer grade-separated roads --- and thus are more planar --- than 20th-century American or 21st-century Chinese metropolises.
+This is a result of the prevailing transportation technologies when the urban form was developed, as well as the local terrain, wealth, culture, and politics.
+
+Street networks are frequently spatially nonplanar because they are embedded in three dimensions, not two: they have a z-coordinate (elevation) along with their x- and y-coordinates.
+But because they are usually \enquote{mostly} planar (average drivable SPR of 88.5\% and average walkable SPR of 92\%), typically with only a few overpasses or underpasses, they could often be described as \emph{approximately planar}.
+However, claiming that urban street networks universally are \enquote{planar} misrepresents them in several ways:
+
+\begin{enumerate}
+	\item{Intersection counts are overestimated due to false nodes where grade-separated edges cross}
+	\item{Average edge lengths are underestimated}
+	\item{Connectivity is misrepresented for routing, accessibility analysis, and other topological studies}
+\end{enumerate}
+
+Our results demonstrate how this is a bigger problem in Los Angeles than in Florence, but even in Florence the walking network is spatially and formally nonplanar due to the \textit{sottopassaggio} (pedestrian subway) near Stazione di Santa Maria Novella, its central train station.
+Our results suggest that spatial planarity is inconsistent both across cities as well as across different neighborhoods within individual cities.
+
+The problem with planar models is particularly pronounced around the downtowns of North American cities, due to the prevalence of freeways, bridges, and underpasses.
+Drivable networks are affected by these in particular.
+Walkable networks are more affected by pedestrian flyovers and subways, as in Florence.
+However, even networks of non-freeway, non-pedestrian-only surface streets could easily be nonplanar due to bridges or tunnels in hilly neighborhoods or over rivers.
+
+Contrary to some of the statements in the urban studies literature, our results suggest that it cannot be universally claimed that urban street networks are planar graphs.
 But perhaps some of it comes down to vocabulary.
@@
 But this ignores the fact that even residential streets sometimes include bridges and tunnels in hilly neighborhoods, and the fact that our analysis earlier showed that walkable circulation networks in city centers often include pedestrian tunnels and footbridges.
-A graph is not planar because its edges \emph{usually} or \emph{approximately} intersect only at nodes: it is planar because its edges \emph{exclusively} intersect at nodes.
+A graph is not planar because its edges \emph{usually} intersect only at nodes: by definition it is planar because its edges \emph{exclusively} intersect at nodes.
 Thus, street networks are not universally planar graphs.
+
+However, as \citet{newman_networks:_2010} points out, debating the semantics of formal planarity may be missing the point -- more interesting is the \emph{extent} to which a network is planar.
 
@@
 
-As George Box famously stated, \enquote{All models are wrong but some are useful.} Can street networks be simplified to planar graphs and still be usefully well-modeled?
-The answer depends on the study site and on the type of analysis.
-In limited circumstances, where the circulation network and built form exhibit few underpasses, overpasses, or grade-separation, then perhaps yes.
+As George Box famously stated, \enquote{All models are wrong but some are useful.} Even if they are not formally planar, can street networks be simplified to planar graphs and still be usefully well-modeled?
+Our results suggest that the answer depends on the study site and on the type of analysis.
+In limited circumstances, where the circulation network and built form exhibit few (or ideally zero) underpasses, overpasses, or grade-separation, then perhaps yes.
+
 But universally, we cannot answer yes.
-
-Most egregiously, imposing planarity on a nonplanar street network forces pseudonodes at underpasses and overpasses, breaking routing and network-based accessibility modeling.
+Most egregiously, imposing planarity on a nonplanar street network forces false nodes at underpasses and overpasses, breaking routing and network-based accessibility modeling.
 For this reason, nonplanar graphs have been the standard for decades in transportation engineering, real-world traffic assignment models, and routing engines.
 
-But planar graphs are often used in the literature to study urban form and morphology.
+But planar graphs are often used in the literature to characterize urban form and morphology.
 So, aside from routing, do planar graphs offer \emph{useful} models for this type of research?
-Again, only in limited circumstances, including examples such as the drivable networks in the three Italian cities we analyzed, which are both formally and spatially planar.
-The results in Table \ref{tab:world_cities} show how common urban form measures such as intersection counts are overstated by planar models, while average edge lengths (a linear proxy for block size) is consequently understated.
-Moreover, this misrepresentation behaves inconsistently from place to place.
-Figure \ref{fig:world_map_bw} demonstrates how the magnitude of bias varies across cities and eras of urbanization.
+Again, only in limited circumstances, such as the drivable networks in the three Italian cities we analyzed.
+Elsewhere, the results in Table \ref{tab:world_cities} showed how common urban form measures such as intersection counts are overstated by planar models, while average street segment lengths are consequently understated.
+Moreover, this misrepresentation behaves inconsistently from place to place: Figure \ref{fig:world_map_bw} demonstrated how the magnitude of bias varies across cities and modes of urbanization.
+Although planar graphs offer computationally tractable models and allow for the polygonal spatial analysis of urban blocks, they often modeled these street networks poorly compared to nonplanar graphs.
 
-So then why use planar graphs? Tractable. In some cities it doesn't matter (Italy). Polygonal analysis of urban blocks. Otherwise use nonplanar, new tools make it easy.
+\subsection{What can nonplanarity tell us?}
+
+We might repurpose this discrepancy between planar and nonplanar representations of a street network to reveal something useful about urban form.
+The SPR and ELR scores of the major US city centers indicate the greater three-dimensionality of their transportation infrastructure compared to that of the European city centers.
+These US city centers feature more automobile-oriented roadways and are often bounded by grade-separated freeways.
+By contrast, these European city centers feature less three-dimensionality as their streets and paths tend to be at-grade.
+Examining how SPR and ELR change over time in different cities could tell us something about the type and pace of urban development.
+
+It is also interesting that most street networks are formally nonplanar while demonstrating SPR and ELR scores that are only slightly nonplanar.
+Approximate planarity is common across cities because of costs, technology, and politics.
+Why?
+It is expensive to engineer a three-dimensional network with z-axis extents (nearly) as extensive as the x- and y-axes.
+Furthermore, a city fully consumed by overpasses and tunnels would present degraded livability and walkability, making it politically infeasible.
 
@@
 
-Urban researchers frequently use planar graphs to model street networks.
+Urban researchers often use planar graphs to model street networks.
 This study demonstrated that, although in limited circumstances these models may be accurate, they behave inconsistently across different kinds of cities by misrepresenting connectivity, accessibility, routing, intersection counts and densities, and street segment lengths.
-It found x, y, and z.
-It also demonstrated how these indicators can be used to characterize the type of urbanization, in particular transportation infrastructure's three-dimensionality, in different cities.
-Future research can explore this latter finding, as it likely correlates with other measures of urbanization, development, and era.
+In most circumstances, nonplanar graphs provide better models.
+It also demonstrated how these indicators can be used to characterize urbanization in different cities, in particular through the transportation infrastructure's three-dimensionality.
+
+Future research can further explore this latter finding, as it likely correlates with other measures of urbanization, development, and era.
 Finally, future research might examine how nonplanar intersection counts represent true intersections if multiple adjacent edges form multiple graph intersections at a point where only one true intersection exists from an urban design perspective.

### 2018-01-19 table_planar_quotes.tex
discussion section, revisions throughout

\begin{tabular}{ | p{\textwidth} | }

\hline

\enquote{In a planar graph, no links intersect, except by nodes. This feature represents a transportation network well.} \citep[p.~6]{dill_measuring_2004} \\ \hline

\enquote{Street networks are planar graphs composed of junctions and street segments...} \citep[p.~18]{batty_network_2005} \\ \hline

\enquote{The number of long-range connections and the number of edges that can be connected to a single node are limited by the spatial embedding. This is particularly evident in planar networks e.g., those networks forming vertices whenever two edges cross, as urban streets or ant networks of galleries...} \citep[p.~1]{crucitti_centrality_2006} \\ \hline

\enquote{Any of these street networks (SNS) can be described by an embedded planar graph... Street networks are planar graphs and such planarity strongly constrains their heterogeneity...} \citep[pp.~514~\&~521]{buhl_topological_2006} \\ \hline

\enquote{Planar graphs are those graphs forming vertices whenever two edges cross, whereas nonplanar graphs can have edge crossings that do not form vertices. The graphs representing urban street patterns are, by construction, planar...} \citep[p.~3]{cardillo_structural_2006} \\ \hline

\enquote{The connection and arrangement of a road network is usually abstracted in network analysis as a directed planar graph...} \citep[p.~340]{xie_measuring_2007} \\ \hline

\enquote{Urban street patterns form planar networks... The simplest description of the street network consists of a graph whose links represent roads and whose vertices represent road intersections and end points. For these graphs, links intersect essentially only at vertices and are thus planar.} \citep[p.~1]{barthelemy_modeling_2008} \\ \hline

\enquote{Urban street networks as spatial networks are embedded in planar space, which give many constraints.} \citep[p.~1]{hu_topological_2008} \\ \hline

\enquote{...a street network is a strange network when compared to other social or biological networks in the sense that it is embedded in the Euclidian [sic] space and the edges do not cross each other. In graph theory, such a network is called a planar graph.} \citep[p.~259]{masucci_random_2009} \\ \hline

\enquote{...street networks are embedded in space and are planar in nature...} \citep[p.~114]{porta_networks_2010} \\ \hline

\enquote{Roads, rail, and other transportation networks are spatial and to a good accuracy planar networks. For many applications, planar spatial networks are the most important...} \citep[p.~3]{barthelemy_spatial_2011} \\ \hline

\enquote{...urban road systems can be (in good approximation) considered as planar networks, i.e., links cannot \enquote{cross} each other without forming a physical intersection (node) as long as there are no tunnels or bridges... The meaningful definition of link angles requires the presence of a planar network, which is assumed to be the case in urban road systems.} \citep[pp.~563~\&~567]{chan_urban_2011} \\ \hline

\enquote{Road networks are planar graphs consisting of a series of land cells surrounded by street segments.} \citep[p.~3]{strano_elementary_2012} \\ \hline

\enquote{Planar graphs are basic tools for understanding transportation systems embedded in two-dimensional space, in particular urban street networks... As these graphs are embedded in a two-dimensional surface, the
planarity criteria requires that the links do not cross each other.} \citep[p.~1]{masucci_limited_2013} \\ \hline

\enquote{...street networks are essentially planar; in the absence of tunnels and bridges, the streets (the links) cannot cross without generating an intersection or a junction, that is, a node.} \citep[p.~1]{gudmundsson_entropy_2013}. \\ \hline

\enquote{Networks of street patterns belong to a particular class of graphs called planar graphs, that is, graphs whose links cross only at nodes.} \citep[p.~1074]{strano_urban_2013} \\ \hline

\enquote{In city science, planar networks are extensively used to represent, to a good approximation, various infrastructure networks... in particular, transportation networks and more recently streets patterns...} \citep[p.~1]{viana_simplicity_2013} \\ \hline

\enquote{...finding a typology of street patterns essentially amounts to classifying planar graphs...} \citep[p.~2]{louf_typology_2014} \\ \hline

\enquote{...we are dealing with spatial graphs, which tend to be planar...} \citep[p.~2191]{zhong_detecting_2014} \\ \hline

\enquote{Urban transport systems as networks can be represented as planar graphs...} \citep[p.~2]{wang_resilience_2015} \\ \hline

\enquote{In city science, planar networks are extensively used to represent various infrastructure networks. In particular, transportation networks and street patterns...} \citep[p.~257]{barthelemy_paths_2017} \\ \hline

\enquote{In graph theory, a spatial street network is a type of planar graph embedded in Euclidean space.} \citep[p.~168]{law_defining_2017} \\ \hline

\end{tabular}

### 2018-01-19 manuscript.tex
edits

@@
 This article is organized as follows.
-The following section introduces the relevant basics of graph theory in urban studies, focusing on discussions in the research literature about street network planarity.
+The following section introduces the basics of graph theory relevant to urban studies, focusing on discussions in the research literature about street network planarity.
 The next section discusses the methods used to acquire and analyze the street networks in this study.
@@
 Graph theory is the mathematical study of networks \citep{newman_networks:_2010}.
-Graphs can mathematically model real-world networks such as friendships or the Internet as well as spatial networks, such urban street networks \citep{barthelemy_spatial_2011}.
+Graphs can model real-world networks such as friendships, the world wide web, or spatial networks such as urban street networks \citep{barthelemy_spatial_2011}.
 A graph $G$ consists of a set of nodes $N$ connected to one another by a set of edges $E$.
-An edge $uv$ in a directed graph points in one direction from some node $u$ to some node $v$, but an undirected graph's edges all mutually point in both directions.
+An edge $uv$ in a directed graph points in one direction from some node $u$ to some node $v$, but an undirected graph's edges all point mutually in both directions.
 In a street network, the nodes represent intersections and dead-ends, and the (directed) edges represent the street segments that connect them.
@@
 A \emph{planar} graph can be drawn on a two-dimensional plane without any of its edges crossing each other, except where they intersect at nodes.
-If the graph \emph{cannot} be drawn --- or redrawn --- to meet this criterion, then the graph is nonplanar \citep{trudeau_introduction_1994}.
+If the graph cannot be drawn --- or redrawn --- to meet this criterion, then the graph is nonplanar \citep{trudeau_introduction_1994}.
 Street networks are embedded in space, which provides them with geometry --- such as geographical coordinates, lengths, areas, and angles --- along with their topology.
@@
 
-This creates a minor wrinkle when we consider planarity: we must distinguish between a graph's topological planarity, which we refer to as \emph{formal planarity}, and the planarity of its particular spatial embedding, which we refer to as \emph{spatial planarity}.
+This creates a minor wrinkle when we consider planarity: we must distinguish between a graph's mathematical/topological planarity, which we refer to as \emph{formal planarity}, and the planarity of its particular spatial embedding, which we refer to as \emph{spatial planarity}.
 For example, a street network might be spatially nonplanar due to its embedding in space (i.e., it contains overpasses or underpasses in the real world), but it could still be formally planar.
@@
 The indicators' log-log relationship is linear, positive, and very strong (drivable $r^2=0.98$ and walkable $r^2=0.99$).
-This strong correlation tells us that these two indicators unsurprisingly provide nearly redundant statistical information about the extent to which a network is planar.
+This strong correlation tells us that these two indicators unsurprisingly provide redundant statistical information about the extent to which a network is planar.
 However, each assesses different implications of this bias for measuring the urban form.

### 2018-01-24 manuscript.tex
add explanatory figure and edits throughout

@@
 \runninghead{Author Redacted}
-\title{The Importance of Being Nonplanar: Street Network Representation in Urban Form Studies}
+\title{Nonplanarity and Street Network Representation in Urban Form Studies}
 \author{Author Redacted \affilnum{1}}
@@
 \begin{abstract}
+
+Mathematical models of street networks, called graphs, are widely used throughout the urban studies literature to examine travel behavior, accessibility, urban design patterns, and morphology.
+These networks are commonly called planar graphs, meaning that they can be represented in two dimensions without any underpasses or overpasses.
+However, real urban street networks exist in three-dimensional space and frequently include grade separation such as bridges and tunnels.
+How well do planar graphs model real-world street networks?
+What does the extent of nonplanarity tell us about a place's infrastructure and urban development?
+We measure the nonplanarity of drivable and walkable street networks in the centers of 50 cities worldwide, then examine the variation of planarity across a single city.
+We find that while some street networks are approximately planar, many are poorly-modeled by planar graphs.
+In particular, planarity can drastically misrepresent routing, connectivity, intersection counts/densities, and block lengths.
 
@@
 In urban planning and design research, street networks are routinely used to calculate accessibility between origins and destinations or to compute indicators of the urban form, such as block sizes or intersection density and connectivity.
-Mathematical models of street networks, called graphs, have grown ubiquitous in the urban studies literature in recent years as they have been used to model household travel patterns, access and equity, pedestrian volume, urban design patterns and spatial morphology, and location centrality and polycentricity \citep{marshall_street_2010,porta_alterations_2014,marshall_community_2014,hajrasouliha_impact_2015,parthasarathi_street_2015,knight_metrics_2015,gil_street_2016,zhong_revealing_2017}.
+Mathematical models of street networks, called graphs, have grown ubiquitous in the urban studies literature in recent years as they have been used to model household travel patterns, access and equity, pedestrian volume, urban design patterns and spatial morphology, and location centrality and polycentricity \citep{southworth_street_1995,marshall_street_2010,porta_alterations_2014,marshall_community_2014,hajrasouliha_impact_2015,parthasarathi_street_2015,knight_metrics_2015,gil_street_2016,zhong_revealing_2017}.
 
@@
 
-Consider two theoretical examples: a medieval European city center and a modern American city center.
-The former's circulation network comprises a set of pedestrian paths and drivable streets without any grade separation, overpasses, or underpasses.
-In two-dimensions, its edges (i.e., streets and paths) cross each other only at nodes (i.e., intersections), so it is by definition planar.
-The latter's circulation network comprises pedestrian paths, drivable surface streets, and a grade-separated freeway with overpasses over some of the surface streets.
-In two-dimensions, its edges occasionally cross each other at non-nodes (i.e., overpasses), so it is by definition nonplanar.
-However, we might call it \emph{approximately} planar because its nonplanar edge crossings are relatively uncommon.
-Approximate planarity constrains nonplanar spatial networks such that they do not exhibit certain characteristics found among nonplanar aspatial graphs (e.g., small-world effects or power-law distributed node degrees).
+\begin{figure}[tbp]
+	\center
+	\includegraphics[width=0.48\textwidth]{planar_vs_not.png}
+	\caption{Formally and spatially planar (left) and nonplanar (right) road networks.}
+	\label{fig:planar_vs_not}
+\end{figure}
+
+Consider the examples in Figure \ref{fig:planar_vs_not}.
+The left network comprises a set of paths and streets without any bridges or tunnels.
+In two-dimensions, its edges intersect each other only at nodes so it is by definition planar.
+The right network comprises surface streets and a grade-separated freeway interchange with many overpasses and underpasses.
+In two-dimensions, its edges sometimes cross each other at non-nodes (i.e., overpasses), so it is spatially nonplanar.
+Furthermore it is impossible to redraw the graph by moving its nodes and edges around in space such that its edges only intersect at nodes.
+Therefore it is also formally nonplanar.
+However, at a city- or region-wide spatial scale, these nonplanar edge crossings may be relatively uncommon; we might call such a network \emph{approximately} planar.
+Approximate planarity constrains nonplanar spatial networks such that they do not exhibit certain characteristics found among nonplanar aspatial graphs, such as small-world effects or power-law distributed node degrees \citep{fischer_spatial_2014}.
 
@@
 
-The graph theory literature offers some measures of how \enquote{far off} a nonplanar graph $G$ is from being planar, including its \emph{crossing number} --- the minimum number of edge crossings of any drawing of $G$ --- and its \emph{skewness} --- the minimum number of edges that must be removed from $G$ to produce a planar graph \citep{liebers_planarizing_2001,szekely_successful_2004,chimani_non-planar_2009}.
-However, these measures are imperfect and hard to compute \citep{chimani_vertex_2012,szekely_successful_2004}.
-They also fail to correct for the size or density of a spatial network.
+The graph theory literature offers some measures of how \enquote{far off} a nonplanar graph $G$ is from being planar, including its \emph{crossing number} --- the minimum number of edge crossings of any drawing of $G$ --- and its \emph{skewness} --- the minimum number of edges that must be removed from $G$ to produce a planar graph \citep{liebers_planarizing_2001,chimani_non-planar_2009}.
+However, these measures are imperfect and hard to compute \citep{szekely_successful_2004,chimani_vertex_2012}.
+They also fail to correct for the size, density, or real-world embedding of a spatial network.
 In his discussion of road networks and approximate spatial planarity, \citet[p.~133]{newman_networks:_2010} argues that due to these drawbacks \enquote{no widely accepted metric for degree of planarity has emerged,} and calls for the development of better indicators.
@@
 Thus the walkable network may contain surface streets, paths through parks, pedestrian passageways between buildings or under roads, and other walkable paths.
-The drivable network may contain surface streets, grade-separated freeways, and other drivable paths.
+The drivable network may contain surface streets, grade-separated freeways, and other drivable routes.
 
@@
 This binary true/false indicator tells us if the graph is formally planar, ignoring its real-world spatial embedding.
-However, street networks \emph{are} spatially embedded, so we have developed the following two indicators to assess the \enquote{extent} to which they are planar.
+However, street networks \emph{are} spatially embedded.
+Accordingly, for this study, we have developed a second and third indicator to assess the \enquote{extent} to which they are planar.
 
-The second measure is the spatial planarity ratio (SPR).
-It represents the ratio of nonplanar intersections $i_n$ (i.e., non-dead-end nodes in the nonplanar, three-dimensional, spatially-embedded graph) to planar intersections $i_p$ (i.e., line crossings in the planar, two-dimensional, spatially-embedded graph):
+The second measure is the spatial planarity ratio $\phi$.
+It represents the ratio of nonplanar intersections $i_n$ (i.e., non-dead-end nodes in the nonplanar, three-dimensional, spatially-embedded graph) to planar intersections $i_p$ (i.e., edge crossings in the planar, two-dimensional, spatially-embedded graph):
 
@@
 	\label{eq:spr}
-	ELR = \frac{i_n}{i_p}
+	\phi = \frac{i_n}{i_p}
 \end{equation}
@@
 This indicates the extent to which planarity overstates intersections and connectivity in each street network.
-In other words, the SPR tells us what proportion of the line crossings in the planar graph are true intersections in the nonplanar graph.
-A truly planar network with no bridges or tunnels will thus have an SPR score of 1.0, while lower values indicate the percentage extent to which the network is planar.
-From the SPR, we can further calculate by what percentage the planar graph overstates intersection counts as:
+In other words, $\phi$ tells us what proportion of the two-dimensional edge crossings in the planar graph are true intersections in the nonplanar graph.
+A truly planar network with no bridges or tunnels will thus have an $\phi$ score of 1.0, while lower values indicate the percentage extent to which the network is planar.
+From the $\phi$, we can further calculate by what percentage the planar graph overstates intersection counts as:
 
@@
 	\label{eq:spr_overstates}
-	1 - \frac{1}{SPR}
+	1 - \frac{1}{\phi}
 \end{equation}
 
-The third measure is the edge length ratio (ELR). It represents the ratio of the mean edge length in the planar graph $l_p$ to the mean edge length in the nonplanar graph $l_n$:
+The third measure is the edge length ratio ($\lambda$). It represents the ratio of the mean edge length in the planar graph $l_p$ to the mean edge length in the nonplanar graph $l_n$:
 
@@
 	\label{eq:elr}
-	ELR = \frac{l_p}{l_n}
+	\lambda = \frac{l_p}{l_n}
 \end{equation}
@@
 This indicates the extent to which planarity understates edge lengths in each street network by fragmenting street segments at overpasses or underpasses.
-A truly planar street network with no bridges or tunnels will thus have an ELR score of 1.0, while lower values indicate the percentage extent to which the network is planar.
-From the ELR, we can further calculate by what percentage the planar graph understates the average edge length as:
+A truly planar street network with no bridges or tunnels will thus have an $\lambda$ score of 1.0, while lower values indicate the percentage extent to which the network is planar.
+From the $\lambda$, we can further calculate by what percentage the planar graph understates the average edge length as:
 
@@
 	\label{eq:elr_understates}
-	1 - ELR
+	1 - \lambda
 \end{equation}
@@
 	\centering
-	\caption{Planarity measures for the central street networks of 50 cities worldwide (Planar = whether street network passed the formal test of planarity; SPR = spatial planarity ratio; ELR = edge length ratio).}
+	\caption{Planarity measures for the central street networks of 50 cities worldwide (Planar = whether street network passed the formal test of planarity; $\phi$ = spatial planarity ratio; $\lambda$ = edge length ratio).}
 	\label{tab:world_cities}
@@
 	\includegraphics[width=\textwidth]{world_map_phi_bw.png}
-	\caption{Map of world cities from Table \ref{tab:world_cities} grouped by SPR score terciles (lower scores mean less planar).}
+	\caption{Map of world cities from Table \ref{tab:world_cities} grouped by $\phi$ terciles (lower values mean less planar).}
 	\label{fig:world_map_bw}
@@
 	\includegraphics[width=0.48\textwidth]{regression_phi_split.png}
-	\caption{Log-log plot of ELR vs SPR, by network type, with simple regression lines.}
+	\caption{Log-log plot of $\lambda$ vs $\phi$, by network type, with simple regression lines.}
 	\label{fig:regression_split}
@@
 
-Table \ref{tab:world_cities} shows the SPR, ELR, and results of the formal planarity test for the walkable and the drivable networks for each of the 50 cities.
+Table \ref{tab:world_cities} shows $\phi$, $\lambda$, and results of the formal planarity test for the walkable and the drivable networks in these 50 city centers.
 Among the drivable street networks, only 20\% are formally planar.
-On average, they are 88.5\% planar by the SPR measure and 90\% planar by the ELR measure.
-The individual SPR values indicate that spatial planarity ranges from a high of 100\% in seven of these cities to a low of 57\% in Moscow.
-The ELR values indicate that spatial planarity ranges from a high of 100\% in seven of these cities to a low of 63.5\% in Los Angeles.
+On average, they are 88.5\% planar by the $\phi$ measure and 90\% planar by the $\lambda$ measure.
+The individual $\phi$ values indicate that spatial planarity ranges from a high of 100\% in seven of these cities to a low of 57\% in Moscow.
+The $\lambda$ values indicate that spatial planarity ranges from a high of 100\% in seven of these cities to a low of 63.5\% in Los Angeles.
+On average across these networks, planar representations overcount intersections by 16\% and underestimate street segment lengths by 10\%.
 
 Among walkable street networks, only 10\% are formally planar.
-On average, they are 92\% planar by the SPR measure and 91.5\% planar by the ELR measure.
-The individual SPR values indicate that spatial planarity ranges from a high of 100\% in two cities to a low of 67\% in Shanghai.
-The ELR values indicate that spatial planarity ranges from a high of 100\% in two cities to a low of 66\% in Shanghai.
+On average, they are 92\% planar by the $\phi$ measure and 91.5\% planar by the $\lambda$ measure.
+The individual $\phi$ values indicate that spatial planarity ranges from a high of 100\% in two cities to a low of 67\% in Shanghai.
+The $\lambda$ values indicate that spatial planarity ranges from a high of 100\% in two cities to a low of 66\% in Shanghai.
+On average across these networks, planar representations overcount intersections by 9\% and underestimate street segment lengths by 8.5\%.
 Fewer walkable than drivable networks are formally planar, but on average these walking networks are slightly more spatially planar than the driving networks.
@@
 Not all formally planar street networks are spatially planar.
-For example, Toronto's drivable network is formally planar but only 93\% and 96\% spatially planar according to its SPR and ELR measures.
+For example, Toronto's drivable network is formally planar but only 93\% and 96\% spatially planar according to its $\phi$ and $\lambda$ measures.
 In total, three drivable networks (Toronto, Jakarta, and Copenhagen) and three walkable networks (Dallas, Delhi, and Bologna) are formally planar but spatially nonplanar to various extents.
-About a third of the cities studied demonstrate spatial planarity of 97\% or higher according to the SPR, suggesting that they are \enquote{approximately} planar.
+About a third of the cities studied demonstrate spatial planarity of 97\% or higher according to the $\phi$, suggesting that they are \enquote{approximately} planar.
 However, another third of the cities studied are less than 87\% planar.
-Dallas, Los Angeles, and Moscow have SPR values below 60\%, suggesting planar graphs poorly model these city centers.
+Dallas, Los Angeles, and Moscow have $\phi$ values below 60\%, suggesting planar graphs poorly model these city centers.
 Moreover, planar graphs overstate the intersection counts in these three networks by 67\%, 72\%, and 74\% respectively (Equation \ref{eq:spr_overstates}).
@@
 All three Italian cities demonstrate perfect planarity in their centers' drivable networks, but not in their walkable networks.
-The extent of planarity is not consistent across network types: Dallas's walkable SPR is 61\% greater than its drivable SPR, while Geneva's drivable SPR is 19\% greater than its walkable SPR.
-Figure \ref{fig:world_map_bw} shows the distribution of SPR values around the world.
+The extent of planarity is not consistent across network types: Dallas's walkable $\phi$ is 61\% greater than its drivable $\phi$, while Geneva's drivable $\phi$ is 19\% greater than its walkable $\phi$.
+Figure \ref{fig:world_map_bw} shows the distribution of $\phi$ values around the world.
 While nearly every European city is in the highest tercile, indicating their networks are more planar, most American cities are in the lowest tercile, indicating their networks are more nonplanar.
 
-Figure \ref{fig:regression_split} shows the relationship between the SPR and ELR indicators across all 50 cities for both network types, as a log-log plot.
+Figure \ref{fig:regression_split} shows the relationship between the $\phi$ and $\lambda$ indicators across all 50 cities for both network types, as a log-log plot.
 The indicators' log-log relationship is linear, positive, and very strong (drivable $r^2=0.98$ and walkable $r^2=0.99$).
@@
 We find that Oakland's city-wide street network is formally nonplanar.
-The city has an SPR score of 91.8\% and an ELR score of 93.6\%.
+The city has a $\phi$ score of 91.8\% and an $\lambda$ score of 93.6\%.
 This suggests that the planar representation of Oakland's drivable street network overstates the number of intersections --- and thus, the network's connectivity --- by 8.9\% city-wide (Equation \ref{eq:spr_overstates}) and understates the average edge length by 6.4\% city-wide (Equation \ref{eq:elr_understates}).
@@
 To explore this statistical variation, Table \ref{tab:samples_city} presents summary statistics of these planarity indicators across 100 square-mile samples of Oakland's drivable street network.
-Our samples' mean SPR and ELR scores are reasonably close to the city-wide values.
-However, the samples range from spatial planarity lows of 56.9\% (SPR) and 63.7\% (ELR) up to highs of 100\%.
+Our samples' mean $\phi$ and $\lambda$ scores are reasonably close to the city-wide values.
+However, the samples range from spatial planarity lows of 56.9\% ($\phi$) and 63.7\% ($\lambda$) up to highs of 100\%.
 67\% of the samples pass the formal planarity test.
-However, 63\% of the samples are at least somewhat spatially nonplanar (i.e., with SPR less than 1).
+However, 63\% of the samples are at least somewhat spatially nonplanar (i.e., with $\phi$ less than 1).
 
@@
 Street networks are frequently spatially nonplanar because they are embedded in three dimensions, not two: they have a z-coordinate (elevation) along with their x- and y-coordinates.
-But because they are usually \enquote{mostly} planar (average drivable SPR of 88.5\% and average walkable SPR of 92\%), typically with only a few overpasses or underpasses, they could often be described as \emph{approximately planar}.
+But because they are usually \enquote{mostly} planar (average drivable $\phi$ of 88.5\% and average walkable $\phi$ of 92\%), typically with only a few overpasses or underpasses, they could often be described as \emph{approximately planar}.
 However, claiming that urban street networks universally are \enquote{planar} misrepresents them in several ways:
@@
 
-As George Box famously stated, \enquote{All models are wrong but some are useful.} Even if they are not formally planar, can street networks be simplified to planar graphs and still be usefully well-modeled?
+As George Box once said, \enquote{All models are wrong but some are useful.} Even if they are not formally planar, can street networks be simplified to planar graphs and still be usefully well-modeled?
 Our results suggest that the answer depends on the study site and on the type of analysis.
@@
 We might repurpose this discrepancy between planar and nonplanar representations of a street network to reveal something useful about urban form.
-The SPR and ELR scores of the major US city centers indicate the greater three-dimensionality of their transportation infrastructure compared to that of the European city centers.
+The $\phi$ and $\lambda$ scores of the major US city centers indicate the greater three-dimensionality of their transportation infrastructure compared to that of the European city centers.
 These US city centers feature more automobile-oriented roadways and are often bounded by grade-separated freeways.
 By contrast, these European city centers feature less three-dimensionality as their streets and paths tend to be at-grade.
-Examining how SPR and ELR change over time in different cities could tell us something about the type and pace of urban development.
+Examining how $\phi$ and $\lambda$ change over time in different cities could tell us something about the type and pace of urban development.
 
-It is also interesting that most street networks are formally nonplanar while demonstrating SPR and ELR scores that are only slightly nonplanar.
+It is also interesting that most street networks are formally nonplanar while demonstrating $\phi$ and $\lambda$ scores that are only slightly nonplanar.
 Approximate planarity is common across cities because of costs, technology, and politics.
@@
 %\clearpage
-\bibliographystyle{apa}
+\bibliographystyle{apalike}
 \bibliography{references}

### 2018-01-24 manuscript.tex
edits

@@
 Mathematical models of street networks, called graphs, are widely used throughout the urban studies literature to examine travel behavior, accessibility, urban design patterns, and morphology.
-These networks are commonly called planar graphs, meaning that they can be represented in two dimensions without any underpasses or overpasses.
+These graphs are commonly described as \enquote{planar,} meaning that they can be represented in two dimensions without any underpasses or overpasses.
 However, real urban street networks exist in three-dimensional space and frequently include grade separation such as bridges and tunnels.
 How well do planar graphs model real-world street networks?
-What does the extent of nonplanarity tell us about a place's infrastructure and urban development?
+What does the extent of nonplanarity tell us about a city's infrastructure and urban development?
 We measure the nonplanarity of drivable and walkable street networks in the centers of 50 cities worldwide, then examine the variation of planarity across a single city.
 We find that while some street networks are approximately planar, many are poorly-modeled by planar graphs.
-In particular, planarity can drastically misrepresent routing, connectivity, intersection counts/densities, and block lengths.
+In particular, planarity can inconsistently but drastically misrepresent routing, connectivity, intersection counts/densities, and block lengths.
+We develop two measures of a network's nonplanarity and argue it can be a useful indicator of transportation infrastructure and development.
 
@@
 In urban planning and design research, street networks are routinely used to calculate accessibility between origins and destinations or to compute indicators of the urban form, such as block sizes or intersection density and connectivity.
-Mathematical models of street networks, called graphs, have grown ubiquitous in the urban studies literature in recent years as they have been used to model household travel patterns, access and equity, pedestrian volume, urban design patterns and spatial morphology, and location centrality and polycentricity \citep{southworth_street_1995,marshall_street_2010,porta_alterations_2014,marshall_community_2014,hajrasouliha_impact_2015,parthasarathi_street_2015,knight_metrics_2015,gil_street_2016,zhong_revealing_2017}.
+Mathematical models of street networks, called graphs, have grown ubiquitous in the urban studies literature in recent years as they have been used to model household travel patterns, access and equity, pedestrian volume, urban design patterns and spatial morphology, and location centrality and polycentricity \citep{marshall_street_2010,porta_alterations_2014,marshall_community_2014,hajrasouliha_impact_2015,parthasarathi_street_2015,knight_metrics_2015,gil_street_2016,zhong_revealing_2017}.
 
@@
 However, at a city- or region-wide spatial scale, these nonplanar edge crossings may be relatively uncommon; we might call such a network \emph{approximately} planar.
-Approximate planarity constrains nonplanar spatial networks such that they do not exhibit certain characteristics found among nonplanar aspatial graphs, such as small-world effects or power-law distributed node degrees \citep{fischer_spatial_2014}.
+Approximate planarity constrains nonplanar spatial networks such that they do not exhibit certain characteristics found among nonplanar aspatial graphs, such as small-world effects or power-law distributed node degrees \citep{crucitti_centrality_2006,fischer_spatial_2014}.
 
@@
 OSMnx automatically simplifies each graph's topology to retain nodes only at intersections and dead-ends, while faithfully retaining each edge's true spatial geometry.
-This provides us with an accurate count of intersections and an accurate measure of edge lengths for comparison between the planar and nonplanar representations of our networks.
+This provides an accurate count of intersections and an accurate measure of edge lengths for comparison between the planar and nonplanar representations of our networks.
 
@@
 
-This indicates the extent to which planarity overstates intersections and connectivity in each street network.
-In other words, $\phi$ tells us what proportion of the two-dimensional edge crossings in the planar graph are true intersections in the nonplanar graph.
-A truly planar network with no bridges or tunnels will thus have an $\phi$ score of 1.0, while lower values indicate the percentage extent to which the network is planar.
-From the $\phi$, we can further calculate by what percentage the planar graph overstates intersection counts as:
+Thus, $\phi$ represents what proportion of the two-dimensional edge crossings in the planar graph are true intersections in the nonplanar graph.
+This indicates the extent to which planarity overstates intersections and connectivity in a street network.
+A truly planar network with no bridges or tunnels will have an $\phi$ score of 1.0, while lower values indicate the percentage extent to which the network is planar.
+From this indicator, we can further calculate by what percentage the planar graph overstates intersection counts as:
 
@@
 Finally, we explore how these three indicators vary across a single city.
-To do so we analyze the drivable street network of Oakland, California, a reasonably representative midsized American city with a variety of urban form types from gridded street patterns in its flatlands, to winding culs-de-sac in its hills, to freeways and dense blocks around its downtown.
+To do so we analyze the drivable street network of Oakland, California as a case study.
+Oakland is a reasonably representative midsized American city with a variety of urban form types from gridded street patterns in its flatlands, to winding culs-de-sac in its hills, to freeways and dense blocks around its downtown.
 First we analyze the entire city of Oakland.
 Then we recreate the aforementioned methodology by randomly sampling 100 points within the city limits and analyzing the square-mile street networks centered on each.
-The resulting statistical dispersion of planarity demonstrates the extent to which analyzing an entire city's neighborhoods as a single graph may obscure neighborhood-scale planarity characteristics.
+The resulting statistical dispersion of planarity demonstrates the extent to which analyzing an entire city's neighborhoods as a single graph may obscure neighborhood-scale infrastructure characteristics.
 
@@
 
-Table \ref{tab:world_cities} shows $\phi$, $\lambda$, and results of the formal planarity test for the walkable and the drivable networks in these 50 city centers.
+Table \ref{tab:world_cities} lists the $\phi$, $\lambda$, and formal planarity of the walkable and drivable networks in these 50 city centers.
 Among the drivable street networks, only 20\% are formally planar.
@@
 The $\lambda$ values indicate that spatial planarity ranges from a high of 100\% in seven of these cities to a low of 63.5\% in Los Angeles.
-On average across these networks, planar representations overcount intersections by 16\% and underestimate street segment lengths by 10\%.
+On average across these networks, planar representations overcount intersections by 16\% and underestimate street segment lengths by 10\% (Equations \ref{eq:spr_overstates} and \ref{eq:elr_understates}).
 
@@
 Not all formally planar street networks are spatially planar.
-For example, Toronto's drivable network is formally planar but only 93\% and 96\% spatially planar according to its $\phi$ and $\lambda$ measures.
+For example, Toronto's drivable network is formally planar but only 93\% ($\phi$) and 96\% ($\lambda$) spatially planar.
 In total, three drivable networks (Toronto, Jakarta, and Copenhagen) and three walkable networks (Dallas, Delhi, and Bologna) are formally planar but spatially nonplanar to various extents.
-About a third of the cities studied demonstrate spatial planarity of 97\% or higher according to the $\phi$, suggesting that they are \enquote{approximately} planar.
+About a third of the cities studied demonstrate $\phi$ spatial planarity of 97\% or higher, suggesting that they are \enquote{approximately} planar.
 However, another third of the cities studied are less than 87\% planar.
 Dallas, Los Angeles, and Moscow have $\phi$ values below 60\%, suggesting planar graphs poorly model these city centers.
-Moreover, planar graphs overstate the intersection counts in these three networks by 67\%, 72\%, and 74\% respectively (Equation \ref{eq:spr_overstates}).
+Moreover, planar graphs overstate the intersection counts in these three networks by 67\%, 72\%, and 74\% respectively.
 
@@
 The extent of planarity is not consistent across network types: Dallas's walkable $\phi$ is 61\% greater than its drivable $\phi$, while Geneva's drivable $\phi$ is 19\% greater than its walkable $\phi$.
-Figure \ref{fig:world_map_bw} shows the distribution of $\phi$ values around the world.
+Figure \ref{fig:world_map_bw} maps the distribution of $\phi$ values around the world.
 While nearly every European city is in the highest tercile, indicating their networks are more planar, most American cities are in the lowest tercile, indicating their networks are more nonplanar.
 
-Figure \ref{fig:regression_split} shows the relationship between the $\phi$ and $\lambda$ indicators across all 50 cities for both network types, as a log-log plot.
+Figure \ref{fig:regression_split} depicts the relationship between the $\phi$ and $\lambda$ indicators across all 50 cities for both network types, as a log-log plot.
 The indicators' log-log relationship is linear, positive, and very strong (drivable $r^2=0.98$ and walkable $r^2=0.99$).
-This strong correlation tells us that these two indicators unsurprisingly provide redundant statistical information about the extent to which a network is planar.
+The coefficients of determination tell us that these two indicators unsurprisingly provide redundant statistical information about the extent to which a network is planar.
 However, each assesses different implications of this bias for measuring the urban form.
@@
 The city has a $\phi$ score of 91.8\% and an $\lambda$ score of 93.6\%.
-This suggests that the planar representation of Oakland's drivable street network overstates the number of intersections --- and thus, the network's connectivity --- by 8.9\% city-wide (Equation \ref{eq:spr_overstates}) and understates the average edge length by 6.4\% city-wide (Equation \ref{eq:elr_understates}).
+This suggests that the planar representation of Oakland's drivable street network overstates the number of intersections --- and thus, the network's connectivity --- by 8.9\% city-wide and understates the average edge length by 6.4\% city-wide.
 
@@
 However, the samples range from spatial planarity lows of 56.9\% ($\phi$) and 63.7\% ($\lambda$) up to highs of 100\%.
-67\% of the samples pass the formal planarity test.
-However, 63\% of the samples are at least somewhat spatially nonplanar (i.e., with $\phi$ less than 1).
+67\% of the samples pass the formal planarity test; however, 63\% of the samples are at least somewhat spatially nonplanar (i.e., with $\phi < 1$).
 
@@
 Medieval European towns or informal settlements in the global south may contain fewer grade-separated roads --- and thus are more planar --- than 20th-century American or 21st-century Chinese metropolises.
-This is a result of the prevailing transportation technologies when the urban form was developed, as well as the local terrain, wealth, culture, and politics.
+This is a result of the prevailing transportation technologies when the urban form was developed, as well as local terrain, wealth, culture, and politics \citep{southworth_street_1995}.
 
@@
 Contrary to some of the statements in the urban studies literature, our results suggest that it cannot be universally claimed that urban street networks are planar graphs.
-But perhaps some of it comes down to vocabulary.
-If road is not synonymous with street, then a road network and a street network may not be synonymous.
-A road network, including freeways and boulevards, may frequently be nonplanar, but a street network, focusing on municipal streets lined by land parcels, may be at-grade and planar.
-But this ignores the fact that even residential streets sometimes include bridges and tunnels in hilly neighborhoods, and the fact that our analysis earlier showed that walkable circulation networks in city centers often include pedestrian tunnels and footbridges.
 A graph is not planar because its edges \emph{usually} intersect only at nodes: by definition it is planar because its edges \emph{exclusively} intersect at nodes.
-Thus, street networks are not universally planar graphs.
-
 However, as \citet{newman_networks:_2010} points out, debating the semantics of formal planarity may be missing the point -- more interesting is the \emph{extent} to which a network is planar.
@@
 
-As George Box once said, \enquote{All models are wrong but some are useful.} Even if they are not formally planar, can street networks be simplified to planar graphs and still be usefully well-modeled?
+As George Box famously said, \enquote{All models are wrong but some are useful.} Even if they are not formally planar, can street networks be simplified to planar graphs and still be usefully well-modeled?
 Our results suggest that the answer depends on the study site and on the type of analysis.
-In limited circumstances, where the circulation network and built form exhibit few (or ideally zero) underpasses, overpasses, or grade-separation, then perhaps yes.
+In limited circumstances --- where the circulation network exhibits few (or ideally zero) underpasses, overpasses, or grade-separation --- perhaps yes.
 
@@
 Again, only in limited circumstances, such as the drivable networks in the three Italian cities we analyzed.
-Elsewhere, the results in Table \ref{tab:world_cities} showed how common urban form measures such as intersection counts are overstated by planar models, while average street segment lengths are consequently understated.
-Moreover, this misrepresentation behaves inconsistently from place to place: Figure \ref{fig:world_map_bw} demonstrated how the magnitude of bias varies across cities and modes of urbanization.
-Although planar graphs offer computationally tractable models and allow for the polygonal spatial analysis of urban blocks, they often modeled these street networks poorly compared to nonplanar graphs.
+Elsewhere, the results in Table \ref{tab:world_cities} showed how common urban form measures such as intersection counts are overstated by planar models (16\% on average in the drivable networks), while average street segment lengths are consequently understated (10\% on average in the drivable networks).
+Moreover, this misrepresentation behaves inconsistently from place to place: Figure \ref{fig:world_map_bw} and Table \ref{tab:samples_city} demonstrated how the magnitude of bias varies across cities and modes of urbanization.
+Although planar graphs offer computationally tractable models and allow for the polygonal spatial analysis of urban blocks, they often model these street networks poorly when compared to nonplanar graphs.
 
@@
 By contrast, these European city centers feature less three-dimensionality as their streets and paths tend to be at-grade.
-Examining how $\phi$ and $\lambda$ change over time in different cities could tell us something about the type and pace of urban development.
+Examining how $\phi$ and $\lambda$ change over time in different cities could reveal the type and pace of urban development.
 
-It is also interesting that most street networks are formally nonplanar while demonstrating $\phi$ and $\lambda$ scores that are only slightly nonplanar.
-Approximate planarity is common across cities because of costs, technology, and politics.
-Why?
-It is expensive to engineer a three-dimensional network with z-axis extents (nearly) as extensive as the x- and y-axes.
-Furthermore, a city fully consumed by overpasses and tunnels would present degraded livability and walkability, making it politically infeasible.
+Most street networks are formally nonplanar while demonstrating $\phi$ and $\lambda$ values that are only slightly nonplanar.
+Approximate planarity is common across cities because of costs, technology, and politics: it is expensive to engineer a three-dimensional network with z-axis extents (nearly) as extensive as the x- and y-axes.
+Furthermore, a city fully consumed by overpasses and tunnels would degrade livability and walkability, making it politically infeasible.
 
@@
 This study demonstrated that, although in limited circumstances these models may be accurate, they behave inconsistently across different kinds of cities by misrepresenting connectivity, accessibility, routing, intersection counts and densities, and street segment lengths.
-In most circumstances, nonplanar graphs provide better models.
-It also demonstrated how these indicators can be used to characterize urbanization in different cities, in particular through the transportation infrastructure's three-dimensionality.
+In most circumstances, nonplanar graphs provide more faithful models.
+It also demonstrated how these indicators can characterize urbanization in different cities, particularly through transportation infrastructure's three-dimensionality.
 

### 2018-01-24 manuscript.tex
edits

@@
 
-\runninghead{Author Redacted}
+\runninghead{Geoff Boeing ROUGH DRAFT}
 \title{Nonplanarity and Street Network Representation in Urban Form Studies}
-\author{Author Redacted \affilnum{1}}
+\author{Geoff Boeing ROUGH DRAFT \affilnum{1}}
 \affiliation{\affilnum{1}Affiliation redacted}
@@
 intersections at grade cannot be distinguished from intersections with an overpass or underpass that do not cross at grade} \citep[p.~395]{fischer_gis_2004}.
-Twenty years ago, \citet[p.~18]{fohl_non-planar_1996} called for a nonplanar model to better represent truly nonplanar spatial networks.
+Twenty years ago, \citet[p.~18]{fohl_non-planar_1996} called for nonplanar models to better represent truly nonplanar spatial networks.
 
 If a planar graph models a street network poorly, it could do so in multiple ways.
-Forcing planarity on a nonplanar street network creates artificial nodes in the graph at bridges and tunnels, which breaks routing.
+Imposing planarity on a nonplanar street network creates artificial nodes in the graph at bridges and tunnels, which breaks routing.
 As \citet[p.~6]{kwan_review_1996} describe it, \enquote{the difficulty in accurately representing overpasses or underpasses may lead to problems when running various routing algorithms (e.g.
@@
 This indicates the extent to which planarity understates edge lengths in each street network by fragmenting street segments at overpasses or underpasses.
-A truly planar street network with no bridges or tunnels will thus have an $\lambda$ score of 1.0, while lower values indicate the percentage extent to which the network is planar.
+A truly planar street network with no bridges or tunnels will thus have a $\lambda$ score of 1.0, while lower values indicate the percentage extent to which the network is planar.
 From the $\lambda$, we can further calculate by what percentage the planar graph understates the average edge length as:
@@
 
-Finally, we explore how these three indicators vary across a single city.
+Finally, we explore how these three indicators vary throughout a single city.
 To do so we analyze the drivable street network of Oakland, California as a case study.
@@
 In total, three drivable networks (Toronto, Jakarta, and Copenhagen) and three walkable networks (Dallas, Delhi, and Bologna) are formally planar but spatially nonplanar to various extents.
-About a third of the cities studied demonstrate $\phi$ spatial planarity of 97\% or higher, suggesting that they are \enquote{approximately} planar.
-However, another third of the cities studied are less than 87\% planar.
+About a third of the city centers studied demonstrate $\phi$ spatial planarity of 97\% or higher, suggesting that they are \enquote{approximately} planar.
+However, another third of the city centers are less than 87\% planar.
 Dallas, Los Angeles, and Moscow have $\phi$ values below 60\%, suggesting planar graphs poorly model these city centers.
@@
 Walkable networks are more affected by pedestrian flyovers and subways, as in Florence.
-However, even networks of non-freeway, non-pedestrian-only surface streets could easily be nonplanar due to bridges or tunnels in hilly neighborhoods or over rivers.
+However, even networks of non-freeway, non-pedestrian-only surface streets could easily be nonplanar due to tunnels in hilly neighborhoods or bridges over rivers.
 
@@
 Our results suggest that the answer depends on the study site and on the type of analysis.
-In limited circumstances --- where the circulation network exhibits few (or ideally zero) underpasses, overpasses, or grade-separation --- perhaps yes.
+In limited circumstances --- where the circulation network exhibits few (or ideally zero) underpasses, overpasses, or grade-separation --- perhaps yes, especially if the study focuses on polygon spatial analysis.
 
-But universally, we cannot answer yes.
+But universally we cannot answer yes, especially for topological studies.
 Most egregiously, imposing planarity on a nonplanar street network forces false nodes at underpasses and overpasses, breaking routing and network-based accessibility modeling.

### 2018-01-29 manuscript.tex
edits

@@
 
+\usepackage{draftwatermark}
+\SetWatermarkText{DRAFT}
+\SetWatermarkScale{1}
+
 \begin{document}
 
-\runninghead{Geoff Boeing ROUGH DRAFT}
+\runninghead{Geoff Boeing}
 \title{Nonplanarity and Street Network Representation in Urban Form Studies}
-\author{Geoff Boeing ROUGH DRAFT \affilnum{1}}
+\author{Geoff Boeing \affilnum{1}}
 \affiliation{\affilnum{1}Affiliation redacted}
@@
 
-Mathematical models of street networks, called graphs, are widely used throughout the urban studies literature to examine travel behavior, accessibility, urban design patterns, and morphology.
-These graphs are commonly described as \enquote{planar,} meaning that they can be represented in two dimensions without any underpasses or overpasses.
-However, real urban street networks exist in three-dimensional space and frequently include grade separation such as bridges and tunnels.
-How well do planar graphs model real-world street networks?
-What does the extent of nonplanarity tell us about a city's infrastructure and urban development?
-We measure the nonplanarity of drivable and walkable street networks in the centers of 50 cities worldwide, then examine the variation of planarity across a single city.
-We find that while some street networks are approximately planar, many are poorly-modeled by planar graphs.
-In particular, planarity can inconsistently but drastically misrepresent routing, connectivity, intersection counts/densities, and block lengths.
-We develop two measures of a network's nonplanarity and argue it can be a useful indicator of transportation infrastructure and development.
+Models of street networks are widely used throughout the urban studies literature to explore travel behavior, accessibility, design patterns, and morphology.
+These models are commonly defined as \enquote{planar,} meaning that they can be represented in two dimensions without any underpasses or overpasses.
+However, real-world urban street networks exist in three-dimensional space and frequently feature grade separation such as bridges and tunnels.
+How well do planar models represent real-world street networks?
+What does a city's extent of nonplanarity reveal about its infrastructure and urban development?
+This study measures the nonplanarity of drivable and walkable street networks in the centers of 50 cities worldwide, then examines the variation of nonplanarity across a single city.
+It develops new measures of a network's nonplanarity as indicators of transportation infrastructure and development.
+While some street networks are approximately planar, many are poorly-modeled by planar simplifications that inconsistently but drastically misrepresent routing, connectivity, intersection counts/densities, and block lengths.
 
@@
 In urban planning and design research, street networks are routinely used to calculate accessibility between origins and destinations or to compute indicators of the urban form, such as block sizes or intersection density and connectivity.
-Mathematical models of street networks, called graphs, have grown ubiquitous in the urban studies literature in recent years as they have been used to model household travel patterns, access and equity, pedestrian volume, urban design patterns and spatial morphology, and location centrality and polycentricity \citep{marshall_street_2010,porta_alterations_2014,marshall_community_2014,hajrasouliha_impact_2015,parthasarathi_street_2015,knight_metrics_2015,gil_street_2016,zhong_revealing_2017}.
+Mathematical models of street networks, called \emph{graphs}, have grown ubiquitous in the urban studies literature in recent years as they have been used to model household travel patterns, access and equity, pedestrian volume, urban design patterns, spatial morphology, and location centrality and polycentricity \citep{marshall_street_2010,porta_alterations_2014,marshall_community_2014,hajrasouliha_impact_2015,parthasarathi_street_2015,knight_metrics_2015,gil_street_2016,zhong_revealing_2017}.
 
-In the urban studies literature, street networks are typically referred to as \enquote{planar} or \enquote{approximately planar} (see Table \ref{tab:planar_quotes}), meaning that they can be well-modeled by a flat two-dimensional model that inherently precludes overpasses or underpasses.
+In the urban studies literature, street networks are often referred to as \enquote{planar} or \enquote{approximately planar}, meaning that they can be well-modeled by a flat two-dimensional model that inherently precludes overpasses or underpasses.
 Of course in the real world, urban street networks are embedded in three-dimensional space and often feature grade separation, bridges, and tunnels.
@@
 These indicators help describe the nature of the urban form and transportation infrastructure.
-Despite common claims in the literature that street networks are planar graphs, this study finds that they generally are nonplanar and that planar graphs poorly model the street networks of many cities.
+Despite common claims in the literature that urban street networks are planar graphs, this study finds that they generally are nonplanar and that planar graphs poorly model the street networks of many cities.
 Further, the magnitude of this bias varies substantially across cities and urbanization types.
 
-This article is organized as follows.
-The following section introduces the basics of graph theory relevant to urban studies, focusing on discussions in the research literature about street network planarity.
-The next section discusses the methods used to acquire and analyze the street networks in this study.
-Then we present the results of this analysis before concluding with a discussion of their ramifications for street network research and urban form studies.
+This article first introduces the basics of graph theory relevant to urban studies, focusing on discussions in the research literature about street network planarity.
+Next it discusses the methods used to acquire and analyze the street networks in this study.
+Then it present the results of this analysis before concluding with a discussion of ramifications for street network research and urban form studies.
 
@@
 Imposing planarity on a nonplanar street network creates artificial nodes in the graph at bridges and tunnels, which breaks routing.
-As \citet[p.~6]{kwan_review_1996} describe it, \enquote{the difficulty in accurately representing overpasses or underpasses may lead to problems when running various routing algorithms (e.g.
-recommending that a traveler make a left-turn at an intersection that proves to be an overpass)}.
-Intersection counts and densities will be overestimated in the presence of these false nodes.
-Consequently, edge lengths will be underestimated due to these artificial breakpoints splitting up street segments.
-Finally, this bias would likely behave inconsistently across different kinds of cities and street network types based on the extents to which they are planar.
+For example, this misrepresentation would allow a routing engine to recommend a left turn at an \enquote{intersection} that is actually an overpass \citep[p.~6]{kwan_review_1996}.
+Intersection counts and densities would be overestimated in the presence of these false nodes.
+Consequently, edge lengths would be underestimated due to these artificial breakpoints splitting up street segments.
+Finally, this bias might behave inconsistently across different kinds of cities and street network types based on the extents to which they are planar.
 
@@
 They also fail to correct for the size, density, or real-world embedding of a spatial network.
-In his discussion of road networks and approximate spatial planarity, \citet[p.~133]{newman_networks:_2010} argues that due to these drawbacks \enquote{no widely accepted metric for degree of planarity has emerged,} and calls for the development of better indicators.
+In his discussion of road networks and approximate spatial planarity, \citet[p.~133]{newman_networks:_2010} argues that due to these drawbacks no widely-accepted measure of the extent of nonplanarity has emerged and calls for the development of better indicators.
 
@@
 
-The second measure is the spatial planarity ratio $\phi$.
+The second measure is the Spatial Planarity Ratio $\phi$.
 It represents the ratio of nonplanar intersections $i_n$ (i.e., non-dead-end nodes in the nonplanar, three-dimensional, spatially-embedded graph) to planar intersections $i_p$ (i.e., edge crossings in the planar, two-dimensional, spatially-embedded graph):
@@
 
-The third measure is the edge length ratio ($\lambda$). It represents the ratio of the mean edge length in the planar graph $l_p$ to the mean edge length in the nonplanar graph $l_n$:
+The third measure is the Edge Length Ratio ($\lambda$). It represents the ratio of the mean edge length in the planar graph $l_p$ to the mean edge length in the nonplanar graph $l_n$:
 
@@
 	\centering
-	\caption{Planarity measures for the central street networks of 50 cities worldwide (Planar = whether street network passed the formal test of planarity; $\phi$ = spatial planarity ratio; $\lambda$ = edge length ratio).}
+	\caption{Planarity measures for the central street networks of 50 cities worldwide (Planar = whether street network passed the formal test of planarity; $\phi$ = Spatial Planarity Ratio; $\lambda$ = Edge Length Ratio).}
 	\label{tab:world_cities}
@@
 Among the drivable street networks, only 20\% are formally planar.
-On average, they are 88.5\% planar by the $\phi$ measure and 90\% planar by the $\lambda$ measure.
+On average, they are 88.5\% spatially planar by the $\phi$ measure and 90\% by the $\lambda$ measure.
 The individual $\phi$ values indicate that spatial planarity ranges from a high of 100\% in seven of these cities to a low of 57\% in Moscow.
@@
 Among walkable street networks, only 10\% are formally planar.
-On average, they are 92\% planar by the $\phi$ measure and 91.5\% planar by the $\lambda$ measure.
+On average, they are 92\% spatially planar by the $\phi$ measure and 91.5\% by the $\lambda$ measure.
 The individual $\phi$ values indicate that spatial planarity ranges from a high of 100\% in two cities to a low of 67\% in Shanghai.
@@
 The indicators' log-log relationship is linear, positive, and very strong (drivable $r^2=0.98$ and walkable $r^2=0.99$).
-The coefficients of determination tell us that these two indicators unsurprisingly provide redundant statistical information about the extent to which a network is planar.
+The coefficients of determination tell us that these two indicators unsurprisingly provide redundant statistical information about the extent to which a network is spatially planar.
 However, each assesses different implications of this bias for measuring the urban form.
@@
 We find that Oakland's city-wide street network is formally nonplanar.
-The city has a $\phi$ score of 91.8\% and an $\lambda$ score of 93.6\%.
+The city has a $\phi$ score of 91.8\% and a $\lambda$ score of 93.6\%.
 This suggests that the planar representation of Oakland's drivable street network overstates the number of intersections --- and thus, the network's connectivity --- by 8.9\% city-wide and understates the average edge length by 6.4\% city-wide.
@@
 
-Our results demonstrate how this is a bigger problem in Los Angeles than in Florence, but even in Florence the walking network is spatially and formally nonplanar due to the \textit{sottopassaggio} (pedestrian subway) near Stazione di Santa Maria Novella, its central train station.
-Our results suggest that spatial planarity is inconsistent both across cities as well as across different neighborhoods within individual cities.
+Our results demonstrate how this is a bigger problem in, say, Los Angeles than in Florence, but even in Florence the walking network is spatially and formally nonplanar due to the \textit{sottopassaggio} (pedestrian subway) near Stazione di Santa Maria Novella, its central train station.
+The results also suggest that spatial planarity is inconsistent both across cities as well as across different neighborhoods within individual cities.
 
 The problem with planar models is particularly pronounced around the downtowns of North American cities, due to the prevalence of freeways, bridges, and underpasses.
-Drivable networks are affected by these in particular.
-Walkable networks are more affected by pedestrian flyovers and subways, as in Florence.
-However, even networks of non-freeway, non-pedestrian-only surface streets could easily be nonplanar due to tunnels in hilly neighborhoods or bridges over rivers.
+Drivable networks are affected by these in particular, but even walkable networks are affected by pedestrian flyovers and subways, as in Florence.
+Furthermore, even networks of non-freeway, non-pedestrian-only surface streets could easily be nonplanar due to tunnels in hilly neighborhoods or bridges over rivers.
 
@@
 Approximate planarity is common across cities because of costs, technology, and politics: it is expensive to engineer a three-dimensional network with z-axis extents (nearly) as extensive as the x- and y-axes.
-Furthermore, a city fully consumed by overpasses and tunnels would degrade livability and walkability, making it politically infeasible.
+Furthermore, a city fully consumed by overpasses and tunnels would degrade quality of life, making it politically infeasible.
 
@@
 
-Urban researchers often use planar graphs to model street networks.
-This study demonstrated that, although in limited circumstances these models may be accurate, they behave inconsistently across different kinds of cities by misrepresenting connectivity, accessibility, routing, intersection counts and densities, and street segment lengths.
-In most circumstances, nonplanar graphs provide more faithful models.
-It also demonstrated how these indicators can characterize urbanization in different cities, particularly through transportation infrastructure's three-dimensionality.
+This study tested the formal planarity of central urban street networks and developed two new measures of their spatial planarity: the Spatial Planarity Ratio $\phi$ and the Edge Length Ratio $\lambda$.
+Although planar models may be accurate in limited circumstances, they behave inconsistently across different kinds of cities by misrepresenting connectivity, accessibility, routing, intersection counts and densities, and street segment lengths.
+Nonplanar graphs usually provide more faithful models.
+It also discussed how $\phi$ and $\lambda$ can characterize urbanization in different cities, particularly through the transportation infrastructure's three-dimensionality.
 

### 2018-02-01 manuscript.tex
revisions

@@
 
-\usepackage{draftwatermark}
-\SetWatermarkText{DRAFT}
-\SetWatermarkScale{1}
+%\usepackage{draftwatermark}
+%\SetWatermarkText{DRAFT}
+%\SetWatermarkScale{1}
 
@@
 \runninghead{Geoff Boeing}
-\title{Nonplanarity and Street Network Representation in Urban Form Studies}
+\title{Street Network Representation and Nonplanarity in Urban Form Studies}
 \author{Geoff Boeing \affilnum{1}}
@@
 
-Models of street networks are widely used throughout the urban studies literature to explore travel behavior, accessibility, design patterns, and morphology.
+Models of street networks underlie research in urban travel behavior, accessibility, design patterns, and morphology.
 These models are commonly defined as \enquote{planar,} meaning that they can be represented in two dimensions without any underpasses or overpasses.
@@
 This study measures the nonplanarity of drivable and walkable street networks in the centers of 50 cities worldwide, then examines the variation of nonplanarity across a single city.
-It develops new measures of a network's nonplanarity as indicators of transportation infrastructure and development.
+It develops new measures of a network's nonplanarity as indicators of infrastructure and urbanization.
 While some street networks are approximately planar, many are poorly-modeled by planar simplifications that inconsistently but drastically misrepresent routing, connectivity, intersection counts/densities, and block lengths.
@@
 
-In urban planning and design research, street networks are routinely used to calculate accessibility between origins and destinations or to compute indicators of the urban form, such as block sizes or intersection density and connectivity.
+In urban planning and transportation research, street networks are routinely used to calculate accessibility between origins and destinations or to compute indicators of the urban form, such as block sizes or intersection density and connectivity.
 Mathematical models of street networks, called \emph{graphs}, have grown ubiquitous in the urban studies literature in recent years as they have been used to model household travel patterns, access and equity, pedestrian volume, urban design patterns, spatial morphology, and location centrality and polycentricity \citep{marshall_street_2010,porta_alterations_2014,marshall_community_2014,hajrasouliha_impact_2015,parthasarathi_street_2015,knight_metrics_2015,gil_street_2016,zhong_revealing_2017}.
 
-In the urban studies literature, street networks are often referred to as \enquote{planar} or \enquote{approximately planar}, meaning that they can be well-modeled by a flat two-dimensional model that inherently precludes overpasses or underpasses.
-Of course in the real world, urban street networks are embedded in three-dimensional space and often feature grade separation, bridges, and tunnels.
+In the urban studies literature, street networks are often referred to as \enquote{planar} or \enquote{approximately planar,} meaning that they can be well-represented by a two-dimensional model that inherently precludes overpasses or underpasses.
+However, this claim has not been well studied or defined empirically.
+In the real world, urban street networks are embedded in three-dimensional space and often feature grade separation, bridges, and tunnels.
 This leads to two intertwined questions.
@@
 This study tests the planarity of street networks around the world.
-It also presents two new measures of the degree of nonplanarity that can be generalized to other types of spatial networks.
+It also presents two new measures of the extent of nonplanarity that can be generalized to other types of spatial networks.
 These indicators help describe the nature of the urban form and transportation infrastructure.
-Despite common claims in the literature that urban street networks are planar graphs, this study finds that they generally are nonplanar and that planar graphs poorly model the street networks of many cities.
+Despite common claims in the literature that urban street networks are planar graphs, this study finds that they generally are nonplanar and that planar graphs poorly model the street networks of many urban centers.
 Further, the magnitude of this bias varies substantially across cities and urbanization types.
 
-This article first introduces the basics of graph theory relevant to urban studies, focusing on discussions in the research literature about street network planarity.
+This paper first introduces the basics of graph theory relevant to urban studies, focusing on discussions in the research literature about street network planarity.
 Next it discusses the methods used to acquire and analyze the street networks in this study.
@@
 If the graph cannot be drawn --- or redrawn --- to meet this criterion, then the graph is nonplanar \citep{trudeau_introduction_1994}.
-Street networks are embedded in space, which provides them with geometry --- such as geographical coordinates, lengths, areas, and angles --- along with their topology.
+Street networks are embedded in space, which provides them with geometry --- such as geographical coordinates, lengths, areas, shapes, and angles --- along with their topology.
 
@@
 \centering
-\caption{Recent statements in the research literature regarding the representation of street networks as planar graphs.}
+\caption{Survey of recent statements in the research literature regarding the representation of street networks as planar graphs.}
 \label{tab:planar_quotes}
@@
 Therefore it is also formally nonplanar.
-However, at a city- or region-wide spatial scale, these nonplanar edge crossings may be relatively uncommon; we might call such a network \emph{approximately} planar.
+At a city- or region-wide spatial scale, these nonplanar edge crossings may be relatively uncommon: we might call such a network \emph{approximately} planar.
 Approximate planarity constrains nonplanar spatial networks such that they do not exhibit certain characteristics found among nonplanar aspatial graphs, such as small-world effects or power-law distributed node degrees \citep{crucitti_centrality_2006,fischer_spatial_2014}.
@@
 Planar graphs offer computational simplicity and tractability.
-They enable easy polygonal spatial analysis of city blocks and form \citep{fohl_non-planar_1996,barthelemy_paths_2017} as well as the Euler characteristic.
-In mathematics, there is a bijection between planar graphs and trees, and classifying planar graphs presents a trivial problem \citep{louf_typology_2014}.
+They enable simple polygonal analysis of city blocks and form \citep{fohl_non-planar_1996,barthelemy_paths_2017} as well as the $\alpha$ and $\gamma$ indices popular in road network analysis \citep{eppstein_studying_2008}.
+In mathematics, a bijection exists planar graphs and trees, and classifying planar graphs presents a trivial problem \citep{louf_typology_2014}.
 Planar graphs are easier to visualize and can be faster to run algorithms on \citep{liebers_planarizing_2001}.
@@
 
-Despite the computational and mathematical advantages of simple planar models, street networks are often nonplanar in reality: many include at least one overpass or underpass that results in the failure of formal proofs of their planarity, such as the \citet{kuratowski_sur_1930} theorem or the \cite{hopcroft_efficient_1974} algorithm \citep[cf.][]{gastner_spatial_2006}.
-As \citet[p.~7]{levinson_network_2012} points out, \enquote{Real networks are neither perfect, nor planar, nor grids, though they may approximate them.}
-
-Other authors have commented on this characteristic of street networks.
+Despite the computational advantages of simple planar models, street networks are often nonplanar in reality \citep[p.~7]{levinson_network_2012}: many include at least one overpass or underpass that results in the failure of formal proofs of their planarity, such as the \citet{kuratowski_sur_1930} theorem or the \cite{hopcroft_efficient_1974} algorithm \citep[cf.][]{gastner_spatial_2006}.
 \citet[p.~199]{jiang_object-oriented_2010} note that \enquote{quite often the transportation network has overpasses and underpasses that require a non-planar network representation.} \citet[p.~1258]{fischer_spatial_2014} explains that \enquote{for many infrastructure networks, {[planarity]} is approximately true, although bridges and tunnels in ground-transport networks are an obvious (but generally minor) exception.} However, the presence of such nonplanar elements can vex models.
@@
 intersections at grade cannot be distinguished from intersections with an overpass or underpass that do not cross at grade} \citep[p.~395]{fischer_gis_2004}.
-Twenty years ago, \citet[p.~18]{fohl_non-planar_1996} called for nonplanar models to better represent truly nonplanar spatial networks.
 
 If a planar graph models a street network poorly, it could do so in multiple ways.
-Imposing planarity on a nonplanar street network creates artificial nodes in the graph at bridges and tunnels, which breaks routing.
+Imposing planarity on a nonplanar street network produces artificial nodes in the graph at bridges and tunnels, which breaks routing.
 For example, this misrepresentation would allow a routing engine to recommend a left turn at an \enquote{intersection} that is actually an overpass \citep[p.~6]{kwan_review_1996}.
@@
 
-Given these issues, what do \emph{approximately planar} and \emph{well-modeled} mean for street network research?
+Given these issues, there are some unanswered questions in the theoretical and empirical literature.
+What do \emph{approximately planar} and \emph{well-modeled} mean for street network research?
 How close is \enquote{close enough} for a planar graph to competently model a formally nonplanar street network?
-Do the biases of planar models behave consistently across geographies and development eras or do they misrepresent different cities to different degrees?
+Do the biases of planar models behave consistently across geographies and development eras or do they misrepresent different cities to different extents?
 And if street networks are at least generally approximately planar, how can we measure just how planar or nonplanar a given street network is?
@@
 However, these measures are imperfect and hard to compute \citep{szekely_successful_2004,chimani_vertex_2012}.
-They also fail to correct for the size, density, or real-world embedding of a spatial network.
+They also fail to adjust for the size, density, or real-world embedding of a spatial network.
 In his discussion of road networks and approximate spatial planarity, \citet[p.~133]{newman_networks:_2010} argues that due to these drawbacks no widely-accepted measure of the extent of nonplanarity has emerged and calls for the development of better indicators.
@@
 
-In this study, we develop two new measures of the extent to which a spatial network is planar.
-We then analyze various world cities to better understand how well planar graphs model their street networks as well as the extent to which bias (i.e., model misrepresentation) varies across different places and types of urbanization.
-Finally, we consider what these indicators suggest about the nature of the urban form and transportation infrastructure in different cities.
+This study develops two new measures of the extent to which a spatial network is planar.
+Then it analyzes various city centers worldwide to better understand how well planar graphs model urban street networks and the extent to which bias (i.e., model misrepresentation) varies across places and types of urbanization.
+Finally, it considers what these indicators suggest about the urban form and transportation infrastructure in different cities.
 
@@
 OSMnx automatically simplifies each graph's topology to retain nodes only at intersections and dead-ends, while faithfully retaining each edge's true spatial geometry.
-This provides an accurate count of intersections and an accurate measure of edge lengths for comparison between the planar and nonplanar representations of our networks.
+This provides an accurate count of intersections and an accurate measure of edge lengths for comparison between the planar and nonplanar representations of these networks.
 
@@
 
-Once we have acquired and prepared our networks, we calculate three measures of planarity.
+Once we have acquired and prepared the networks, we calculate three measures of planarity.
 The first is an aspatial test of formal planarity using the algorithm described by \citet{boyer_subgraph_2012}.
@@
 This binary true/false indicator tells us if the graph is formally planar, ignoring its real-world spatial embedding.
-However, street networks \emph{are} spatially embedded.
+However, street networks \emph{are} spatially embedded: every spatially planar network is formally planar, but not every formally planar network is spatially planar.
 Accordingly, for this study, we have developed a second and third indicator to assess the \enquote{extent} to which they are planar.
@@
 The second measure is the Spatial Planarity Ratio $\phi$.
-It represents the ratio of nonplanar intersections $i_n$ (i.e., non-dead-end nodes in the nonplanar, three-dimensional, spatially-embedded graph) to planar intersections $i_p$ (i.e., edge crossings in the planar, two-dimensional, spatially-embedded graph):
+It represents the ratio of nonplanar intersections $i_n$ (i.e., non-dead-end nodes in the nonplanar, three-dimensional, spatially-embedded graph) to planar intersections $i_p$ (i.e., edge crossings in the two-dimensional spatially-embedded graph):
 
@@
 This indicates the extent to which planarity overstates intersections and connectivity in a street network.
-A truly planar network with no bridges or tunnels will have an $\phi$ score of 1.0, while lower values indicate the percentage extent to which the network is planar.
+A spatially planar network with no bridges or tunnels will have an $\phi$ score of 1.0, while lower values indicate the percentage extent to which the network is planar.
 From this indicator, we can further calculate by what percentage the planar graph overstates intersection counts as:
@@
 This indicates the extent to which planarity understates edge lengths in each street network by fragmenting street segments at overpasses or underpasses.
-A truly planar street network with no bridges or tunnels will thus have a $\lambda$ score of 1.0, while lower values indicate the percentage extent to which the network is planar.
+A spatially planar street network will thus have a $\lambda$ score of 1.0, while lower values indicate the percentage extent to which the network is planar.
 From the $\lambda$, we can further calculate by what percentage the planar graph understates the average edge length as:
@@
 	\centering
-	\caption{Planarity measures for the central street networks of 50 cities worldwide (Planar = whether street network passed the formal test of planarity; $\phi$ = Spatial Planarity Ratio; $\lambda$ = Edge Length Ratio).}
+	\caption{Planarity measures for central street networks in 50 cities worldwide (Planar = whether street network passed the formal test of planarity; $\phi$ = Spatial Planarity Ratio; $\lambda$ = Edge Length Ratio).}
 	\label{tab:world_cities}
@@
 	\center
-	\includegraphics[width=\textwidth]{world_map_phi_bw.png}
+	\includegraphics[width=\textwidth]{world_map_bw.png}
 	\caption{Map of world cities from Table \ref{tab:world_cities} grouped by $\phi$ terciles (lower values mean less planar).}
@@
 	\includegraphics[width=0.48\textwidth]{regression_phi_split.png}
-	\caption{Log-log plot of $\lambda$ vs $\phi$, by network type, with simple regression lines.}
+	\caption{Log-log plot of $\lambda$ vs $\phi$ by network type with simple regression lines.}
 	\label{fig:regression_split}
@@
 
-Table \ref{tab:world_cities} lists the $\phi$, $\lambda$, and formal planarity of the walkable and drivable networks in these 50 city centers.
+Table \ref{tab:world_cities} lists the planarity indicators for these 50 city centers.
 Among the drivable street networks, only 20\% are formally planar.
@@
 	\centering
-	\caption{Summary statistics of planarity indicators across 100 random samples of Oakland, California's drivable network.}
+	\caption{Summary statistics of planarity indicators across 100 square-mile samples of Oakland, California's drivable network.}
 	\label{tab:samples_city}
@@
 
-Our findings suggest that the street networks at the centers of most major cities are formally nonplanar.
-However, this depends on the scale of measurement: across an entire city there is likely to be at least one overpass or underpass somewhere, while individual neighborhoods or small towns might be formally planar in their entirety.
+These findings suggest that the street networks at the centers of most major cities are formally and spatially nonplanar.
+However, this depends on the scale of measurement: across an entire city there is likely to be at least one overpass or underpass somewhere, while individual neighborhoods or small towns may be entirely planar.
 The type and era of urbanization represent another factor.
-Medieval European towns or informal settlements in the global south may contain fewer grade-separated roads --- and thus are more planar --- than 20th-century American or 21st-century Chinese metropolises.
-This is a result of the prevailing transportation technologies when the urban form was developed, as well as local terrain, wealth, culture, and politics \citep{southworth_street_1995}.
+Medieval European towns or informal settlements in the global south may contain fewer grade-separated roads --- and thus would be more planar --- than 20th-century American or 21st-century Chinese metropolises.
+This results from the prevailing transportation and engineering technologies when the urban form developed, as well as local terrain, wealth, culture, and politics \citep{southworth_street_1995}.
 
-Street networks are frequently spatially nonplanar because they are embedded in three dimensions, not two: they have a z-coordinate (elevation) along with their x- and y-coordinates.
-But because they are usually \enquote{mostly} planar (average drivable $\phi$ of 88.5\% and average walkable $\phi$ of 92\%), typically with only a few overpasses or underpasses, they could often be described as \emph{approximately planar}.
-However, claiming that urban street networks universally are \enquote{planar} misrepresents them in several ways:
+Street networks are frequently spatially nonplanar because they are embedded in three dimensions, not two: they have a (less-extensive) z-axis along with their x- and y-axes.
+But because they are usually \enquote{mostly} planar (average drivable $\phi$ of 88.5\% and average walkable $\phi$ of 92\%), typically with only a few overpasses or underpasses, they could often be described as \emph{approximately} planar.
+However, claiming that urban street networks universally are \enquote{planar,} or modeling them as such, misrepresents them in several ways:
 
@@
 
-Our results demonstrate how this is a bigger problem in, say, Los Angeles than in Florence, but even in Florence the walking network is spatially and formally nonplanar due to the \textit{sottopassaggio} (pedestrian subway) near Stazione di Santa Maria Novella, its central train station.
+Our results demonstrate how this is a bigger problem in Los Angeles than in Florence.
+But even in Florence the walking network is both spatially and formally nonplanar due to the \textit{sottopassaggio} (pedestrian subway) near Stazione di Santa Maria Novella, its central train station.
 The results also suggest that spatial planarity is inconsistent both across cities as well as across different neighborhoods within individual cities.
@@
 
-Contrary to some of the statements in the urban studies literature, our results suggest that it cannot be universally claimed that urban street networks are planar graphs.
 A graph is not planar because its edges \emph{usually} intersect only at nodes: by definition it is planar because its edges \emph{exclusively} intersect at nodes.
-However, as \citet{newman_networks:_2010} points out, debating the semantics of formal planarity may be missing the point -- more interesting is the \emph{extent} to which a network is planar.
+Contrary to some of the statements in the urban studies literature, our results suggest that it cannot be universally claimed that urban street networks are planar graphs --- formally or spatially.
+However, as \citet{newman_networks:_2010} points out, debating the semantics of formal planarity may be missing the point -- more interesting is the \emph{extent} to which a network is planar and its implications for modeling.
 
@@
 Our results suggest that the answer depends on the study site and on the type of analysis.
-In limited circumstances --- where the circulation network exhibits few (or ideally zero) underpasses, overpasses, or grade-separation --- perhaps yes, especially if the study focuses on polygon spatial analysis.
-
+In limited circumstances --- where the circulation network exhibits few (or ideally zero) underpasses, overpasses, or grade-separation --- then perhaps yes, especially if the study focuses on polygonal spatial analysis.
 But universally we cannot answer yes, especially for topological studies.
@@
 But planar graphs are often used in the literature to characterize urban form and morphology.
-So, aside from routing, do planar graphs offer \emph{useful} models for this type of research?
+Aside from routing, do planar graphs offer useful models for this type of research?
 Again, only in limited circumstances, such as the drivable networks in the three Italian cities we analyzed.
@@
 Moreover, this misrepresentation behaves inconsistently from place to place: Figure \ref{fig:world_map_bw} and Table \ref{tab:samples_city} demonstrated how the magnitude of bias varies across cities and modes of urbanization.
-Although planar graphs offer computationally tractable models and allow for the polygonal spatial analysis of urban blocks, they often model these street networks poorly when compared to nonplanar graphs.
+
+We might shoehorn real-world data into a chosen model and be able to perform a desired computation, but we lose the underlying ability to reason when the real world no longer obeys the rules and constraints of the chosen model.
+Although planar graphs offer computational tractability and allow for the polygonal analysis of urban blocks, they often model these street networks poorly when compared to nonplanar graphs.
 
@@
 Examining how $\phi$ and $\lambda$ change over time in different cities could reveal the type and pace of urban development.
+Future research can further explore how these indicators correlate with other measures of urbanization, development, and era.
 
@@
 
+Street networks are often defined and modeled as planar graphs in the urban studies literature despite typically being nonplanar in reality.
 This study tested the formal planarity of central urban street networks and developed two new measures of their spatial planarity: the Spatial Planarity Ratio $\phi$ and the Edge Length Ratio $\lambda$.
-Although planar models may be accurate in limited circumstances, they behave inconsistently across different kinds of cities by misrepresenting connectivity, accessibility, routing, intersection counts and densities, and street segment lengths.
-Nonplanar graphs usually provide more faithful models.
-It also discussed how $\phi$ and $\lambda$ can characterize urbanization in different cities, particularly through the transportation infrastructure's three-dimensionality.
-
-Future research can further explore this latter finding, as it likely correlates with other measures of urbanization, development, and era.
-Finally, future research might examine how nonplanar intersection counts represent true intersections if multiple adjacent edges form multiple graph intersections at a point where only one true intersection exists from an urban design perspective.
+Although planar models may be useful in limited circumstances, they behave inconsistently within and across cities by misrepresenting connectivity, accessibility, routing, intersection counts and densities, and street segment lengths.
+Nonplanar graphs usually provide more faithful models, and $\phi$ and $\lambda$ can characterize urbanization in different cities, particularly through the transportation infrastructure's three-dimensionality.
 

### 2018-02-01 manuscript.tex
revise intro conclusion

@@
 However, real-world urban street networks exist in three-dimensional space and frequently feature grade separation such as bridges and tunnels.
-How well do planar models represent real-world street networks?
-What does a city's extent of nonplanarity reveal about its infrastructure and urban development?
+Planar simplifications affect the results of real-world street network analysis.
+A city's extent of nonplanarity can describe its infrastructure and urban development.
 This study measures the nonplanarity of drivable and walkable street networks in the centers of 50 cities worldwide, then examines the variation of nonplanarity across a single city.
-It develops new measures of a network's nonplanarity as indicators of infrastructure and urbanization.
-While some street networks are approximately planar, many are poorly-modeled by planar simplifications that inconsistently but drastically misrepresent routing, connectivity, intersection counts/densities, and block lengths.
+It develops new nonplanarity indicators of infrastructure and urbanization.
+While some street networks are approximately planar, we empirically quantify how planar simplifications can inconsistently but drastically misrepresent routing, connectivity, intersection counts/densities, and block lengths.
 
@@
 
-In the urban studies literature, street networks are often referred to as \enquote{planar} or \enquote{approximately planar,} meaning that they can be well-represented by a two-dimensional model that inherently precludes overpasses or underpasses.
+The urban studies literature often defines street networks as \enquote{planar} or \enquote{approximately planar,} meaning they can be well-represented by a two-dimensional model that inherently precludes overpasses or underpasses.
 However, this claim has not been well studied or defined empirically.
 In the real world, urban street networks are embedded in three-dimensional space and often feature grade separation, bridges, and tunnels.
-This leads to two intertwined questions.
-First, how well do these planar graphs model urban street networks?
-Second, what does the extent to which a network is nonplanar tell us about a city's infrastructure and development?
+How well do planar graphs model urban street networks?
+To what extent do convenient planar simplifications alter network analysis results?
+What does the extent to which a network is nonplanar tell us about a city's infrastructure and development?
 
-This study tests the planarity of street networks around the world.
-It also presents two new measures of the extent of nonplanarity that can be generalized to other types of spatial networks.
-These indicators help describe the nature of the urban form and transportation infrastructure.
-Despite common claims in the literature that urban street networks are planar graphs, this study finds that they generally are nonplanar and that planar graphs poorly model the street networks of many urban centers.
+This study tests the planarity of street networks around the world and demonstrates how planar simplifications alter the results of street network analysis.
+It also presents two new measures of the \enquote{extent of nonplanarity} that describe the urban form and transportation infrastructure and can be generalized to other types of spatial networks.
+Despite common assumptions in the literature that urban street networks are planar graphs, this study finds that they generally are nonplanar and empirically quantifies how they misrepresent the street networks of many urban centers.
 Further, the magnitude of this bias varies substantially across cities and urbanization types.
@@
 Next it discusses the methods used to acquire and analyze the street networks in this study.
-Then it present the results of this analysis before concluding with a discussion of ramifications for street network research and urban form studies.
+Then it present the results of this analysis before concluding with a discussion of implications for street network research and urban form studies.
 
@@
 Planar graphs offer computational simplicity and tractability.
-They enable simple polygonal analysis of city blocks and form \citep{fohl_non-planar_1996,barthelemy_paths_2017} as well as the $\alpha$ and $\gamma$ indices popular in road network analysis \citep{eppstein_studying_2008}.
+They enable simple polygonal analysis of city blocks and form \citep{fohl_non-planar_1996,barthelemy_paths_2017} as well as the \emph{alpha} and \emph{gamma} indices popular in road network analysis \citep{eppstein_studying_2008}.
 In mathematics, a bijection exists planar graphs and trees, and classifying planar graphs presents a trivial problem \citep{louf_typology_2014}.
@@
 This study develops two new measures of the extent to which a spatial network is planar.
-Then it analyzes various city centers worldwide to better understand how well planar graphs model urban street networks and the extent to which bias (i.e., model misrepresentation) varies across places and types of urbanization.
+It empirically analyzes various city centers worldwide to quantify how planar models impact street network analysis results and the extent to which bias (i.e., model misrepresentation) varies across places and types of urbanization.
 Finally, it considers what these indicators suggest about the urban form and transportation infrastructure in different cities.
@@
 
-Table \ref{tab:world_cities} lists the planarity indicators for these 50 city centers.
+Table \ref{tab:world_cities} demonstrates how planar simplifications impact street network analysis results in these 50 city centers.
 Among the drivable street networks, only 20\% are formally planar.
@@
 The $\lambda$ values indicate that spatial planarity ranges from a high of 100\% in seven of these cities to a low of 63.5\% in Los Angeles.
-On average across these networks, planar representations overcount intersections by 16\% and underestimate street segment lengths by 10\% (Equations \ref{eq:spr_overstates} and \ref{eq:elr_understates}).
+On average across these networks, planar models overcount intersections by 16\% and underestimate street segment lengths by 10\% (Equations \ref{eq:spr_overstates} and \ref{eq:elr_understates}).
 
@@
 The $\lambda$ values indicate that spatial planarity ranges from a high of 100\% in two cities to a low of 66\% in Shanghai.
-On average across these networks, planar representations overcount intersections by 9\% and underestimate street segment lengths by 8.5\%.
+On average across these networks, planar models overcount intersections by 9\% and underestimate street segment lengths by 8.5\%.
 Fewer walkable than drivable networks are formally planar, but on average these walking networks are slightly more spatially planar than the driving networks.
@@
 However, the samples range from spatial planarity lows of 56.9\% ($\phi$) and 63.7\% ($\lambda$) up to highs of 100\%.
-67\% of the samples pass the formal planarity test; however, 63\% of the samples are at least somewhat spatially nonplanar (i.e., with $\phi < 1$).
+67\% of the samples pass the formal planarity test; however, 63\% of the samples are at least somewhat spatially nonplanar (i.e., $\phi < 1$).
 
@@
 
-Street networks are frequently spatially nonplanar because they are embedded in three dimensions, not two: they have a (less-extensive) z-axis along with their x- and y-axes.
+Street networks are frequently spatially nonplanar because they are embedded in three dimensions, not two.
 But because they are usually \enquote{mostly} planar (average drivable $\phi$ of 88.5\% and average walkable $\phi$ of 92\%), typically with only a few overpasses or underpasses, they could often be described as \emph{approximately} planar.
@@
 A graph is not planar because its edges \emph{usually} intersect only at nodes: by definition it is planar because its edges \emph{exclusively} intersect at nodes.
-Contrary to some of the statements in the urban studies literature, our results suggest that it cannot be universally claimed that urban street networks are planar graphs --- formally or spatially.
-However, as \citet{newman_networks:_2010} points out, debating the semantics of formal planarity may be missing the point -- more interesting is the \emph{extent} to which a network is planar and its implications for modeling.
+These results suggest that it cannot be universally claimed that urban street networks are planar graphs --- formally or spatially.
+However, as \citet{newman_networks:_2010} points out, debating the semantics of formal planarity may be missing the point -- more interesting is the \emph{extent} to which a network is planar.
+How do planar simplifications impact modeling and analysis?
 
@@
 
-But planar graphs are often used in the literature to characterize urban form and morphology.
-Aside from routing, do planar graphs offer useful models for this type of research?
+But aside from routing, planar graphs are often used to characterize urban form and morphology.
+Do they provide useful models for this type of research?
 Again, only in limited circumstances, such as the drivable networks in the three Italian cities we analyzed.
-Elsewhere, the results in Table \ref{tab:world_cities} showed how common urban form measures such as intersection counts are overstated by planar models (16\% on average in the drivable networks), while average street segment lengths are consequently understated (10\% on average in the drivable networks).
+Elsewhere, Table \ref{tab:world_cities} showed how the results of common urban form analyses such as intersection counts are overstated by planar models (16\% on average in the drivable networks), while average street segment lengths are consequently understated (10\% on average in the drivable networks).
 Moreover, this misrepresentation behaves inconsistently from place to place: Figure \ref{fig:world_map_bw} and Table \ref{tab:samples_city} demonstrated how the magnitude of bias varies across cities and modes of urbanization.
 
-We might shoehorn real-world data into a chosen model and be able to perform a desired computation, but we lose the underlying ability to reason when the real world no longer obeys the rules and constraints of the chosen model.
-Although planar graphs offer computational tractability and allow for the polygonal analysis of urban blocks, they often model these street networks poorly when compared to nonplanar graphs.
+We may be able to shoehorn real-world data into a chosen model and perform a desired computation, but we lose the underlying ability to reason when the real world no longer obeys the rules and constraints of the chosen model.
+Although planar simplifications offer algorithmic tractability and allow for the polygonal analysis of urban blocks, these results demonstrate how they can model urban street networks poorly and bias analytical results.
 
@@
 
-We might repurpose this discrepancy between planar and nonplanar representations of a street network to reveal something useful about urban form.
+We might repurpose this discrepancy between planar and nonplanar models of a street network to reveal something useful about urban form.
 The $\phi$ and $\lambda$ scores of the major US city centers indicate the greater three-dimensionality of their transportation infrastructure compared to that of the European city centers.
@@
 Street networks are often defined and modeled as planar graphs in the urban studies literature despite typically being nonplanar in reality.
-This study tested the formal planarity of central urban street networks and developed two new measures of their spatial planarity: the Spatial Planarity Ratio $\phi$ and the Edge Length Ratio $\lambda$.
+This study tested the formal planarity of central urban street networks and  empirically demonstrated how planar simplifications affect analysis results.
+It developed two new measures of the extent to which they are nonplanar --- the Spatial Planarity Ratio $\phi$ and the Edge Length Ratio $\lambda$ ---  to characterize urbanization particularly through the transportation infrastructure's three-dimensionality.
 Although planar models may be useful in limited circumstances, they behave inconsistently within and across cities by misrepresenting connectivity, accessibility, routing, intersection counts and densities, and street segment lengths.
-Nonplanar graphs usually provide more faithful models, and $\phi$ and $\lambda$ can characterize urbanization in different cities, particularly through the transportation infrastructure's three-dimensionality.
 

### 2018-02-02 manuscript.tex
edits

@@
 This study measures the nonplanarity of drivable and walkable street networks in the centers of 50 cities worldwide, then examines the variation of nonplanarity across a single city.
-It develops new nonplanarity indicators of infrastructure and urbanization.
+It develops two new nonplanarity indicators of infrastructure and urbanization: the Spatial Planarity Ratio and the Edge Length Ratio.
 While some street networks are approximately planar, we empirically quantify how planar simplifications can inconsistently but drastically misrepresent routing, connectivity, intersection counts/densities, and block lengths.
@@
 
-Despite the computational advantages of simple planar models, street networks are often nonplanar in reality \citep[p.~7]{levinson_network_2012}: many include at least one overpass or underpass that results in the failure of formal proofs of their planarity, such as the \citet{kuratowski_sur_1930} theorem or the \cite{hopcroft_efficient_1974} algorithm \citep[cf.][]{gastner_spatial_2006}.
+Despite the computational advantages of planar simplifications, street networks obviously are often nonplanar in reality \citep[p.~7]{levinson_network_2012}: many include at least one overpass or underpass that results in the failure of formal proofs of their planarity, such as the \citet{kuratowski_sur_1930} theorem or the \cite{hopcroft_efficient_1974} algorithm \citep[cf.][]{gastner_spatial_2006}.
 \citet[p.~199]{jiang_object-oriented_2010} note that \enquote{quite often the transportation network has overpasses and underpasses that require a non-planar network representation.} \citet[p.~1258]{fischer_spatial_2014} explains that \enquote{for many infrastructure networks, {[planarity]} is approximately true, although bridges and tunnels in ground-transport networks are an obvious (but generally minor) exception.} However, the presence of such nonplanar elements can vex models.
@@
 To acquire these street networks, we use OSMnx to download the data for each city and network type from OpenStreetMap.
-OpenStreetMap is a collaborative online mapping platform commonly used by researchers because of its good worldwide coverage \citep{haklay_how_2010,jokar_arsanjani_openstreetmap_2015}.
+OpenStreetMap is a collaborative online mapping platform commonly used by researchers because of its high-quality worldwide coverage \citep{haklay_how_2010,jokar_arsanjani_openstreetmap_2015}.
 OSMnx is a Python-based software tool that allows us to automatically download a street network from OpenStreetMap for any study site in the world, and automatically process it into a length-weighted nonplanar directed graph \citep{boeing_osmnx:_2017}.
@@
 
-The second measure is the Spatial Planarity Ratio $\phi$.
+The second measure is the Spatial Planarity Ratio, $\phi$.
 It represents the ratio of nonplanar intersections $i_n$ (i.e., non-dead-end nodes in the nonplanar, three-dimensional, spatially-embedded graph) to planar intersections $i_p$ (i.e., edge crossings in the two-dimensional spatially-embedded graph):
@@
 Thus, $\phi$ represents what proportion of the two-dimensional edge crossings in the planar graph are true intersections in the nonplanar graph.
-This indicates the extent to which planarity overstates intersections and connectivity in a street network.
-A spatially planar network with no bridges or tunnels will have an $\phi$ score of 1.0, while lower values indicate the percentage extent to which the network is planar.
-From this indicator, we can further calculate by what percentage the planar graph overstates intersection counts as:
+A spatially planar network with no bridges or tunnels will have a $\phi$ score of 1.0, while lower values indicate the percentage extent to which the network is planar.
+From this indicator, we can calculate how planar simplifications overstate intersection counts and connectivity in a street network (i.e., by what percentage the planar model overcounts intersections) as:
 
@@
 
-The third measure is the Edge Length Ratio ($\lambda$). It represents the ratio of the mean edge length in the planar graph $l_p$ to the mean edge length in the nonplanar graph $l_n$:
+The third measure is the Edge Length Ratio, $\lambda$. It represents the ratio of the planar graph's mean edge length $l_p$ to the nonplanar graph's mean edge length $l_n$:
 
@@
 
-This indicates the extent to which planarity understates edge lengths in each street network by fragmenting street segments at overpasses or underpasses.
+This measures how planar simplifications underestimate edge lengths in each street network by fragmenting street segments at overpasses or underpasses.
 A spatially planar street network will thus have a $\lambda$ score of 1.0, while lower values indicate the percentage extent to which the network is planar.
-From the $\lambda$, we can further calculate by what percentage the planar graph understates the average edge length as:
+We can calculate by what percentage the planar graph understates the average edge length as:
 
@@
 First we analyze the entire city of Oakland.
-Then we recreate the aforementioned methodology by randomly sampling 100 points within the city limits and analyzing the square-mile street networks centered on each.
+Then we randomly sample 100 points within the city limits and analyze the square-mile street networks centered on each.
 The resulting statistical dispersion of planarity demonstrates the extent to which analyzing an entire city's neighborhoods as a single graph may obscure neighborhood-scale infrastructure characteristics.
@@
 Dallas, Los Angeles, and Moscow have $\phi$ values below 60\%, suggesting planar graphs poorly model these city centers.
-Moreover, planar graphs overstate the intersection counts in these three networks by 67\%, 72\%, and 74\% respectively.
+In fact, planarity overstates the intersection counts in these three networks by 67\%, 72\%, and 74\% respectively.
 
@@
 All three Italian cities demonstrate perfect planarity in their centers' drivable networks, but not in their walkable networks.
-The extent of planarity is not consistent across network types: Dallas's walkable $\phi$ is 61\% greater than its drivable $\phi$, while Geneva's drivable $\phi$ is 19\% greater than its walkable $\phi$.
+The extent of nonplanarity is not consistent across network types: Dallas's walkable $\phi$ is 61\% greater than its drivable $\phi$, while Geneva's drivable $\phi$ is 19\% greater than its walkable $\phi$.
 Figure \ref{fig:world_map_bw} maps the distribution of $\phi$ values around the world.
@@
 Finally, we examine how these measures behave across an entire city.
-We find that Oakland's city-wide street network is formally nonplanar.
-The city has a $\phi$ score of 91.8\% and a $\lambda$ score of 93.6\%.
+Oakland's city-wide street network is formally nonplanar.
+In terms of spatial planarity, the city has a $\phi$ score of 91.8\% and a $\lambda$ score of 93.6\%.
 This suggests that the planar representation of Oakland's drivable street network overstates the number of intersections --- and thus, the network's connectivity --- by 8.9\% city-wide and understates the average edge length by 6.4\% city-wide.
-
-However, these indicators' values vary across the city.
+However, these indicators vary across the city.
 To explore this statistical variation, Table \ref{tab:samples_city} presents summary statistics of these planarity indicators across 100 square-mile samples of Oakland's drivable street network.
@@
 However, the samples range from spatial planarity lows of 56.9\% ($\phi$) and 63.7\% ($\lambda$) up to highs of 100\%.
-67\% of the samples pass the formal planarity test; however, 63\% of the samples are at least somewhat spatially nonplanar (i.e., $\phi < 1$).
+67\% of the samples pass the formal planarity test but 63\% of the samples are at least somewhat spatially nonplanar (i.e., $\phi < 1$).
 
@@
 
-These findings suggest that the street networks at the centers of most major cities are formally and spatially nonplanar.
+These findings suggest that the street networks at the centers of most major cities are both formally and spatially nonplanar.
 However, this depends on the scale of measurement: across an entire city there is likely to be at least one overpass or underpass somewhere, while individual neighborhoods or small towns may be entirely planar.
@@
 	\item{Average edge lengths are underestimated}
-	\item{Connectivity is misrepresented for routing, accessibility analysis, and other topological studies}
+	\item{Connectivity is misrepresented for routing, accessibility analysis, and topological studies}
 \end{enumerate}
 
-Our results demonstrate how this is a bigger problem in Los Angeles than in Florence.
+Our results empirically demonstrate how this is a bigger problem in Los Angeles than in Florence.
 But even in Florence the walking network is both spatially and formally nonplanar due to the \textit{sottopassaggio} (pedestrian subway) near Stazione di Santa Maria Novella, its central train station.
@@
 A graph is not planar because its edges \emph{usually} intersect only at nodes: by definition it is planar because its edges \emph{exclusively} intersect at nodes.
-These results suggest that it cannot be universally claimed that urban street networks are planar graphs --- formally or spatially.
+Urban street networks are not universally planar graphs, formally or spatially.
 However, as \citet{newman_networks:_2010} points out, debating the semantics of formal planarity may be missing the point -- more interesting is the \emph{extent} to which a network is planar.
@@
 Our results suggest that the answer depends on the study site and on the type of analysis.
-In limited circumstances --- where the circulation network exhibits few (or ideally zero) underpasses, overpasses, or grade-separation --- then perhaps yes, especially if the study focuses on polygonal spatial analysis.
+In limited circumstances --- where the circulation network exhibits few (or ideally zero) underpasses, overpasses, or grade-separation --- then certainly yes, especially if the study focuses on polygonal spatial analysis.
 But universally we cannot answer yes, especially for topological studies.
@@
 Do they provide useful models for this type of research?
-Again, only in limited circumstances, such as the drivable networks in the three Italian cities we analyzed.
-Elsewhere, Table \ref{tab:world_cities} showed how the results of common urban form analyses such as intersection counts are overstated by planar models (16\% on average in the drivable networks), while average street segment lengths are consequently understated (10\% on average in the drivable networks).
-Moreover, this misrepresentation behaves inconsistently from place to place: Figure \ref{fig:world_map_bw} and Table \ref{tab:samples_city} demonstrated how the magnitude of bias varies across cities and modes of urbanization.
+Again, only in certain circumstances, such as the urban forms of Bologna and Florence.
+Meanwhile, in our central Los Angeles network, they overcount intersections by 72\% and underestimate edge lengths by 37\%.
+In aggregate, the results of common urban form analyses such as intersection counts are overstated by planar models (by 16\% on average in the drivable networks), while average street segment lengths are consequently understated (by 10\% on average in the drivable networks).
+Moreover, this misrepresentation behaves inconsistently from place to place: Figure \ref{fig:world_map_bw} and Table \ref{tab:samples_city} demonstrated how the magnitude of bias varies within and across cities and modes of urbanization.
 
-We may be able to shoehorn real-world data into a chosen model and perform a desired computation, but we lose the underlying ability to reason when the real world no longer obeys the rules and constraints of the chosen model.
+We may be able to shoehorn real-world data into a chosen model and perform a desired computation, but we lose the underlying ability to reason with it when the real world no longer obeys the rules and constraints of the model.
 Although planar simplifications offer algorithmic tractability and allow for the polygonal analysis of urban blocks, these results demonstrate how they can model urban street networks poorly and bias analytical results.
@@
 The $\phi$ and $\lambda$ scores of the major US city centers indicate the greater three-dimensionality of their transportation infrastructure compared to that of the European city centers.
-These US city centers feature more automobile-oriented roadways and are often bounded by grade-separated freeways.
+These US city centers experienced massive investment in automobile-oriented roadways and are often bounded by grade-separated freeways.
 By contrast, these European city centers feature less three-dimensionality as their streets and paths tend to be at-grade.
 Examining how $\phi$ and $\lambda$ change over time in different cities could reveal the type and pace of urban development.
-Future research can further explore how these indicators correlate with other measures of urbanization, development, and era.
 
@@
 
-\section{Conclusion}
+Future research can further explore how these indicators correlate with other measures of urbanization, development, investment, technology, politics, and era.
+While this study focused on measures of the urban form, particularly intersection counts and block lengths, future research can also further explore impacts of planar simplifications on topological analyses and routing.
+
+\subsection{Conclusion}
 
 Street networks are often defined and modeled as planar graphs in the urban studies literature despite typically being nonplanar in reality.
-This study tested the formal planarity of central urban street networks and  empirically demonstrated how planar simplifications affect analysis results.
-It developed two new measures of the extent to which they are nonplanar --- the Spatial Planarity Ratio $\phi$ and the Edge Length Ratio $\lambda$ ---  to characterize urbanization particularly through the transportation infrastructure's three-dimensionality.
-Although planar models may be useful in limited circumstances, they behave inconsistently within and across cities by misrepresenting connectivity, accessibility, routing, intersection counts and densities, and street segment lengths.
+This study tested the formal planarity of central urban street networks empirically and demonstrated how planar simplifications inconsistently affect analysis results.
+It developed two new measures of the extent to which spatial networks are nonplanar --- the Spatial Planarity Ratio, $\phi$, and the Edge Length Ratio, $\lambda$ ---  to characterize urbanization particularly through the transportation infrastructure's three-dimensionality.
+Although planar models are useful in certain circumstances, they behave inconsistently within and across cities by misrepresenting connectivity, accessibility, routing, intersection counts and densities, and street segment lengths.
 

### 2018-02-07 manuscript.tex
revise discussion

@@
 \runninghead{Geoff Boeing}
-\title{Street Network Representation and Nonplanarity in Urban Form Studies}
+\title{Planarity and Street Network Representation in Urban Form Studies}
 \author{Geoff Boeing \affilnum{1}}
@@
 Models of street networks underlie research in urban travel behavior, accessibility, design patterns, and morphology.
-These models are commonly defined as \enquote{planar,} meaning that they can be represented in two dimensions without any underpasses or overpasses.
+These models are commonly defined as planar, meaning that they can be represented in two dimensions without any underpasses or overpasses.
 However, real-world urban street networks exist in three-dimensional space and frequently feature grade separation such as bridges and tunnels.
-Planar simplifications affect the results of real-world street network analysis.
-A city's extent of nonplanarity can describe its infrastructure and urban development.
+Thus, planar simplifications affect the results of real-world street network analysis and a city's extent of nonplanarity can describe its infrastructure and urban development.
 This study measures the nonplanarity of drivable and walkable street networks in the centers of 50 cities worldwide, then examines the variation of nonplanarity across a single city.
@@
 
-The urban studies literature often defines street networks as \enquote{planar} or \enquote{approximately planar,} meaning they can be well-represented by a two-dimensional model that inherently precludes overpasses or underpasses.
-However, this claim has not been well studied or defined empirically.
+The urban studies literature often defines street networks as \enquote{planar} or \enquote{approximately planar,} meaning they can be well-represented by a simplified two-dimensional model that inherently precludes overpasses or underpasses.
+However, the analytical impacts of this computationally-useful simplification have not been studied empirically.
 In the real world, urban street networks are embedded in three-dimensional space and often feature grade separation, bridges, and tunnels.
@@
 It also presents two new measures of the \enquote{extent of nonplanarity} that describe the urban form and transportation infrastructure and can be generalized to other types of spatial networks.
-Despite common assumptions in the literature that urban street networks are planar graphs, this study finds that they generally are nonplanar and empirically quantifies how they misrepresent the street networks of many urban centers.
-Further, the magnitude of this bias varies substantially across cities and urbanization types.
+This study empirically quantifies how planarity misrepresents the street networks of many urban centers: while some street networks are well-modeled by planar graphs, others are substantially misrepresented and the magnitude of this bias varies across cities and urbanization types.
 
@@
 Next it discusses the methods used to acquire and analyze the street networks in this study.
-Then it present the results of this analysis before concluding with a discussion of implications for street network research and urban form studies.
+Then it presents the results of this analysis before concluding with a discussion of implications for street network research and urban form studies.
 
@@
 For example, a node's \emph{degree} is a topological trait that represents how many edges connect to that node.
+
 A \emph{planar} graph can be drawn on a two-dimensional plane without any of its edges crossing each other, except where they intersect at nodes.
@@
 Street networks are embedded in space, which provides them with geometry --- such as geographical coordinates, lengths, areas, shapes, and angles --- along with their topology.
+This creates a minor wrinkle when we consider planarity: we must distinguish between a graph's mathematical/topological planarity, which we refer to as \emph{formal planarity}, and the planarity of its real-world spatial embedding, which we refer to as \emph{spatial planarity}.
+For example, a street network might be spatially nonplanar due to its embedding in space (i.e., it contains overpasses or underpasses in the real world), but it could still be formally planar.
+If we \enquote{redraw} the graph by moving its nodes and edges around in space without changing how they are connected to one another (i.e., altering its geometry without altering its topology), there may exist some alternative spatial embedding that prevents edges crossing anywhere but at nodes \citep[p.~6]{barthelemy_morphogenesis_2017}.
+In such a case, the street network is formally planar from a topological perspective, but its real-world embedding is spatially nonplanar.
 
@@
 \end{table*}
-
-This creates a minor wrinkle when we consider planarity: we must distinguish between a graph's mathematical/topological planarity, which we refer to as \emph{formal planarity}, and the planarity of its particular spatial embedding, which we refer to as \emph{spatial planarity}.
-For example, a street network might be spatially nonplanar due to its embedding in space (i.e., it contains overpasses or underpasses in the real world), but it could still be formally planar.
-If we \enquote{redraw} the graph by moving its nodes and edges around in space without changing how they are connected to one another (i.e., altering its geometry without altering its topology), there may exist some spatial embedding that prevents edges crossing anywhere but at nodes \citep[p.~6]{barthelemy_morphogenesis_2017}.
-In such a case, the street network is formally planar from a topological perspective, but its real-world embedding is spatially nonplanar.
 
@@
 The left network comprises a set of paths and streets without any bridges or tunnels.
-In two-dimensions, its edges intersect each other only at nodes so it is by definition planar.
+In two dimensions, its edges intersect each other only at nodes so it is by definition planar.
 The right network comprises surface streets and a grade-separated freeway interchange with many overpasses and underpasses.
-In two-dimensions, its edges sometimes cross each other at non-nodes (i.e., overpasses), so it is spatially nonplanar.
+In two dimensions, its edges sometimes cross each other at non-nodes (i.e., overpasses), so it is spatially nonplanar.
 Furthermore it is impossible to redraw the graph by moving its nodes and edges around in space such that its edges only intersect at nodes.
@@
 
-In the urban studies literature, street networks are commonly referred to as planar graphs.
-Table \ref{tab:planar_quotes} presents a survey of statements and reasoning around this claim.
-Some authors prefer to hedge slightly, arguing that street networks are \emph{approximately} or \emph{essentially} planar graphs that are close enough to be well-modeled as such.
+In the urban studies literature, street networks are often modeled as planar graphs.
+Table \ref{tab:planar_quotes} surveys recent statements and reasoning around this approach.
+Some authors prefer to hedge slightly, arguing that street networks are \emph{approximately} or \emph{essentially} planar, but close enough to be well-modeled as such.
 
 If street networks can be sufficiently well-modeled by planar graphs, there are certain methodological benefits to doing so.
-Planar graphs offer computational simplicity and tractability.
-They enable simple polygonal analysis of city blocks and form \citep{fohl_non-planar_1996,barthelemy_paths_2017} as well as the \emph{alpha} and \emph{gamma} indices popular in road network analysis \citep{eppstein_studying_2008}.
-In mathematics, a bijection exists planar graphs and trees, and classifying planar graphs presents a trivial problem \citep{louf_typology_2014}.
-Planar graphs are easier to visualize and can be faster to run algorithms on \citep{liebers_planarizing_2001}.
+Planar graphs offer computational simplicity and algorithmic tractability.
+They enable simple polygonal analysis of city blocks and form \citep{fohl_non-planar_1996,barthelemy_paths_2017} and the calculation of the \emph{alpha} and \emph{gamma} indices popular in road network analysis \citep{eppstein_studying_2008}.
+In mathematics, a bijection exists between planar graphs and trees, and classifying planar graphs is a trivial problem \citep{louf_typology_2014}.
+Planar graphs are easier to visualize and can be much faster to run algorithms on \citep{liebers_planarizing_2001}.
 Accordingly, \citet[p.~3]{barthelemy_spatial_2011} argues that \enquote{planar spatial networks are the most important and most studies have focused on these examples}.
-But in contrast, \citet{masucci_random_2009} and \citet{masucci_limited_2013} argue that planar graphs remain a compelling research domain for urban scholars because they were understudied until recently for two reasons: they appear topologically trivial and planarity does not lend itself to certain popular graph-theoretic analyses.
-Discussing the open research area around street networks as planar graphs, \citet[p.~1]{viana_simplicity_2013} state, \enquote{there is still a lack of global, high-level metrics allowing to characterize their structure and geometrical patterns.}
+But in contrast, \citet{masucci_random_2009} and \citet{masucci_limited_2013} argue that planar graphs remain a compelling research domain for urban scholars because they were understudied until recently for two reasons: they appear topologically trivial and planarity does not lend itself to certain popular analyses of nonplanar aspatial graphs.
 
-Despite the computational advantages of planar simplifications, street networks obviously are often nonplanar in reality \citep[p.~7]{levinson_network_2012}: many include at least one overpass or underpass that results in the failure of formal proofs of their planarity, such as the \citet{kuratowski_sur_1930} theorem or the \cite{hopcroft_efficient_1974} algorithm \citep[cf.][]{gastner_spatial_2006}.
+Researchers use planar models for a useful simplification of real-world complexity; most street networks obviously are nonplanar in reality \citep[p.~7]{levinson_network_2012}.
+Many include at least one overpass or underpass that results in the failure of formal proofs of planarity, such as the \citet{kuratowski_sur_1930} theorem or the \cite{hopcroft_efficient_1974} algorithm \citep[cf.][]{gastner_spatial_2006}.
 \citet[p.~199]{jiang_object-oriented_2010} note that \enquote{quite often the transportation network has overpasses and underpasses that require a non-planar network representation.} \citet[p.~1258]{fischer_spatial_2014} explains that \enquote{for many infrastructure networks, {[planarity]} is approximately true, although bridges and tunnels in ground-transport networks are an obvious (but generally minor) exception.} However, the presence of such nonplanar elements can vex models.
-\enquote{The planar network data model has received widespread acceptance and use.
-Despite its popularity, the model has limitations for some areas of transportation analysis, especially where complex network structures are involved.
+\enquote{Despite its popularity, the [planar] model has limitations for some areas of transportation analysis, especially where complex network structures are involved.
 One major problem is caused by the planar embedding requirement...
@@
 
-Given these issues, there are some unanswered questions in the theoretical and empirical literature.
+Given these issues, there are some unanswered questions in the empirical literature.
 What do \emph{approximately planar} and \emph{well-modeled} mean for street network research?
-How close is \enquote{close enough} for a planar graph to competently model a formally nonplanar street network?
+How do planar simplifications impact the results of street network analyses?
 Do the biases of planar models behave consistently across geographies and development eras or do they misrepresent different cities to different extents?
-And if street networks are at least generally approximately planar, how can we measure just how planar or nonplanar a given street network is?
+And if street networks are at least generally approximately planar, how can we measure the extent to which a given street network is nonplanar?
 
@@
 This study develops two new measures of the extent to which a spatial network is planar.
-It empirically analyzes various city centers worldwide to quantify how planar models impact street network analysis results and the extent to which bias (i.e., model misrepresentation) varies across places and types of urbanization.
+It empirically analyzes various city centers worldwide to quantify how planar models impact street network analysis results (focusing on the common urban form measures of intersection density and block length) and the extent to which bias (i.e., model misrepresentation) varies across places and types of urbanization.
 Finally, it considers what these indicators suggest about the urban form and transportation infrastructure in different cities.
@@
 
-\subsection{Are street networks planar graphs?}
-
-These findings suggest that the street networks at the centers of most major cities are both formally and spatially nonplanar.
+Street networks at the centers of most major cities are both formally and spatially nonplanar.
 However, this depends on the scale of measurement: across an entire city there is likely to be at least one overpass or underpass somewhere, while individual neighborhoods or small towns may be entirely planar.
 The type and era of urbanization represent another factor.
-Medieval European towns or informal settlements in the global south may contain fewer grade-separated roads --- and thus would be more planar --- than 20th-century American or 21st-century Chinese metropolises.
-This results from the prevailing transportation and engineering technologies when the urban form developed, as well as local terrain, wealth, culture, and politics \citep{southworth_street_1995}.
+Old European towns or informal settlements in the Global South may contain fewer grade-separated roads --- and thus would be more planar --- than 20th-century American or 21st-century Chinese metropolises, resulting from prevailing transportation and engineering technologies when the urban form developed, as well as local terrain, wealth, culture, and politics \citep{southworth_street_1995}.
 
-Street networks are frequently spatially nonplanar because they are embedded in three dimensions, not two.
-But because they are usually \enquote{mostly} planar (average drivable $\phi$ of 88.5\% and average walkable $\phi$ of 92\%), typically with only a few overpasses or underpasses, they could often be described as \emph{approximately} planar.
-However, claiming that urban street networks universally are \enquote{planar,} or modeling them as such, misrepresents them in several ways:
-
-\begin{enumerate}
-	\item{Intersection counts are overestimated due to false nodes where grade-separated edges cross}
-	\item{Average edge lengths are underestimated}
-	\item{Connectivity is misrepresented for routing, accessibility analysis, and topological studies}
-\end{enumerate}
-
-Our results empirically demonstrate how this is a bigger problem in Los Angeles than in Florence.
-But even in Florence the walking network is both spatially and formally nonplanar due to the \textit{sottopassaggio} (pedestrian subway) near Stazione di Santa Maria Novella, its central train station.
-The results also suggest that spatial planarity is inconsistent both across cities as well as across different neighborhoods within individual cities.
-
-The problem with planar models is particularly pronounced around the downtowns of North American cities, due to the prevalence of freeways, bridges, and underpasses.
-Drivable networks are affected by these in particular, but even walkable networks are affected by pedestrian flyovers and subways, as in Florence.
-Furthermore, even networks of non-freeway, non-pedestrian-only surface streets could easily be nonplanar due to tunnels in hilly neighborhoods or bridges over rivers.
-
-A graph is not planar because its edges \emph{usually} intersect only at nodes: by definition it is planar because its edges \emph{exclusively} intersect at nodes.
-Urban street networks are not universally planar graphs, formally or spatially.
-However, as \citet{newman_networks:_2010} points out, debating the semantics of formal planarity may be missing the point -- more interesting is the \emph{extent} to which a network is planar.
+Street networks worldwide are commonly nonplanar because they are embedded in three dimensions, not two.
+But because they are usually \enquote{mostly} planar (average drivable $\phi$ of 88.5\% and average walkable $\phi$ of 92\%), typically with only a few overpasses or underpasses, many could be described as \emph{approximately} planar.
+But formally, a graph is not planar because its edges \emph{usually} intersect only at nodes: by definition it is planar because its edges \emph{exclusively} intersect at nodes.
+However, as \citet{newman_networks:_2010} points out, debating the semantics of formal planarity may be missing the point -- more interesting is the \emph{extent} to which a network is nonplanar.
 How do planar simplifications impact modeling and analysis?
 
-\subsection{Are street networks well-modeled by planar graphs?}
+\subsection{Planarity and model bias}
 
-As George Box famously said, \enquote{All models are wrong but some are useful.} Even if they are not formally planar, can street networks be simplified to planar graphs and still be usefully well-modeled?
+George Box famously said, \enquote{All models are wrong but some are useful.} Even if they are not formally planar, can street networks be simplified to planar graphs and still be usefully well-modeled?
 Our results suggest that the answer depends on the study site and on the type of analysis.
@@
 Again, only in certain circumstances, such as the urban forms of Bologna and Florence.
-Meanwhile, in our central Los Angeles network, they overcount intersections by 72\% and underestimate edge lengths by 37\%.
-In aggregate, the results of common urban form analyses such as intersection counts are overstated by planar models (by 16\% on average in the drivable networks), while average street segment lengths are consequently understated (by 10\% on average in the drivable networks).
-Moreover, this misrepresentation behaves inconsistently from place to place: Figure \ref{fig:world_map_bw} and Table \ref{tab:samples_city} demonstrated how the magnitude of bias varies within and across cities and modes of urbanization.
+Meanwhile, in our central Los Angeles network, planarity overcounted intersections by 72\% and underestimated edge lengths by 37\%.
+We can see this bias in aggregate too: the results of common urban form analyses such as intersection counts are overstated by planar models (by 16\% on average in the drivable networks), while average street segment lengths are consequently understated (by 10\% on average in the drivable networks).
+
+Moreover, this misrepresentation behaves inconsistently from place to place.
+The magnitude of bias varies within and across cities and modes of urbanization: these results empirically demonstrate how planar simplifications pose a bigger problem in Los Angeles than in Florence.
+But even in Florence the walking network is both spatially and formally nonplanar due to the \textit{sottopassaggio} (pedestrian subway) near Stazione di Santa Maria Novella, its central train station.
+The problem with planar models is particularly pronounced around the downtowns of North American cities, due to the prevalence of freeways, bridges, and underpasses.
+Drivable networks are affected by these in particular, but even walkable networks are affected by pedestrian flyovers and subways, as in Florence.
+Furthermore, even networks of non-freeway, non-pedestrian-only surface streets could easily be nonplanar due to tunnels in hilly neighborhoods or bridges over rivers.
+
+To summarize, planarity varies both across cities as well as across different neighborhoods within individual cities.
+Modeling urban street networks as planar graphs can bias urban form analyses in several ways:
+
+\begin{enumerate}
+	\item{Intersection counts are overestimated due to false nodes where grade-separated edges cross}
+	\item{Average edge lengths are underestimated}
+	\item{Connectivity is misrepresented for routing, accessibility analysis, and topological studies}
+	\item{Bias varies inconsistently across places}
+\end{enumerate}
 
 We may be able to shoehorn real-world data into a chosen model and perform a desired computation, but we lose the underlying ability to reason with it when the real world no longer obeys the rules and constraints of the model.
-Although planar simplifications offer algorithmic tractability and allow for the polygonal analysis of urban blocks, these results demonstrate how they can model urban street networks poorly and bias analytical results.
+Although planar simplifications offer algorithmic tractability and allow for the polygonal analysis of urban blocks, these results demonstrate how they can model many urban street networks poorly and bias analytical results.
 
@@
 Street networks are often defined and modeled as planar graphs in the urban studies literature despite typically being nonplanar in reality.
-This study tested the formal planarity of central urban street networks empirically and demonstrated how planar simplifications inconsistently affect analysis results.
+This study tested the formal and spatial planarity of central urban street networks empirically, demonstrating how planar simplifications inconsistently affect urban form analysis results.
 It developed two new measures of the extent to which spatial networks are nonplanar --- the Spatial Planarity Ratio, $\phi$, and the Edge Length Ratio, $\lambda$ ---  to characterize urbanization particularly through the transportation infrastructure's three-dimensionality.
-Although planar models are useful in certain circumstances, they behave inconsistently within and across cities by misrepresenting connectivity, accessibility, routing, intersection counts and densities, and street segment lengths.
+Although planar models are useful in some circumstances, they behave inconsistently within and across cities by misrepresenting connectivity, accessibility, routing, intersection densities, and block lengths.
 

### 2018-02-08 manuscript.tex
revise discussion

@@
 
-%\usepackage{draftwatermark}
-%\SetWatermarkText{DRAFT}
-%\SetWatermarkScale{1}
+\usepackage{draftwatermark}
+\SetWatermarkText{DRAFT}
+\SetWatermarkScale{1}
+\SetWatermarkLightness{0.9}
 
@@
 \runninghead{Geoff Boeing}
-\title{Planarity and Street Network Representation in Urban Form Studies}
+\title{Planarity and Street Network Representation in Urban Form Analysis}
 \author{Geoff Boeing \affilnum{1}}
@@
 These models are commonly defined as planar, meaning that they can be represented in two dimensions without any underpasses or overpasses.
-However, real-world urban street networks exist in three-dimensional space and frequently feature grade separation such as bridges and tunnels.
-Thus, planar simplifications affect the results of real-world street network analysis and a city's extent of nonplanarity can describe its infrastructure and urban development.
+However, real-world urban street networks exist in three-dimensional space and frequently feature grade separation such as bridges and tunnels: planar simplifications can be useful but might affect the results of real-world street network analysis.
+Further, a city's extent of nonplanarity could describe its infrastructure and urbanization.
 This study measures the nonplanarity of drivable and walkable street networks in the centers of 50 cities worldwide, then examines the variation of nonplanarity across a single city.
-It develops two new nonplanarity indicators of infrastructure and urbanization: the Spatial Planarity Ratio and the Edge Length Ratio.
+It develops two new indicators: the Spatial Planarity Ratio and the Edge Length Ratio.
 While some street networks are approximately planar, we empirically quantify how planar simplifications can inconsistently but drastically misrepresent routing, connectivity, intersection counts/densities, and block lengths.
@@
 
-This study tests the planarity of street networks around the world and demonstrates how planar simplifications alter the results of street network analysis.
+This study tests the planarity of street networks around the world and demonstrates how planar simplifications affect the results of street network analysis.
 It also presents two new measures of the \enquote{extent of nonplanarity} that describe the urban form and transportation infrastructure and can be generalized to other types of spatial networks.
@@
 
-Researchers use planar models for a useful simplification of real-world complexity; most street networks obviously are nonplanar in reality \citep[p.~7]{levinson_network_2012}.
-Many include at least one overpass or underpass that results in the failure of formal proofs of planarity, such as the \citet{kuratowski_sur_1930} theorem or the \cite{hopcroft_efficient_1974} algorithm \citep[cf.][]{gastner_spatial_2006}.
+Researchers use planar models as a useful simplification of real-world complexity but recognize that most street networks obviously are nonplanar in reality.
+Many include at least one overpass or underpass that results in the failure of formal proofs of planarity, such as the \citet{kuratowski_sur_1930} theorem or the \cite{hopcroft_efficient_1974} algorithm \citep[cf.][]{gastner_spatial_2006,levinson_network_2012}.
 \citet[p.~199]{jiang_object-oriented_2010} note that \enquote{quite often the transportation network has overpasses and underpasses that require a non-planar network representation.} \citet[p.~1258]{fischer_spatial_2014} explains that \enquote{for many infrastructure networks, {[planarity]} is approximately true, although bridges and tunnels in ground-transport networks are an obvious (but generally minor) exception.} However, the presence of such nonplanar elements can vex models.
@@
 Such measures would be particularly useful for street networks, as the extent to which a network is (or is not) planar can characterize the nature of its circulation infrastructure and urban form.
-For instance, late 20th-century freeway-oriented American cities might exhibit lower planarity than older walkable European cities.
-What about informal settlements in developing countries or rapidly urbanizing Chinese cities?
+For instance, wealthy 20th- and 21st-century cities oriented around automobility and freeways might exhibit more nonplanarity than do older cities or informal settlements more oriented around walking.
 Beyond the question of graph model goodness-of-fit, such indicators could provide useful information about urban development, civil infrastructure, and transportation system character.
@@
 In limited circumstances --- where the circulation network exhibits few (or ideally zero) underpasses, overpasses, or grade-separation --- then certainly yes, especially if the study focuses on polygonal spatial analysis.
-But universally we cannot answer yes, especially for topological studies.
-Most egregiously, imposing planarity on a nonplanar street network forces false nodes at underpasses and overpasses, breaking routing and network-based accessibility modeling.
+But universally we cannot answer yes, especially for topological studies: most egregiously, imposing planarity on a nonplanar street network forces false nodes at underpasses and overpasses, breaking routing and network-based accessibility modeling.
 For this reason, nonplanar graphs have been the standard for decades in transportation engineering, real-world traffic assignment models, and routing engines.
@@
 Do they provide useful models for this type of research?
-Again, only in certain circumstances, such as the urban forms of Bologna and Florence.
+Again, only in certain circumstances, such as the urban form of Florence.
 Meanwhile, in our central Los Angeles network, planarity overcounted intersections by 72\% and underestimated edge lengths by 37\%.
-We can see this bias in aggregate too: the results of common urban form analyses such as intersection counts are overstated by planar models (by 16\% on average in the drivable networks), while average street segment lengths are consequently understated (by 10\% on average in the drivable networks).
-
-Moreover, this misrepresentation behaves inconsistently from place to place.
-The magnitude of bias varies within and across cities and modes of urbanization: these results empirically demonstrate how planar simplifications pose a bigger problem in Los Angeles than in Florence.
-But even in Florence the walking network is both spatially and formally nonplanar due to the \textit{sottopassaggio} (pedestrian subway) near Stazione di Santa Maria Novella, its central train station.
-The problem with planar models is particularly pronounced around the downtowns of North American cities, due to the prevalence of freeways, bridges, and underpasses.
-Drivable networks are affected by these in particular, but even walkable networks are affected by pedestrian flyovers and subways, as in Florence.
+this misrepresentation behaves inconsistently from place to place as the magnitude of bias varies within and across cities and modes of urbanization.
+Misrepresentation is particularly pronounced around the downtowns of North American cities because of the prevalence of freeways, bridges, and underpasses.
+Drivable networks are mainly affected by these features but even walkable networks are affected by pedestrian flyovers and subways.
+In Florence, the walking network is both spatially and formally nonplanar due to the \textit{sottopassaggio} (pedestrian subway) near Stazione di Santa Maria Novella, its central train station.
 Furthermore, even networks of non-freeway, non-pedestrian-only surface streets could easily be nonplanar due to tunnels in hilly neighborhoods or bridges over rivers.
 
+We can see this bias in aggregate too.
+The results of common urban form analyses such as intersection counts are overstated by planar models (by 16\% on average in the drivable networks), while average street segment lengths are consequently understated (by 10\% on average in the drivable networks).
 To summarize, planarity varies both across cities as well as across different neighborhoods within individual cities.
@@
 	\item{Connectivity is misrepresented for routing, accessibility analysis, and topological studies}
-	\item{Bias varies inconsistently across places}
 \end{enumerate}
 
-We may be able to shoehorn real-world data into a chosen model and perform a desired computation, but we lose the underlying ability to reason with it when the real world no longer obeys the rules and constraints of the model.
-Although planar simplifications offer algorithmic tractability and allow for the polygonal analysis of urban blocks, these results demonstrate how they can model many urban street networks poorly and bias analytical results.
+We may be able to shoehorn real-world data into a chosen model and perform a desired computation, but we lose the underlying ability to reason with it when the real world no longer obeys the model's rules and assumptions.
+Although planar simplifications offer algorithmic tractability and allow for the polygonal analysis of urban blocks, these results demonstrate how they can model certain urban street networks poorly and bias analytical results.
 

### 2018-02-08 figures_tables/table_planar_quotes.tex
reorganize repo

--- 2018-01-19 table_planar_quotes.tex
@@
 \enquote{Street networks are planar graphs composed of junctions and street segments...} \citep[p.~18]{batty_network_2005} \\ \hline
-
-\enquote{The number of long-range connections and the number of edges that can be connected to a single node are limited by the spatial embedding.
-This is particularly evident in planar networks e.g., those networks forming vertices whenever two edges cross, as urban streets or ant networks of galleries...} \citep[p.~1]{crucitti_centrality_2006} \\ \hline
 
@@
 
+\enquote{Modeling a road network as a planar graph seems very natural.} \citep[p.~42]{aldous_routed_2016} \\ \hline
+
 \enquote{In city science, planar networks are extensively used to represent various infrastructure networks.

### 2018-02-08 main.tex
reorganize repo

--- 2018-02-08 manuscript.tex
@@
 	\label{tab:planar_quotes}
-	\input{table_planar_quotes.tex}
+	\input{figures_tables/table_planar_quotes.tex}
 \end{table*}
@@
 	\center
-	\includegraphics[width=0.48\textwidth]{planar_vs_not.png}
+	\includegraphics[width=0.48\textwidth]{figures_tables/planar_vs_not.png}
 	\caption{Formally and spatially planar (left) and nonplanar (right) road networks.}
@@
 	\label{tab:world_cities}
-	\input{table_world_cities.tex}
+	\input{figures_tables/table_world_cities.tex}
 \end{table*}
@@
 	\center
-	\includegraphics[width=\textwidth]{world_map_bw.png}
+	\includegraphics[width=\textwidth]{figures_tables/world_map_bw.png}
 	\caption{Map of world cities from Table \ref{tab:world_cities} grouped by $\phi$ terciles (lower values mean less planar).}
@@
 \begin{figure}[htbp]
-	\includegraphics[width=0.48\textwidth]{regression_phi_split.png}
+	\includegraphics[width=0.48\textwidth]{figures_tables/regression_phi_split.png}
 	\caption{Log-log plot of $\lambda$ vs $\phi$ by network type with simple regression lines.}
@@
 	\label{tab:samples_city}
-	\input{table_samples_city.tex}
+	\input{figures_tables/table_samples_city.tex}
 \end{table}

### 2018-02-08 main.tex
edits

@@
 In the urban studies literature, street networks are often modeled as planar graphs.
+Many street networks are obviously nonplanar in reality, but researchers build these planar models as a useful simplification of real-world complexity.
 Table \ref{tab:planar_quotes} surveys recent statements and reasoning around this approach.
-Some authors prefer to hedge slightly, arguing that street networks are \emph{approximately} or \emph{essentially} planar, but close enough to be well-modeled as such.
+Some scholars argue that street networks are planar graphs, while others prefer to hedge slightly, arguing that street networks are \emph{approximately} or \emph{essentially} planar but close enough to be well-modeled as such.
 
@@
 
-Researchers use planar models as a useful simplification of real-world complexity but recognize that most street networks obviously are nonplanar in reality.
-Many include at least one overpass or underpass that results in the failure of formal proofs of planarity, such as the \citet{kuratowski_sur_1930} theorem or the \cite{hopcroft_efficient_1974} algorithm \citep[cf.][]{gastner_spatial_2006,levinson_network_2012}.
+Although planar models can be computationally useful, real-world street networks often include at least one overpass or underpass that results in the failure of formal proofs of planarity, such as the \citet{kuratowski_sur_1930} theorem or the \cite{hopcroft_efficient_1974} algorithm \citep[cf.][]{gastner_spatial_2006,levinson_network_2012}.
 \citet[p.~199]{jiang_object-oriented_2010} note that \enquote{quite often the transportation network has overpasses and underpasses that require a non-planar network representation.} \citet[p.~1258]{fischer_spatial_2014} explains that \enquote{for many infrastructure networks, {[planarity]} is approximately true, although bridges and tunnels in ground-transport networks are an obvious (but generally minor) exception.} However, the presence of such nonplanar elements can vex models.
@@
 How do planar simplifications impact the results of street network analyses?
-Do the biases of planar models behave consistently across geographies and development eras or do they misrepresent different cities to different extents?
+Do the biases of planar models behave consistently across geographies and development eras or do they misrepresent different cities (or neighborhoods) to different extents?
 And if street networks are at least generally approximately planar, how can we measure the extent to which a given street network is nonplanar?
@@
 Such measures would be particularly useful for street networks, as the extent to which a network is (or is not) planar can characterize the nature of its circulation infrastructure and urban form.
-For instance, wealthy 20th- and 21st-century cities oriented around automobility and freeways might exhibit more nonplanarity than do older cities or informal settlements more oriented around walking.
+For instance, wealthy 20th- and 21st-century cities oriented around automobility and freeways might exhibit more nonplanarity than do older cities or informal settlements oriented around walking.
 Beyond the question of graph model goodness-of-fit, such indicators could provide useful information about urban development, civil infrastructure, and transportation system character.
@@
 
-This study develops two new measures of the extent to which a spatial network is planar.
+This study builds on this rich body of street network research by developing two new measures of the extent to which a spatial network is planar.
 It empirically analyzes various city centers worldwide to quantify how planar models impact street network analysis results (focusing on the common urban form measures of intersection density and block length) and the extent to which bias (i.e., model misrepresentation) varies across places and types of urbanization.
@@
 	\includegraphics[width=\textwidth]{figures_tables/world_map_bw.png}
-	\caption{Map of world cities from Table \ref{tab:world_cities} grouped by $\phi$ terciles (lower values mean less planar).}
+	\caption{Map of world cities from Table \ref{tab:world_cities} grouped by drivable $\phi$ terciles (lower values mean less planar).}
 	\label{fig:world_map_bw}
@@
 All three Italian cities demonstrate perfect planarity in their centers' drivable networks, but not in their walkable networks.
-The extent of nonplanarity is not consistent across network types: Dallas's walkable $\phi$ is 61\% greater than its drivable $\phi$, while Geneva's drivable $\phi$ is 19\% greater than its walkable $\phi$.
+Further, the extent of nonplanarity is not consistent across network types: Dallas's walkable $\phi$ is 61\% greater than its drivable $\phi$, while Geneva's drivable $\phi$ is 19\% greater than its walkable $\phi$.
 Figure \ref{fig:world_map_bw} maps the distribution of $\phi$ values around the world.
@@
 The indicators' log-log relationship is linear, positive, and very strong (drivable $r^2=0.98$ and walkable $r^2=0.99$).
-The coefficients of determination tell us that these two indicators unsurprisingly provide redundant statistical information about the extent to which a network is spatially planar.
+The coefficients of determination suggest that these two indicators unsurprisingly provide redundant statistical information about the extent to which a network is spatially planar.
 However, each assesses different implications of this bias for measuring the urban form.
@@
 Oakland's city-wide street network is formally nonplanar.
-In terms of spatial planarity, the city has a $\phi$ score of 91.8\% and a $\lambda$ score of 93.6\%.
+In terms of spatial planarity, the city has a $\phi$ of 91.8\% and a $\lambda$ of 93.6\%.
 This suggests that the planar representation of Oakland's drivable street network overstates the number of intersections --- and thus, the network's connectivity --- by 8.9\% city-wide and understates the average edge length by 6.4\% city-wide.
-However, these indicators vary across the city.
-To explore this statistical variation, Table \ref{tab:samples_city} presents summary statistics of these planarity indicators across 100 square-mile samples of Oakland's drivable street network.
+However, these indicators vary across the city: Table \ref{tab:samples_city} presents summary statistics of these planarity indicators across 100 square-mile samples of Oakland's drivable street network.
 Our samples' mean $\phi$ and $\lambda$ scores are reasonably close to the city-wide values.
@@
 The type and era of urbanization represent another factor.
-Old European towns or informal settlements in the Global South may contain fewer grade-separated roads --- and thus would be more planar --- than 20th-century American or 21st-century Chinese metropolises, resulting from prevailing transportation and engineering technologies when the urban form developed, as well as local terrain, wealth, culture, and politics \citep{southworth_street_1995}.
+Old European towns or informal settlements in the Global South may contain fewer grade-separated roads --- and thus would be more planar --- than 20th-century American or 21st-century Chinese metropolises, a result of prevailing transportation and engineering technologies when the urban form developed, as well as local terrain, wealth, culture, and politics \citep{southworth_street_1995}.
 
@@
 Do they provide useful models for this type of research?
-Again, only in certain circumstances, such as the urban form of Florence.
+Again, only in certain circumstances, such as the urban form of Florence where the planar bias is negligible.
 Meanwhile, in our central Los Angeles network, planarity overcounted intersections by 72\% and underestimated edge lengths by 37\%.
@@
 Misrepresentation is particularly pronounced around the downtowns of North American cities because of the prevalence of freeways, bridges, and underpasses.
-Drivable networks are mainly affected by these features but even walkable networks are affected by pedestrian flyovers and subways.
+
+Drivable networks are primarily affected by these features, but even walkable networks are affected by pedestrian flyovers and subways.
 In Florence, the walking network is both spatially and formally nonplanar due to the \textit{sottopassaggio} (pedestrian subway) near Stazione di Santa Maria Novella, its central train station.
-Furthermore, even networks of non-freeway, non-pedestrian-only surface streets could easily be nonplanar due to tunnels in hilly neighborhoods or bridges over rivers.
+But even networks of non-freeway, non-pedestrian-only surface streets could easily be nonplanar due to tunnels in hilly neighborhoods or bridges over rivers.
 
-We can see this bias in aggregate too.
-The results of common urban form analyses such as intersection counts are overstated by planar models (by 16\% on average in the drivable networks), while average street segment lengths are consequently understated (by 10\% on average in the drivable networks).
+We can see the impact of this bias in aggregate across our study sites: the results of common urban form analyses such as intersection counts are overstated by planar models (by 16\% on average in the drivable networks), while average street segment lengths are consequently understated (by 10\% on average in the drivable networks).
 To summarize, planarity varies both across cities as well as across different neighborhoods within individual cities.
@@
 
-We might repurpose this discrepancy between planar and nonplanar models of a street network to reveal something useful about urban form.
-The $\phi$ and $\lambda$ scores of the major US city centers indicate the greater three-dimensionality of their transportation infrastructure compared to that of the European city centers.
-These US city centers experienced massive investment in automobile-oriented roadways and are often bounded by grade-separated freeways.
+We might repurpose this discrepancy between planar and nonplanar models of a street network to describe characteristics of the urban form.
+For example, the $\phi$ and $\lambda$ scores of these US city centers indicate the greater three-dimensionality of their transportation infrastructure compared to that of these European city centers.
+During the 20\textsuperscript{th} century, US cities experienced massive investment in automobile-oriented roadways and are commonly laced with grade-separated freeways.
 By contrast, these European city centers feature less three-dimensionality as their streets and paths tend to be at-grade.
@@
 
-Most street networks are formally nonplanar while demonstrating $\phi$ and $\lambda$ values that are only slightly nonplanar.
-Approximate planarity is common across cities because of costs, technology, and politics: it is expensive to engineer a three-dimensional network with z-axis extents (nearly) as extensive as the x- and y-axes.
+Many street networks are formally nonplanar while demonstrating $\phi$ and $\lambda$ values that are only slightly nonplanar.
+This approximate planarity is common across cities because of costs, technology, and politics: it is expensive to engineer an extensively three-dimensional network.
 Furthermore, a city fully consumed by overpasses and tunnels would degrade quality of life, making it politically infeasible.
@@
 
-Street networks are often defined and modeled as planar graphs in the urban studies literature despite typically being nonplanar in reality.
-This study tested the formal and spatial planarity of central urban street networks empirically, demonstrating how planar simplifications inconsistently affect urban form analysis results.
+Street networks are often defined and modeled as planar graphs in the urban studies literature --- despite typically being nonplanar in reality --- to simplify real-world complexity and make certain analyses tractable.
+This study measured the formal and spatial planarity of central urban street networks empirically, demonstrating how planar simplifications inconsistently affect urban form analysis results.
 It developed two new measures of the extent to which spatial networks are nonplanar --- the Spatial Planarity Ratio, $\phi$, and the Edge Length Ratio, $\lambda$ ---  to characterize urbanization particularly through the transportation infrastructure's three-dimensionality.
-Although planar models are useful in some circumstances, they behave inconsistently within and across cities by misrepresenting connectivity, accessibility, routing, intersection densities, and block lengths.
+Although planar graphs are useful in some circumstances, they behave inconsistently within and across cities by misrepresenting connectivity, accessibility, routing, intersection densities, and block lengths.
+Individual urban street network analyses must consider these location-dependent tractability and accuracy trade-offs when choosing between planar and nonplanar models.
 

### 2018-02-08 main.tex
update figs

@@
 It develops two new indicators: the Spatial Planarity Ratio and the Edge Length Ratio.
-While some street networks are approximately planar, we empirically quantify how planar simplifications can inconsistently but drastically misrepresent routing, connectivity, intersection counts/densities, and block lengths.
+While some street networks are approximately planar, we empirically quantify how planar models can inconsistently but drastically misrepresent intersection counts/densities, block lengths, routing, and connectivity.
 
@@
 \begin{figure}[htbp]
-	\includegraphics[width=0.48\textwidth]{figures_tables/regression_phi_split.png}
-	\caption{Log-log plot of $\lambda$ vs $\phi$ by network type with simple regression lines.}
-	\label{fig:regression_split}
+	\includegraphics[width=0.48\textwidth]{figures_tables/regression_lambda_phi_linear.png}
+	\caption{Plot of $\lambda$ vs $\phi$ by network type with simple regression lines (dashed=drive, dotted=walk).}
+	\label{fig:regression_lambda_phi}
 \end{figure}
@@
 
-Figure \ref{fig:regression_split} depicts the relationship between the $\phi$ and $\lambda$ indicators across all 50 cities for both network types, as a log-log plot.
-The indicators' log-log relationship is linear, positive, and very strong (drivable $r^2=0.98$ and walkable $r^2=0.99$).
+Figure \ref{fig:regression_lambda_phi} depicts the relationship between $\phi$ and $\lambda$ across all 50 cities for both network types: it is linear, positive, and very strong (drivable $r^2=0.985$ and walkable $r^2=0.994$).
 The coefficients of determination suggest that these two indicators unsurprisingly provide redundant statistical information about the extent to which a network is spatially planar.

### 2018-02-16 main.tex
update fig and author
rework quotes table into text

@@
 
-\usepackage{draftwatermark}
+\usepackage[final]{draftwatermark}
 \SetWatermarkText{DRAFT}
@@
 
-\runninghead{Geoff Boeing}
+\runninghead{Author Redacted}
 \title{Planarity and Street Network Representation in Urban Form Analysis}
-\author{Geoff Boeing \affilnum{1}}
+\author{Author Redacted\affilnum{1}}
 \affiliation{\affilnum{1}Affiliation redacted}
@@
 
-\keywords{street network, GIS, urban form, transportation, urban design}
+\keywords{GIS, street networks, transportation, urban form, urban morphology}
 
@@
 In urban planning and transportation research, street networks are routinely used to calculate accessibility between origins and destinations or to compute indicators of the urban form, such as block sizes or intersection density and connectivity.
-Mathematical models of street networks, called \emph{graphs}, have grown ubiquitous in the urban studies literature in recent years as they have been used to model household travel patterns, access and equity, pedestrian volume, urban design patterns, spatial morphology, and location centrality and polycentricity \citep{marshall_street_2010,porta_alterations_2014,marshall_community_2014,hajrasouliha_impact_2015,parthasarathi_street_2015,knight_metrics_2015,gil_street_2016,zhong_revealing_2017}.
+Mathematical models of street networks, called \emph{graphs}, have grown ubiquitous in the urban studies literature in recent years as they have been used to model household travel patterns, access and equity, pedestrian volume, urban design patterns, spatial morphology, and location centrality and polycentricity \citep{marshall_street_2010,porta_alterations_2014,marshall_community_2014,hajrasouliha_impact_2015,parthasarathi_street_2015,knight_metrics_2015,xiao_identifying_2016,zhong_revealing_2017}.
 
@@
 
-\section{Background and Motivation}
+\section{Background and motivation}
+
+\subsection{Street network models}
 
@@
 
-\begin{table*}[htbp]
-	\centering
-	\caption{Survey of recent statements in the research literature regarding the representation of street networks as planar graphs.}
-	\label{tab:planar_quotes}
-	\input{figures_tables/table_planar_quotes.tex}
-\end{table*}
-
 \begin{figure}[tbp]
 	\center
-	\includegraphics[width=0.48\textwidth]{figures_tables/planar_vs_not.png}
+	\includegraphics[width=0.49\textwidth]{figures_tables/planar_vs_not.png}
 	\caption{Formally and spatially planar (left) and nonplanar (right) road networks.}
@@
 
-In the urban studies literature, street networks are often modeled as planar graphs.
-Many street networks are obviously nonplanar in reality, but researchers build these planar models as a useful simplification of real-world complexity.
-Table \ref{tab:planar_quotes} surveys recent statements and reasoning around this approach.
-Some scholars argue that street networks are planar graphs, while others prefer to hedge slightly, arguing that street networks are \emph{approximately} or \emph{essentially} planar but close enough to be well-modeled as such.
+\subsection{Planar models in urban research}
+
+In the urban studies literature, street networks are commonly modeled as planar graphs.
+Many street networks are obviously nonplanar in reality, but researchers use planar models as a potentially useful simplification of real-world complexity.
+Many scholars argue that street networks are planar graphs (e.g., \citealp[p.~18]{batty_network_2005}; \citealp[p.~521]{buhl_topological_2006}; \citealp[p.~1]{hu_topological_2008}; \citealp[p.~259]{masucci_random_2009}; \citealp[p.~114]{porta_networks_2010}; \citealp[p.~3]{strano_elementary_2012}; \citealp[p.~1]{masucci_limited_2013}; \citealp[p.~1074]{strano_urban_2013}; \citealp[p.~168]{law_defining_2017}).
+Others prefer to hedge slightly, arguing that street networks are approximately or essentially planar --- close enough to be well-modeled as such (e.g., \citealp[p.~6]{dill_measuring_2004}; \citealp[p.~3]{cardillo_structural_2006};  \citealp[p.~340]{xie_measuring_2007}; \citealp[p.~1]{barthelemy_modeling_2008}; \citealp[p.~3]{barthelemy_spatial_2011}; \citealp[pp.~563]{chan_urban_2011}; \citealp[p.~1]{gudmundsson_entropy_2013}; \citealp[p.~1]{viana_simplicity_2013}; \citealp[p.~2]{louf_typology_2014}; \citealp[p.~2191]{zhong_detecting_2014}; \citealp[p.~2]{wang_resilience_2015}; \citealp[p.~42]{aldous_routed_2016}; \citealp[p.~257]{barthelemy_paths_2017}).
 
@@
 For example, this misrepresentation would allow a routing engine to recommend a left turn at an \enquote{intersection} that is actually an overpass \citep[p.~6]{kwan_review_1996}.
-Intersection counts and densities would be overestimated in the presence of these false nodes.
-Consequently, edge lengths would be underestimated due to these artificial breakpoints splitting up street segments.
+Intersection counts and densities would be overestimated in the presence of these false nodes, and edge lengths would be underestimated due to these artificial breakpoints splitting up street segments.
+This consequently impacts interpretation of urban form density, grain, pattern, connectedness, and permeability.
 Finally, this bias might behave inconsistently across different kinds of cities and street network types based on the extents to which they are planar.
 
-Given these issues, there are some unanswered questions in the empirical literature.
+\subsection{Open questions}
+
+Given these issues, some unanswered questions remain in the empirical literature.
 What do \emph{approximately planar} and \emph{well-modeled} mean for street network research?
@@
 They also fail to adjust for the size, density, or real-world embedding of a spatial network.
-In his discussion of road networks and approximate spatial planarity, \citet[p.~133]{newman_networks:_2010} argues that due to these drawbacks no widely-accepted measure of the extent of nonplanarity has emerged and calls for the development of better indicators.
+Discussing road networks and approximate spatial planarity, \citet[p.~133]{newman_networks:_2010} argues that due to these drawbacks no widely-accepted measure of the extent of nonplanarity has emerged and calls for the development of better indicators.
 
@@
 This study builds on this rich body of street network research by developing two new measures of the extent to which a spatial network is planar.
-It empirically analyzes various city centers worldwide to quantify how planar models impact street network analysis results (focusing on the common urban form measures of intersection density and block length) and the extent to which bias (i.e., model misrepresentation) varies across places and types of urbanization.
+It empirically analyzes various city centers worldwide to quantify how planar models impact street network analysis results (focusing on the common urban form measures of intersection density and street segment length) and the extent to which bias (i.e., model misrepresentation) varies across places and types of urbanization.
 Finally, it considers what these indicators suggest about the urban form and transportation infrastructure in different cities.
@@
 Following \citet{jacobs_great_1995} and \citet{cardillo_structural_2006}, we analyze a consistently sized, square-mile network at the centers of 50 cities worldwide.
-This allows us to consistently examine central urban street networks without being swamped by metropolitan-scale variation or the idiosyncrasies of individual municipalities' spatial extents.
+This allows us to consistently examine central urban street networks without being swamped by metropolitan-scale variation or the idiosyncrasies of individual municipalities' spatial extents \citep[cf.][]{horner_excess_2002,gil_street_2017}.
 The 50 sampled cities span Africa, Asia, Australia, Europe, and North and South America.
@@
 \begin{figure*}[htbp]
-	\center
+	\centering
 	\includegraphics[width=\textwidth]{figures_tables/world_map_bw.png}
@@
 \begin{figure}[htbp]
-	\includegraphics[width=0.48\textwidth]{figures_tables/regression_lambda_phi_linear.png}
-	\caption{Plot of $\lambda$ vs $\phi$ by network type with simple regression lines (dashed=drive, dotted=walk).}
+	\centering
+	\includegraphics[width=0.49\textwidth]{figures_tables/regression_lambda_phi_linear.png}
+	\caption{Plot of $\lambda$ vs $\phi$ by network type with regression lines.}
 	\label{fig:regression_lambda_phi}
@@
 We can see the impact of this bias in aggregate across our study sites: the results of common urban form analyses such as intersection counts are overstated by planar models (by 16\% on average in the drivable networks), while average street segment lengths are consequently understated (by 10\% on average in the drivable networks).
+These impact our understanding of the urban form's density, grain, pattern, connectedness, and permeability.
 To summarize, planarity varies both across cities as well as across different neighborhoods within individual cities.
@@
 Future research can further explore how these indicators correlate with other measures of urbanization, development, investment, technology, politics, and era.
-While this study focused on measures of the urban form, particularly intersection counts and block lengths, future research can also further explore impacts of planar simplifications on topological analyses and routing.
+While this study focused on measures of the urban form, particularly intersection counts and street segment lengths, future research can also further explore impacts of planar simplifications on topological analyses and routing.
 
@@
 It developed two new measures of the extent to which spatial networks are nonplanar --- the Spatial Planarity Ratio, $\phi$, and the Edge Length Ratio, $\lambda$ ---  to characterize urbanization particularly through the transportation infrastructure's three-dimensionality.
-Although planar graphs are useful in some circumstances, they behave inconsistently within and across cities by misrepresenting connectivity, accessibility, routing, intersection densities, and block lengths.
+Although planar graphs are useful in some circumstances, they behave inconsistently within and across cities by misrepresenting connectivity, accessibility, routing, intersection densities, and street lengths.
 Individual urban street network analyses must consider these location-dependent tractability and accuracy trade-offs when choosing between planar and nonplanar models.
+
+\section{Acknowledgments}
+
+Redacted.
 

### 2018-02-17 main.tex
update figures and tables

@@
 \documentclass[Afour,sageh,times]{sagej}
+\usepackage[T1]{fontenc}
 \usepackage[utf8]{inputenc}
 \usepackage[USenglish]{babel}
+\usepackage{microtype}
 \usepackage{booktabs}
@@
 In urban planning and transportation research, street networks are routinely used to calculate accessibility between origins and destinations or to compute indicators of the urban form, such as block sizes or intersection density and connectivity.
-Mathematical models of street networks, called \emph{graphs}, have grown ubiquitous in the urban studies literature in recent years as they have been used to model household travel patterns, access and equity, pedestrian volume, urban design patterns, spatial morphology, and location centrality and polycentricity \citep{marshall_street_2010,porta_alterations_2014,marshall_community_2014,hajrasouliha_impact_2015,parthasarathi_street_2015,knight_metrics_2015,xiao_identifying_2016,zhong_revealing_2017}.
+Mathematical models of street networks, called \emph{graphs}, have grown ubiquitous in the urban studies literature in recent years as they have been used to model household travel patterns, access and equity, pedestrian volume, urban design patterns, spatial morphology, and location centrality and polycentricity \citep{marshall_street_2010,pflieger_switzerland_2010,porta_street_2012,lee_identifying_2014,porta_alterations_2014,marshall_community_2014,hajrasouliha_impact_2015,parthasarathi_street_2015,knight_metrics_2015,xiao_identifying_2016,zhong_revealing_2017}.
 
@@
 Following \citet{jacobs_great_1995} and \citet{cardillo_structural_2006}, we analyze a consistently sized, square-mile network at the centers of 50 cities worldwide.
-This allows us to consistently examine central urban street networks without being swamped by metropolitan-scale variation or the idiosyncrasies of individual municipalities' spatial extents \citep[cf.][]{horner_excess_2002,gil_street_2017}.
+This allows us to consistently examine central urban street networks without being swamped by metropolitan-scale variation or the idiosyncrasies of individual municipalities' spatial extents \citep{gil_street_2017}.
 The 50 sampled cities span Africa, Asia, Australia, Europe, and North and South America.
@@
 Although planar graphs are useful in some circumstances, they behave inconsistently within and across cities by misrepresenting connectivity, accessibility, routing, intersection densities, and street lengths.
-Individual urban street network analyses must consider these location-dependent tractability and accuracy trade-offs when choosing between planar and nonplanar models.
+Individual analyses must consider these location-dependent tractability and accuracy trade-offs when choosing between planar and nonplanar models.
 

### 2018-02-18 main.tex
update results values in text

@@
 It develops two new indicators: the Spatial Planarity Ratio and the Edge Length Ratio.
-While some street networks are approximately planar, we empirically quantify how planar models can inconsistently but drastically misrepresent intersection counts/densities, block lengths, routing, and connectivity.
+While some street networks are approximately planar, we empirically quantify how planar models can inconsistently but drastically misrepresent intersection density, street lengths, routing, and connectivity.
 
@@
 For example, a street network might be spatially nonplanar due to its embedding in space (i.e., it contains overpasses or underpasses in the real world), but it could still be formally planar.
-If we \enquote{redraw} the graph by moving its nodes and edges around in space without changing how they are connected to one another (i.e., altering its geometry without altering its topology), there may exist some alternative spatial embedding that prevents edges crossing anywhere but at nodes \citep[p.~6]{barthelemy_morphogenesis_2017}.
+If we \enquote{redraw} the graph by moving its nodes and edges around in space without changing how they are connected to one another (i.e., altering its geometry without altering its topology), there may exist some alternative spatial embedding that prevents edges crossing anywhere but at nodes \citep[for a thorough introduction see][p.~6]{barthelemy_morphogenesis_2017}.
 In such a case, the street network is formally planar from a topological perspective, but its real-world embedding is spatially nonplanar.
@@
 	\label{eq:spr_overstates}
-	1 - \frac{1}{\phi}
+	\frac{1}{\phi} - 1
 \end{equation}
@@
 Among the drivable street networks, only 20\% are formally planar.
-On average, they are 88.5\% spatially planar by the $\phi$ measure and 90\% by the $\lambda$ measure.
-The individual $\phi$ values indicate that spatial planarity ranges from a high of 100\% in seven of these cities to a low of 57\% in Moscow.
-The $\lambda$ values indicate that spatial planarity ranges from a high of 100\% in seven of these cities to a low of 63.5\% in Los Angeles.
-On average across these networks, planar models overcount intersections by 16\% and underestimate street segment lengths by 10\% (Equations \ref{eq:spr_overstates} and \ref{eq:elr_understates}).
+On average, they are 88\% spatially planar by the $\phi$ measure and 89\% by the $\lambda$ measure.
+The individual $\phi$ values indicate that spatial planarity ranges from a high of 100\% in six of these cities to a low of 54\% in Moscow.
+The $\lambda$ values indicate that spatial planarity ranges from a high of 100\% in six of these cities to a low of 60\% in Moscow.
+On average across these networks, planar models overcount intersections by 16\% and underestimate street segment lengths by 11\% (Equations \ref{eq:spr_overstates} and \ref{eq:elr_understates}).
 
 Among walkable street networks, only 10\% are formally planar.
-On average, they are 92\% spatially planar by the $\phi$ measure and 91.5\% by the $\lambda$ measure.
-The individual $\phi$ values indicate that spatial planarity ranges from a high of 100\% in two cities to a low of 67\% in Shanghai.
-The $\lambda$ values indicate that spatial planarity ranges from a high of 100\% in two cities to a low of 66\% in Shanghai.
-On average across these networks, planar models overcount intersections by 9\% and underestimate street segment lengths by 8.5\%.
+On average, they are 92\% spatially planar by the $\phi$ measure and 91\% by the $\lambda$ measure.
+The individual $\phi$ values indicate that spatial planarity ranges from a high of 100\% in two cities to a low of 66\% in Shanghai.
+The $\lambda$ values indicate that spatial planarity ranges from a high of 100\% in two cities to a low of 64\% in Shanghai.
+On average across these networks, planar models overcount intersections by 10\% and underestimate street segment lengths by 9\%.
 Fewer walkable than drivable networks are formally planar, but on average these walking networks are slightly more spatially planar than the driving networks.
@@
 Not all formally planar street networks are spatially planar.
-For example, Toronto's drivable network is formally planar but only 93\% ($\phi$) and 96\% ($\lambda$) spatially planar.
-In total, three drivable networks (Toronto, Jakarta, and Copenhagen) and three walkable networks (Dallas, Delhi, and Bologna) are formally planar but spatially nonplanar to various extents.
-About a third of the city centers studied demonstrate $\phi$ spatial planarity of 97\% or higher, suggesting that they are \enquote{approximately} planar.
+For example, Toronto's drivable network is formally planar but only 92\% ($\phi$) and 95\% ($\lambda$) spatially planar.
+In total, four drivable networks (Toronto, Jakarta, Florence, and Copenhagen) and three walkable networks (Dallas, Delhi, and Bologna) are formally planar but spatially nonplanar.
+About a third of the city centers studied demonstrate $\phi$ spatial planarity of 96\% or higher, suggesting that they are \enquote{approximately} planar.
 However, another third of the city centers are less than 87\% planar.
-Dallas, Los Angeles, and Moscow have $\phi$ values below 60\%, suggesting planar graphs poorly model these city centers.
-In fact, planarity overstates the intersection counts in these three networks by 67\%, 72\%, and 74\% respectively.
+Dallas, Los Angeles, and Moscow have $\phi$ values below 59\%, suggesting planar graphs poorly model these city centers.
+In fact, planarity overstates the intersection counts in these three networks by 71\%, 72\%, and 85\% respectively.
 
 Mogadishu is the only city studied that demonstrates perfect planarity across all three indicators for both network types.
-All three Italian cities demonstrate perfect planarity in their centers' drivable networks, but not in their walkable networks.
-Further, the extent of nonplanarity is not consistent across network types: Dallas's walkable $\phi$ is 61\% greater than its drivable $\phi$, while Geneva's drivable $\phi$ is 19\% greater than its walkable $\phi$.
+Cities like Milan and Barcelona demonstrate perfect planarity in their centers' drivable networks, but not in their walkable networks.
+Further, the extent of nonplanarity is not consistent across network types: Dallas's walkable $\phi$ is 65\% greater than its drivable $\phi$, while Geneva's drivable $\phi$ is 15\% greater than its walkable $\phi$.
 Figure \ref{fig:world_map_bw} maps the distribution of $\phi$ values around the world.
@@
 
-Figure \ref{fig:regression_lambda_phi} depicts the relationship between $\phi$ and $\lambda$ across all 50 cities for both network types: it is linear, positive, and very strong (drivable $r^2=0.985$ and walkable $r^2=0.994$).
+Figure \ref{fig:regression_lambda_phi} depicts the relationship between $\phi$ and $\lambda$ across all 50 cities for both network types: it is linear, positive, and very strong (drivable $r^2=0.990$ and walkable $r^2=0.993$).
 The coefficients of determination suggest that these two indicators unsurprisingly provide redundant statistical information about the extent to which a network is spatially planar.
@@
 Oakland's city-wide street network is formally nonplanar.
-In terms of spatial planarity, the city has a $\phi$ of 91.8\% and a $\lambda$ of 93.6\%.
-This suggests that the planar representation of Oakland's drivable street network overstates the number of intersections --- and thus, the network's connectivity --- by 8.9\% city-wide and understates the average edge length by 6.4\% city-wide.
-However, these indicators vary across the city: Table \ref{tab:samples_city} presents summary statistics of these planarity indicators across 100 square-mile samples of Oakland's drivable street network.
-Our samples' mean $\phi$ and $\lambda$ scores are reasonably close to the city-wide values.
-However, the samples range from spatial planarity lows of 56.9\% ($\phi$) and 63.7\% ($\lambda$) up to highs of 100\%.
+In terms of spatial planarity, the city has a $\phi$ of 91.6\% and a $\lambda$ of 92.9\%.
+This suggests that the planar representation of Oakland's drivable street network overstates the number of intersections --- and thus, the network's connectivity --- by 9.2\% city-wide and understates the average edge length by 7.1\% city-wide.
+However, these indicators vary across the city: Table \ref{tab:samples_city} presents summary statistics of these planarity indicators across the 100 square-mile samples of its drivable network.
+The samples' mean $\phi$ and $\lambda$ scores are reasonably close to the city-wide values.
+However, the samples range from spatial planarity lows of 56.9\% ($\phi$) and 56.7\% ($\lambda$) up to highs of 100\%.
 67\% of the samples pass the formal planarity test but 63\% of the samples are at least somewhat spatially nonplanar (i.e., $\phi < 1$).
@@
 Street networks worldwide are commonly nonplanar because they are embedded in three dimensions, not two.
-But because they are usually \enquote{mostly} planar (average drivable $\phi$ of 88.5\% and average walkable $\phi$ of 92\%), typically with only a few overpasses or underpasses, many could be described as \emph{approximately} planar.
+But because they are usually \enquote{mostly} planar (average drivable $\phi$ of 88\% and average walkable $\phi$ of 89\%), typically with only a few overpasses or underpasses, many could be described as \emph{approximately} planar.
 But formally, a graph is not planar because its edges \emph{usually} intersect only at nodes: by definition it is planar because its edges \emph{exclusively} intersect at nodes.
-However, as \citet{newman_networks:_2010} points out, debating the semantics of formal planarity may be missing the point -- more interesting is the \emph{extent} to which a network is nonplanar.
+However, as \citet{newman_networks:_2010} points out, debating the semantics of formal planarity may be missing the point --- more interesting is the \emph{extent} to which a network is nonplanar.
 How do planar simplifications impact modeling and analysis?
@@
 Do they provide useful models for this type of research?
-Again, only in certain circumstances, such as the urban form of Florence where the planar bias is negligible.
+Again, only in certain circumstances, such as the urban form of Florence where the planar bias is arguably negligible.
 Meanwhile, in our central Los Angeles network, planarity overcounted intersections by 72\% and underestimated edge lengths by 37\%.
@@
 
-We can see the impact of this bias in aggregate across our study sites: the results of common urban form analyses such as intersection counts are overstated by planar models (by 16\% on average in the drivable networks), while average street segment lengths are consequently understated (by 10\% on average in the drivable networks).
+We can see the impact of this bias in aggregate across our study sites: the results of common urban form analyses such as intersection counts are overstated by planar models (by 16\% on average in the drivable networks), while average street segment lengths are consequently understated (by 11\% on average in the drivable networks).
 These impact our understanding of the urban form's density, grain, pattern, connectedness, and permeability.

### 2018-02-19 main.tex
update planar vs non demo figure
revise formal vs spatial planar explanation

@@
 These models are commonly defined as planar, meaning that they can be represented in two dimensions without any underpasses or overpasses.
-However, real-world urban street networks exist in three-dimensional space and frequently feature grade separation such as bridges and tunnels: planar simplifications can be useful but might affect the results of real-world street network analysis.
-Further, a city's extent of nonplanarity could describe its infrastructure and urbanization.
+However, real-world urban street networks exist in three-dimensional space and frequently feature grade separation such as bridges and tunnels: planar simplifications can be useful but they impact the results of real-world street network analysis.
 This study measures the nonplanarity of drivable and walkable street networks in the centers of 50 cities worldwide, then examines the variation of nonplanarity across a single city.
-It develops two new indicators: the Spatial Planarity Ratio and the Edge Length Ratio.
+It develops two new indicators---the Spatial Planarity Ratio and the Edge Length Ratio---to measure planarity and describe infrastructure and urbanization.
 While some street networks are approximately planar, we empirically quantify how planar models can inconsistently but drastically misrepresent intersection density, street lengths, routing, and connectivity.
@@
 
-\keywords{GIS, street networks, transportation, urban form, urban morphology}
+\keywords{GIS, OpenStreetMap, planar graphs, street networks, urban form, urban morphology}
 
@@
 A graph $G$ consists of a set of nodes $N$ connected to one another by a set of edges $E$.
-An edge $uv$ in a directed graph points in one direction from some node $u$ to some node $v$, but an undirected graph's edges all point mutually in both directions.
+An edge $uv$ in a directed graph points in one direction from some node $u$ to some node $v$, while an undirected graph's edges all point mutually in both directions.
 In a street network, the nodes represent intersections and dead-ends, and the (directed) edges represent the street segments that connect them.
-How a graph's nodes and edges connect to one another defines its \emph{topology}.
+The way a graph's nodes and edges connect to one another defines its \emph{topology}.
 For example, a node's \emph{degree} is a topological trait that represents how many edges connect to that node.
@@
 A \emph{planar} graph can be drawn on a two-dimensional plane without any of its edges crossing each other, except where they intersect at nodes.
-If the graph cannot be drawn --- or redrawn --- to meet this criterion, then the graph is nonplanar \citep{trudeau_introduction_1994}.
-Street networks are embedded in space, which provides them with geometry --- such as geographical coordinates, lengths, areas, shapes, and angles --- along with their topology.
+If it cannot be drawn to meet this criterion, it is nonplanar \citep{trudeau_introduction_1994}.
+Street networks are embedded in space, which provides them with geometry---such as geographical coordinates, lengths, areas, shapes, and angles---along with their topology.
 This creates a minor wrinkle when we consider planarity: we must distinguish between a graph's mathematical/topological planarity, which we refer to as \emph{formal planarity}, and the planarity of its real-world spatial embedding, which we refer to as \emph{spatial planarity}.
@@
 	\includegraphics[width=0.49\textwidth]{figures_tables/planar_vs_not.png}
-	\caption{Formally and spatially planar (left) and nonplanar (right) road networks.}
+	\caption{Spatial network examples.
+Gray lines = edges (i.e., street segments), circles = nodes (i.e., intersections and dead ends), red dots = two-dimensional line crossings (i.e., artificial nodes in planar representation).}
 	\label{fig:planar_vs_not}
@@
 
-Consider the examples in Figure \ref{fig:planar_vs_not}.
-The left network comprises a set of paths and streets without any bridges or tunnels.
-In two dimensions, its edges intersect each other only at nodes so it is by definition planar.
-The right network comprises surface streets and a grade-separated freeway interchange with many overpasses and underpasses.
-In two dimensions, its edges sometimes cross each other at non-nodes (i.e., overpasses), so it is spatially nonplanar.
-Furthermore it is impossible to redraw the graph by moving its nodes and edges around in space such that its edges only intersect at nodes.
+Figure \ref{fig:planar_vs_not}'s top row presents trivial examples of this.
+\ref{fig:planar_vs_not}A depicts a spatially nonplanar network (with one non-node line crossing represented by a red dot) that is formally planar since it can be \enquote{redrawn} such that its edges only cross at nodes (\ref{fig:planar_vs_not}B).
+By contrast, \ref{fig:planar_vs_not}C is both formally and spatially nonplanar because it cannot be redrawn to prevent line crossings.
+Figure \ref{fig:planar_vs_not}'s bottom row presents real-world examples.
+\ref{fig:planar_vs_not}D's network contains a single overpass (red dot) and is thus spatially nonplanar.
+However, it is formally planar as we can redraw it by moving its nodes and/or edges around in space (without breaking any connections) such that its edges only intersect at nodes (\ref{fig:planar_vs_not}E).
+The network in Figure \ref{fig:planar_vs_not}F includes a freeway interchange with overpasses and underpasses.
+In two dimensions, its edges frequently cross each other at non-nodes, so it is spatially nonplanar.
+Furthermore, it is mathematically impossible to redraw the graph such that its edges only intersect at nodes.
 Therefore it is also formally nonplanar.
-At a city- or region-wide spatial scale, these nonplanar edge crossings may be relatively uncommon: we might call such a network \emph{approximately} planar.
+
+Imposing a planar model on a street network forces artificial nodes at any line crossings.
+At a citywide or regional spatial scale, these nonplanar edge crossings may be relatively uncommon: we might call such a network \emph{approximately} planar.
 Approximate planarity constrains nonplanar spatial networks such that they do not exhibit certain characteristics found among nonplanar aspatial graphs, such as small-world effects or power-law distributed node degrees \citep{crucitti_centrality_2006,fischer_spatial_2014}.
@@
 Many scholars argue that street networks are planar graphs (e.g., \citealp[p.~18]{batty_network_2005}; \citealp[p.~521]{buhl_topological_2006}; \citealp[p.~1]{hu_topological_2008}; \citealp[p.~259]{masucci_random_2009}; \citealp[p.~114]{porta_networks_2010}; \citealp[p.~3]{strano_elementary_2012}; \citealp[p.~1]{masucci_limited_2013}; \citealp[p.~1074]{strano_urban_2013}; \citealp[p.~168]{law_defining_2017}).
-Others prefer to hedge slightly, arguing that street networks are approximately or essentially planar --- close enough to be well-modeled as such (e.g., \citealp[p.~6]{dill_measuring_2004}; \citealp[p.~3]{cardillo_structural_2006};  \citealp[p.~340]{xie_measuring_2007}; \citealp[p.~1]{barthelemy_modeling_2008}; \citealp[p.~3]{barthelemy_spatial_2011}; \citealp[pp.~563]{chan_urban_2011}; \citealp[p.~1]{gudmundsson_entropy_2013}; \citealp[p.~1]{viana_simplicity_2013}; \citealp[p.~2]{louf_typology_2014}; \citealp[p.~2191]{zhong_detecting_2014}; \citealp[p.~2]{wang_resilience_2015}; \citealp[p.~42]{aldous_routed_2016}; \citealp[p.~257]{barthelemy_paths_2017}).
+Others prefer to hedge slightly, arguing that street networks are approximately or essentially planar---close enough to be well-modeled as such (e.g., \citealp[p.~6]{dill_measuring_2004}; \citealp[p.~3]{cardillo_structural_2006};  \citealp[p.~340]{xie_measuring_2007}; \citealp[p.~1]{barthelemy_modeling_2008}; \citealp[p.~3]{barthelemy_spatial_2011}; \citealp[pp.~563]{chan_urban_2011}; \citealp[p.~1]{gudmundsson_entropy_2013}; \citealp[p.~1]{viana_simplicity_2013}; \citealp[p.~2]{louf_typology_2014}; \citealp[p.~2191]{zhong_detecting_2014}; \citealp[p.~2]{wang_resilience_2015}; \citealp[p.~42]{aldous_routed_2016}; \citealp[p.~257]{barthelemy_paths_2017}).
 
@@
 
-The graph theory literature offers some measures of how \enquote{far off} a nonplanar graph $G$ is from being planar, including its \emph{crossing number} --- the minimum number of edge crossings of any drawing of $G$ --- and its \emph{skewness} --- the minimum number of edges that must be removed from $G$ to produce a planar graph \citep{liebers_planarizing_2001,chimani_non-planar_2009}.
+The graph theory literature offers some measures of how \enquote{far off} a nonplanar graph $G$ is from being planar, including its \emph{crossing number}---the minimum number of edge crossings of any drawing of $G$---and its \emph{skewness}---the minimum number of edges that must be removed from $G$ to produce a planar graph \citep{liebers_planarizing_2001,chimani_non-planar_2009}.
 However, these measures are imperfect and hard to compute \citep{szekely_successful_2004,chimani_vertex_2012}.
@@
 Following \citet{jacobs_great_1995} and \citet{cardillo_structural_2006}, we analyze a consistently sized, square-mile network at the centers of 50 cities worldwide.
-This allows us to consistently examine central urban street networks without being swamped by metropolitan-scale variation or the idiosyncrasies of individual municipalities' spatial extents \citep{gil_street_2017}.
+This allows us to consistently examine central urban street networks without being swamped by metropolitan-scale variation or the idiosyncrasies of individual municipalities' spatial extents.
 The 50 sampled cities span Africa, Asia, Australia, Europe, and North and South America.
 We look separately at each city's drivable and walkable street networks.
-For cities such as Moscow, with a newer commercial central business district (CBD) that lies apart from an \enquote{old town} center, we take the modern CBD as the city center.
+For cities with a newer central business district (CBD) that lies apart from an \enquote{old town} center, we take the modern CBD as the city center.
 
@@
 OpenStreetMap is a collaborative online mapping platform commonly used by researchers because of its high-quality worldwide coverage \citep{haklay_how_2010,jokar_arsanjani_openstreetmap_2015}.
-OSMnx is a Python-based software tool that allows us to automatically download a street network from OpenStreetMap for any study site in the world, and automatically process it into a length-weighted nonplanar directed graph \citep{boeing_osmnx:_2017}.
+OSMnx is a Python-based software tool that allows us to download a street network from OpenStreetMap for any study site in the world, and process it into a length-weighted nonplanar directed graph \citep{boeing_osmnx:_2017}.
 It differentiates between walkable and drivable routes in the circulation network based on individual elements' metadata that describe how the route may be used.
@@
 OpenStreetMap's raw data contain many interstitial nodes in the middle of street segments (forming an expansion graph) to allow streets to curve through space via a series of straight-line approximations.
-OSMnx automatically simplifies each graph's topology to retain nodes only at intersections and dead-ends, while faithfully retaining each edge's true spatial geometry.
+OSMnx automatically simplifies each graph's topology to retain nodes only at intersections and dead-ends, while retaining each edge's true spatial geometry.
 This provides an accurate count of intersections and an accurate measure of edge lengths for comparison between the planar and nonplanar representations of these networks.
@@
 Thus, $\phi$ represents what proportion of the two-dimensional edge crossings in the planar graph are true intersections in the nonplanar graph.
-A spatially planar network with no bridges or tunnels will have a $\phi$ score of 1.0, while lower values indicate the percentage extent to which the network is planar.
+A spatially planar network with no bridges or tunnels will have a $\phi$ score of 1.0, while lower values indicate the extent to which the network is planar.
 From this indicator, we can calculate how planar simplifications overstate intersection counts and connectivity in a street network (i.e., by what percentage the planar model overcounts intersections) as:
@@
 This measures how planar simplifications underestimate edge lengths in each street network by fragmenting street segments at overpasses or underpasses.
-A spatially planar street network will thus have a $\lambda$ score of 1.0, while lower values indicate the percentage extent to which the network is planar.
+A spatially planar street network will thus have a $\lambda$ score of 1.0, while lower values indicate the extent to which the network is planar.
 We can calculate by what percentage the planar graph understates the average edge length as:
@@
 
-Finally, we explore how these three indicators vary throughout a single city.
-To do so we analyze the drivable street network of Oakland, California as a case study.
-Oakland is a reasonably representative midsized American city with a variety of urban form types from gridded street patterns in its flatlands, to winding culs-de-sac in its hills, to freeways and dense blocks around its downtown.
+Finally, street network models (almost) always impose an artificial boundary on the network, and moving the study site even slightly might affect these results \citep{gil_street_2017}.
+We explore how these three indicators vary throughout a single city by analyzing the drivable street network of Oakland, California as a case study.
+Oakland is a midsized American city with a variety of urban form types from gridded street patterns in its flatlands, to winding culs-de-sac in its hills, to freeways and dense blocks around its downtown.
 First we analyze the entire city of Oakland.
@@
 Among the drivable street networks, only 20\% are formally planar.
-On average, they are 88\% spatially planar by the $\phi$ measure and 89\% by the $\lambda$ measure.
+On average, they are 88\% spatially planar by the $\phi$ measure (Spatial Planarity Ratio) and 89\% by the $\lambda$ measure (Edge Length Ratio).
 The individual $\phi$ values indicate that spatial planarity ranges from a high of 100\% in six of these cities to a low of 54\% in Moscow.
@@
 In terms of spatial planarity, the city has a $\phi$ of 91.6\% and a $\lambda$ of 92.9\%.
-This suggests that the planar representation of Oakland's drivable street network overstates the number of intersections --- and thus, the network's connectivity --- by 9.2\% city-wide and understates the average edge length by 7.1\% city-wide.
+This suggests that the planar representation of Oakland's drivable street network overstates the number of intersections---and thus, the network's connectivity---by 9.2\% city-wide and understates the average edge length by 7.1\% city-wide.
 However, these indicators vary across the city: Table \ref{tab:samples_city} presents summary statistics of these planarity indicators across the 100 square-mile samples of its drivable network.
@@
 The type and era of urbanization represent another factor.
-Old European towns or informal settlements in the Global South may contain fewer grade-separated roads --- and thus would be more planar --- than 20th-century American or 21st-century Chinese metropolises, a result of prevailing transportation and engineering technologies when the urban form developed, as well as local terrain, wealth, culture, and politics \citep{southworth_street_1995}.
+Old European towns or informal settlements in the Global South may contain fewer grade-separated roads---and thus would be more planar---than 20th-century American or 21st-century Chinese metropolises, a result of prevailing transportation and engineering technologies when the urban form developed, as well as local terrain, wealth, culture, and politics \citep{southworth_street_1995}.
 
@@
 But formally, a graph is not planar because its edges \emph{usually} intersect only at nodes: by definition it is planar because its edges \emph{exclusively} intersect at nodes.
-However, as \citet{newman_networks:_2010} points out, debating the semantics of formal planarity may be missing the point --- more interesting is the \emph{extent} to which a network is nonplanar.
+However, as \citet{newman_networks:_2010} points out, debating the semantics of formal planarity may be missing the point---more interesting is the \emph{extent} to which a network is nonplanar.
 How do planar simplifications impact modeling and analysis?
@@
 Our results suggest that the answer depends on the study site and on the type of analysis.
-In limited circumstances --- where the circulation network exhibits few (or ideally zero) underpasses, overpasses, or grade-separation --- then certainly yes, especially if the study focuses on polygonal spatial analysis.
+In limited circumstances---where the circulation network exhibits few (or ideally zero) underpasses, overpasses, or grade-separation---then certainly yes, especially if the study focuses on polygonal spatial analysis.
 But universally we cannot answer yes, especially for topological studies: most egregiously, imposing planarity on a nonplanar street network forces false nodes at underpasses and overpasses, breaking routing and network-based accessibility modeling.
@@
 
-Street networks are often defined and modeled as planar graphs in the urban studies literature --- despite typically being nonplanar in reality --- to simplify real-world complexity and make certain analyses tractable.
+Street networks are often defined and modeled as planar graphs in the urban studies literature---despite typically being nonplanar in reality---to simplify real-world complexity and make certain analyses tractable.
 This study measured the formal and spatial planarity of central urban street networks empirically, demonstrating how planar simplifications inconsistently affect urban form analysis results.
-It developed two new measures of the extent to which spatial networks are nonplanar --- the Spatial Planarity Ratio, $\phi$, and the Edge Length Ratio, $\lambda$ ---  to characterize urbanization particularly through the transportation infrastructure's three-dimensionality.
+It developed two new measures of the extent to which spatial networks are nonplanar---the Spatial Planarity Ratio, $\phi$, and the Edge Length Ratio, $\lambda$--- to characterize urbanization particularly through the transportation infrastructure's three-dimensionality.
 Although planar graphs are useful in some circumstances, they behave inconsistently within and across cities by misrepresenting connectivity, accessibility, routing, intersection densities, and street lengths.

### 2018-02-19 main.tex
edits

@@
 \usepackage{moreverb}
-\usepackage[colorlinks,bookmarksopen,bookmarksnumbered,citecolor=red,urlcolor=red]{hyperref}
+\usepackage[hidelinks,bookmarksopen,bookmarksnumbered,citecolor=red,urlcolor=red]{hyperref}
 \newcommand\BibTeX{{\rmfamily B\kern-.05em \textsc{i\kern-.025em b}\kern-.08em T\kern-.1667em\lower.7ex\hbox{E}\kern-.125emX}}
@@
 Models of street networks underlie research in urban travel behavior, accessibility, design patterns, and morphology.
-These models are commonly defined as planar, meaning that they can be represented in two dimensions without any underpasses or overpasses.
-However, real-world urban street networks exist in three-dimensional space and frequently feature grade separation such as bridges and tunnels: planar simplifications can be useful but they impact the results of real-world street network analysis.
+These models are commonly defined as planar, meaning they can be represented in two dimensions without any underpasses or overpasses.
+However, real-world urban street networks exist in three-dimensional space and frequently feature grade separation such as bridges and tunnels: planar simplifications can be useful but they also impact the results of real-world street network analysis.
 This study measures the nonplanarity of drivable and walkable street networks in the centers of 50 cities worldwide, then examines the variation of nonplanarity across a single city.
@@
 
-In urban planning and transportation research, street networks are routinely used to calculate accessibility between origins and destinations or to compute indicators of the urban form, such as block sizes or intersection density and connectivity.
-Mathematical models of street networks, called \emph{graphs}, have grown ubiquitous in the urban studies literature in recent years as they have been used to model household travel patterns, access and equity, pedestrian volume, urban design patterns, spatial morphology, and location centrality and polycentricity \citep{marshall_street_2010,pflieger_switzerland_2010,porta_street_2012,lee_identifying_2014,porta_alterations_2014,marshall_community_2014,hajrasouliha_impact_2015,parthasarathi_street_2015,knight_metrics_2015,xiao_identifying_2016,zhong_revealing_2017}.
+In urban planning and transportation research, street networks are routinely used to calculate accessibility between origins and destinations or to compute indicators of the urban form such as block sizes, intersection density, and connectivity.
+Mathematical models of street networks, called \emph{graphs}, have grown ubiquitous in the urban studies literature in recent years as they have been used to model household travel behavior, access and equity, pedestrian volume, urban design patterns, spatial morphology, and location centrality and polycentricity \citep{marshall_street_2010,pflieger_switzerland_2010,porta_street_2012,lee_identifying_2014,porta_alterations_2014,marshall_community_2014,hajrasouliha_impact_2015,parthasarathi_street_2015,knight_metrics_2015,xiao_identifying_2016,zhong_revealing_2017}.
 
 The urban studies literature often defines street networks as \enquote{planar} or \enquote{approximately planar,} meaning they can be well-represented by a simplified two-dimensional model that inherently precludes overpasses or underpasses.
-However, the analytical impacts of this computationally-useful simplification have not been studied empirically.
-In the real world, urban street networks are embedded in three-dimensional space and often feature grade separation, bridges, and tunnels.
+However, the analytical impacts of this computationally useful simplification have not been studied empirically.
+In the real world, street networks are embedded in three-dimensional space and often feature grade separation, bridges, and tunnels.
 How well do planar graphs model urban street networks?
@@
 This paper first introduces the basics of graph theory relevant to urban studies, focusing on discussions in the research literature about street network planarity.
-Next it discusses the methods used to acquire and analyze the street networks in this study.
+Next it discusses the methods used to acquire and analyze street networks in this study.
 Then it presents the results of this analysis before concluding with a discussion of implications for street network research and urban form studies.
@@
 The way a graph's nodes and edges connect to one another defines its \emph{topology}.
-For example, a node's \emph{degree} is a topological trait that represents how many edges connect to that node.
+For example, a node's \emph{degree} is the topological trait that represents how many edges connect to that node.
 
@@
 For example, a street network might be spatially nonplanar due to its embedding in space (i.e., it contains overpasses or underpasses in the real world), but it could still be formally planar.
-If we \enquote{redraw} the graph by moving its nodes and edges around in space without changing how they are connected to one another (i.e., altering its geometry without altering its topology), there may exist some alternative spatial embedding that prevents edges crossing anywhere but at nodes \citep[for a thorough introduction see][p.~6]{barthelemy_morphogenesis_2017}.
+That is, if we \enquote{redraw} the graph by moving its nodes and edges around in space without changing how they are connected to one another (i.e., altering its geometry without altering its topology), there may exist some alternative spatial embedding that prevents edges crossing anywhere but at nodes \citep[for a thorough introduction see][p.~6]{barthelemy_morphogenesis_2017}.
 In such a case, the street network is formally planar from a topological perspective, but its real-world embedding is spatially nonplanar.
@@
 \begin{figure}[tbp]
-	\center
-	\includegraphics[width=0.49\textwidth]{figures_tables/planar_vs_not.png}
+	\includegraphics[width=0.48\textwidth]{figures_tables/planar_vs_not_labels.png}
 	\caption{Spatial network examples.
@@
 
-Figure \ref{fig:planar_vs_not}'s top row presents trivial examples of this.
-\ref{fig:planar_vs_not}A depicts a spatially nonplanar network (with one non-node line crossing represented by a red dot) that is formally planar since it can be \enquote{redrawn} such that its edges only cross at nodes (\ref{fig:planar_vs_not}B).
+Figure \ref{fig:planar_vs_not}'s top row illustrates trivial examples of this.
+\ref{fig:planar_vs_not}A depicts a spatially nonplanar network (with one non-node line crossing represented by a red dot) that is formally planar since it can be \enquote{redrawn} (\ref{fig:planar_vs_not}B) such that its edges only intersect at nodes.
 By contrast, \ref{fig:planar_vs_not}C is both formally and spatially nonplanar because it cannot be redrawn to prevent line crossings.
-Figure \ref{fig:planar_vs_not}'s bottom row presents real-world examples.
-\ref{fig:planar_vs_not}D's network contains a single overpass (red dot) and is thus spatially nonplanar.
-However, it is formally planar as we can redraw it by moving its nodes and/or edges around in space (without breaking any connections) such that its edges only intersect at nodes (\ref{fig:planar_vs_not}E).
-The network in Figure \ref{fig:planar_vs_not}F includes a freeway interchange with overpasses and underpasses.
+Figure \ref{fig:planar_vs_not}'s bottom row depicts real-world examples.
+\ref{fig:planar_vs_not}D's road network contains a single overpass (red dot) and is thus spatially nonplanar.
+However, it is formally planar as we can redraw it (\ref{fig:planar_vs_not}E), by rearranging an edge in space without breaking any connections, so its edges now intersect only at nodes.
+The network in \ref{fig:planar_vs_not}F includes a freeway interchange with overpasses and underpasses.
 In two dimensions, its edges frequently cross each other at non-nodes, so it is spatially nonplanar.
@@
 If a planar graph models a street network poorly, it could do so in multiple ways.
-Imposing planarity on a nonplanar street network produces artificial nodes in the graph at bridges and tunnels, which breaks routing.
+Imposing planarity on a nonplanar street network produces artificial nodes in the graph at overpasses and underpasses, which breaks routing.
 For example, this misrepresentation would allow a routing engine to recommend a left turn at an \enquote{intersection} that is actually an overpass \citep[p.~6]{kwan_review_1996}.
-Intersection counts and densities would be overestimated in the presence of these false nodes, and edge lengths would be underestimated due to these artificial breakpoints splitting up street segments.
+Intersection counts and densities would be overestimated by the presence of these false nodes, and edge lengths would be underestimated due to these artificial breakpoints splitting up street segments.
 This consequently impacts interpretation of urban form density, grain, pattern, connectedness, and permeability.
@@
 They also fail to adjust for the size, density, or real-world embedding of a spatial network.
-Discussing road networks and approximate spatial planarity, \citet[p.~133]{newman_networks:_2010} argues that due to these drawbacks no widely-accepted measure of the extent of nonplanarity has emerged and calls for the development of better indicators.
+Discussing road networks and approximate spatial planarity, \citet[p.~133]{newman_networks:_2010} argues that due to such drawbacks no widely-accepted measure of the extent of nonplanarity has emerged, and calls for the development of better indicators.
 
 Such measures would be particularly useful for street networks, as the extent to which a network is (or is not) planar can characterize the nature of its circulation infrastructure and urban form.
-For instance, wealthy 20th- and 21st-century cities oriented around automobility and freeways might exhibit more nonplanarity than do older cities or informal settlements oriented around walking.
+For instance, wealthy 20th- and 21st-century cities organized around automobility and freeways might exhibit more nonplanarity than do older cities or informal settlements oriented around walking.
 Beyond the question of graph model goodness-of-fit, such indicators could provide useful information about urban development, civil infrastructure, and transportation system character.
@@
 OSMnx is a Python-based software tool that allows us to download a street network from OpenStreetMap for any study site in the world, and process it into a length-weighted nonplanar directed graph \citep{boeing_osmnx:_2017}.
-It differentiates between walkable and drivable routes in the circulation network based on individual elements' metadata that describe how the route may be used.
-Thus the walkable network may contain surface streets, paths through parks, pedestrian passageways between buildings or under roads, and other walkable paths.
+It differentiates between walkable and drivable routes based on individual elements' metadata that describe how the route may be used.
+Thus the walkable network may contain surface streets, paths through parks, pedestrian flyovers, passageways between buildings, and other walkable paths.
 The drivable network may contain surface streets, grade-separated freeways, and other drivable routes.
@@
 OSMnx automatically simplifies each graph's topology to retain nodes only at intersections and dead-ends, while retaining each edge's true spatial geometry.
-This provides an accurate count of intersections and an accurate measure of edge lengths for comparison between the planar and nonplanar representations of these networks.
+This provides accurate intersection counts and edge length measures for comparison between these networks' planar and nonplanar representations.
 
@@
 
-This measures how planar simplifications underestimate edge lengths in each street network by fragmenting street segments at overpasses or underpasses.
+This measures how planar simplifications fragment street segments at overpasses or underpasses.
 A spatially planar street network will thus have a $\lambda$ score of 1.0, while lower values indicate the extent to which the network is planar.
-We can calculate by what percentage the planar graph understates the average edge length as:
+We can calculate by what percentage the planar graph underestimates the average edge length as:
 
@@
 
-Finally, street network models (almost) always impose an artificial boundary on the network, and moving the study site even slightly might affect these results \citep{gil_street_2017}.
+Thus, equations \ref{eq:spr} and \ref{eq:elr} measure the extent to which a spatial network is planar, while equations \ref{eq:spr_overstates} and \ref{eq:elr_understates} measure the extent to which a planar model misrepresents the network's characteristics.
+
+Finally, street network models almost always impose an artificial boundary on the network, and moving the study site even slightly might affect these results \citep{gil_street_2017}.
 We explore how these three indicators vary throughout a single city by analyzing the drivable street network of Oakland, California as a case study.
@@
 	\centering
-	\includegraphics[width=0.49\textwidth]{figures_tables/regression_lambda_phi_linear.png}
+	\includegraphics[width=0.48\textwidth]{figures_tables/regression_lambda_phi_linear.png}
 	\caption{Plot of $\lambda$ vs $\phi$ by network type with regression lines.}
@@
 The $\lambda$ values indicate that spatial planarity ranges from a high of 100\% in six of these cities to a low of 60\% in Moscow.
-On average across these networks, planar models overcount intersections by 16\% and underestimate street segment lengths by 11\% (Equations \ref{eq:spr_overstates} and \ref{eq:elr_understates}).
+On average across these networks, planar models overcount intersections by 16\% and underestimate street segment lengths by 11\%.
 
@@
 
-Figure \ref{fig:regression_lambda_phi} depicts the relationship between $\phi$ and $\lambda$ across all 50 cities for both network types: it is linear, positive, and very strong (drivable $r^2=0.990$ and walkable $r^2=0.993$).
+Figure \ref{fig:regression_lambda_phi} depicts the relationship between $\phi$ and $\lambda$ across all 50 cities for both network types: it is linear, positive, and strong (drivable $r^2=0.990$ and walkable $r^2=0.993$).
 The coefficients of determination suggest that these two indicators unsurprisingly provide redundant statistical information about the extent to which a network is spatially planar.
@@
 This suggests that the planar representation of Oakland's drivable street network overstates the number of intersections---and thus, the network's connectivity---by 9.2\% city-wide and understates the average edge length by 7.1\% city-wide.
-However, these indicators vary across the city: Table \ref{tab:samples_city} presents summary statistics of these planarity indicators across the 100 square-mile samples of its drivable network.
+However, these indicators vary across the city (Table \ref{tab:samples_city}).
 The samples' mean $\phi$ and $\lambda$ scores are reasonably close to the city-wide values.
 However, the samples range from spatial planarity lows of 56.9\% ($\phi$) and 56.7\% ($\lambda$) up to highs of 100\%.
-67\% of the samples pass the formal planarity test but 63\% of the samples are at least somewhat spatially nonplanar (i.e., $\phi < 1$).
+67\% of the samples pass the formal planarity test, but 63\% of the samples are at least somewhat spatially nonplanar (i.e., $\phi < 1$).
 
@@
 Street networks worldwide are commonly nonplanar because they are embedded in three dimensions, not two.
-But because they are usually \enquote{mostly} planar (average drivable $\phi$ of 88\% and average walkable $\phi$ of 89\%), typically with only a few overpasses or underpasses, many could be described as \emph{approximately} planar.
+But because they are usually \enquote{mostly} planar (average drivable $\phi$ of 88\% and average walkable $\phi$ of 92\%), typically with only a few overpasses or underpasses, many could be described as \emph{approximately} planar.
 But formally, a graph is not planar because its edges \emph{usually} intersect only at nodes: by definition it is planar because its edges \emph{exclusively} intersect at nodes.
@@
 Meanwhile, in our central Los Angeles network, planarity overcounted intersections by 72\% and underestimated edge lengths by 37\%.
-this misrepresentation behaves inconsistently from place to place as the magnitude of bias varies within and across cities and modes of urbanization.
-Misrepresentation is particularly pronounced around the downtowns of North American cities because of the prevalence of freeways, bridges, and underpasses.
+This misrepresentation behaves inconsistently from place to place, as the magnitude of bias varies within and across cities and modes of urbanization.
+Misrepresentation is particularly pronounced around the downtowns of American cities due to the prevalence of freeways, overpasses, and underpasses.
 
@@
 In Florence, the walking network is both spatially and formally nonplanar due to the \textit{sottopassaggio} (pedestrian subway) near Stazione di Santa Maria Novella, its central train station.
-But even networks of non-freeway, non-pedestrian-only surface streets could easily be nonplanar due to tunnels in hilly neighborhoods or bridges over rivers.
+But even networks of non-freeway, non-pedestrian-only surface streets could easily become nonplanar due to tunnels in hilly neighborhoods or bridges over rivers.
 
@@
 This approximate planarity is common across cities because of costs, technology, and politics: it is expensive to engineer an extensively three-dimensional network.
-Furthermore, a city fully consumed by overpasses and tunnels would degrade quality of life, making it politically infeasible.
-
+Moreover, a city fully consumed by overpasses and tunnels would degrade quality of life, making it politically infeasible.
 Future research can further explore how these indicators correlate with other measures of urbanization, development, investment, technology, politics, and era.
-While this study focused on measures of the urban form, particularly intersection counts and street segment lengths, future research can also further explore impacts of planar simplifications on topological analyses and routing.
+While this study focused on measures of the urban form, particularly intersection counts and street segment lengths, future research can further explore impacts of planar simplifications on topological analyses and routing.
 
@@
 This study measured the formal and spatial planarity of central urban street networks empirically, demonstrating how planar simplifications inconsistently affect urban form analysis results.
-It developed two new measures of the extent to which spatial networks are nonplanar---the Spatial Planarity Ratio, $\phi$, and the Edge Length Ratio, $\lambda$--- to characterize urbanization particularly through the transportation infrastructure's three-dimensionality.
+It developed two new measures of the extent to which spatial networks are nonplanar---the Spatial Planarity Ratio, $\phi$, and the Edge Length Ratio, $\lambda$---to characterize urbanization particularly through the transportation infrastructure's three-dimensionality.
 Although planar graphs are useful in some circumstances, they behave inconsistently within and across cities by misrepresenting connectivity, accessibility, routing, intersection densities, and street lengths.

### 2018-02-19 main.tex
update fig, table folders

@@
 \begin{figure}[tbp]
-	\includegraphics[width=0.48\textwidth]{figures_tables/planar_vs_not_labels.png}
+	\includegraphics[width=0.48\textwidth]{figures/planar_vs_not_labels.png}
 	\caption{Spatial network examples.
@@
 	\label{tab:world_cities}
-	\input{figures_tables/table_world_cities.tex}
+	\input{tables/table_world_cities.tex}
 \end{table*}
@@
 	\centering
-	\includegraphics[width=\textwidth]{figures_tables/world_map_bw.png}
+	\includegraphics[width=\textwidth]{figures/world_map_bw.png}
 	\caption{Map of world cities from Table \ref{tab:world_cities} grouped by drivable $\phi$ terciles (lower values mean less planar).}
@@
 	\centering
-	\includegraphics[width=0.48\textwidth]{figures_tables/regression_lambda_phi_linear.png}
+	\includegraphics[width=0.48\textwidth]{figures/regression_lambda_phi_linear.png}
 	\caption{Plot of $\lambda$ vs $\phi$ by network type with regression lines.}
@@
 	\label{tab:samples_city}
-	\input{figures_tables/table_samples_city.tex}
+	\input{tables/table_samples_city.tex}
 \end{table}

### 2018-03-02 main.tex
edits for submission

@@
-\documentclass[Afour,sageh,times]{sagej}
+\documentclass[Afour,doublespace,sageh,times]{sagej}
 \usepackage[T1]{fontenc}
@@
 \usepackage{url}
-\usepackage{moreverb}
-\usepackage[hidelinks,bookmarksopen,bookmarksnumbered,citecolor=red,urlcolor=red]{hyperref}
-\newcommand\BibTeX{{\rmfamily B\kern-.05em \textsc{i\kern-.025em b}\kern-.08em T\kern-.1667em\lower.7ex\hbox{E}\kern-.125emX}}
+\usepackage[colorlinks,bookmarksopen,bookmarksnumbered,citecolor=red,urlcolor=red]{hyperref}
 
@@
 
-\usepackage[final]{draftwatermark}
-\SetWatermarkText{DRAFT}
-\SetWatermarkScale{1}
-\SetWatermarkLightness{0.9}
-
 \begin{document}
 
-\runninghead{Author Redacted}
+\runninghead{}
 \title{Planarity and Street Network Representation in Urban Form Analysis}
-\author{Author Redacted\affilnum{1}}
-\affiliation{\affilnum{1}Affiliation redacted}
-\corrauth{Author Redacted, Address Redacted}
-\email{Email Redacted}
+\author{}
+\affiliation{}
+\corrauth{}
+\email{}
 
@@
 \begin{figure}[tbp]
-	\includegraphics[width=0.48\textwidth]{figures/planar_vs_not_labels.png}
+	\centering
+	\includegraphics[width=0.48\textwidth]{fig_planar_vs_not_labels.png}
 	\caption{Spatial network examples.
@@
 	\label{tab:world_cities}
-	\input{tables/table_world_cities.tex}
+	\input{table_world_cities.tex}
 \end{table*}
@@
 	\centering
-	\includegraphics[width=\textwidth]{figures/world_map_bw.png}
+	\includegraphics[width=\textwidth]{fig_world_map_bw.png}
 	\caption{Map of world cities from Table \ref{tab:world_cities} grouped by drivable $\phi$ terciles (lower values mean less planar).}
@@
 	\centering
-	\includegraphics[width=0.48\textwidth]{figures/regression_lambda_phi_linear.png}
+	\includegraphics[width=0.48\textwidth]{fig_regression_lambda_phi_linear.png}
 	\caption{Plot of $\lambda$ vs $\phi$ by network type with regression lines.}
@@
 	\label{tab:samples_city}
-	\input{tables/table_samples_city.tex}
+	\input{table_samples_city.tex}
 \end{table}
@@
 
-%\clearpage
-\bibliographystyle{apalike}
-\bibliography{references}
+\bibliographystyle{SageH}
+\bibliography{references.bib}
 
